### 2016-10-24 sec-related-work.tex
Begin writing the related work section of the paper.

\section{Related Work}

\citet{Faucon:2016:EDM} propose a semi-Markov model for simulating MOOC
students. They produce behavior profiles that characterize groups of
students in the form of semi-Markov models like our proposed model does,
but they differ from our work in two major ways. First, they assume that a
student can belong to only one behavior profile across the entire course
rather than allowing this profile to change over time.  Because we do not
have this restriction, our model is also able to learn the transition
probabilities between the different behavior profiles we discover.

\citet{Shih:2010:EDM} investigate the use of HMM-based clustering
techniques for automatic discovery of student learning strategies when
solving a particular problem. This is similar to our approach in that the
description of behavior profiles is a Markov-model, but differs in focus:
we investigate macro-level MOOC behavior here as opposed to looking at
individual problem solving behavior. \citet{Ypma:2002:Springer} use
mixtures of HMMs to categorize web pages and cluster users by investigating
web log data, which is quite similar to the clickstream log data we obtain
from MOOCs.

\citet{Kizilcec:2013:LAK} assign students  to states following a rule-based
approach based upon when the student submitted the assignment for a
particular week in the course. They investigated how students transitioned
between these states as the course progressed, and used the sequence of
states a student exhibited as a representation for performing $k$-means
clustering of students into related groups. This differs from our method
substantially: we assign students to states using a probabilistic
framework to account for uncertainty in this state assignment, and jointly
learn representations for these states, which are treated as being
\emph{latent} as opposed to pre-definied using some rule (or set of rules).
Furthermore, our model provides more flexibility in how the time segments
are defined, allowing for both finer (for example, day-by-day) or courser
(for example, month-by-month) granularity.

### 2016-10-24 sec-model.tex
Begin writing section that describes the model formulation.

\section{Model}

Given a MOOC log, we can define a set $\mathbf{A}$ of actions that a
student can take at any given time. For each student $\ell \in \mathbf{L}$,
we then extract a list of action sequences $\mathbf{S}_\ell$ that he or she
produced as observed in the log. Each sequence can be divided flexibly; in
this paper we chose to denote the end of a sequence as occurring when no
further actions occur within a 10 hour window of time (and thus our
sequences roughly correspond to one day's worth of activity). Each action
sequence $s_i \in \mathbf{S}_\ell$ is a list of actions from $\mathbf{A}$.

Each sequence $s_i$ is associated with a latent state $u_i$, and the
actions within the sequence $(a_1, a_2, \ldots, a_{|s_i|})$ are modeled as
a first-order Markov process, where each action is drawn from a
distribution conditioned upon the previous action. (There is also an initial
distribution that governs the probability of starting the sequence with a
particular action $a_1 \in A$.)

For a given list of action sequences $\mathbf{S}_\ell$, we have the
likelihood function
\begin{equation}
  P(\mathbf{S}_\ell \mid \Lambda)
  = \prod_{s_i \in \mathbf{S}_\ell}
  \sum_{U} P(u_i \mid u_{i-1})
  \prod_{a_t \in s_i}
  P(a_t \mid a_{t-1}, u_i)
\end{equation}
where $\Lambda$ is the set of parameters for our model, and the summation
over $U$ refers to summing over all possible latent state sequences of the
same length as $\mathbf{S}_\ell$. (For $i=1$ and for $t=1$, we define the
transition probability as drawing from the initial distribution; this can
be thought of as padding the sequences with a dummy start symbol $u_0$ and
$a_0$, respectively). This can be seen as a modification of the traditional
hidden Markov model~\cite{Rabiner:1990:RSR} where instead of discrete
observations (one for each latent state transition) we have observations
that take the form of entire sequences $(a_1, a_2, \ldots a_T)$ whose
probabilities are computed using another (non-hidden) Markov model
conditioned upon the latent state $u_i$.

\subsection{Parameter Estimation}
To learn the parameters of our model, we may attempt to use maximum
likelihood estimation. Unfortunately, a closed-form solution does not
exist, so we must appeal to the EM algorithm~\cite{Dempster:1977:JRSS}. In
particular, we propose a minor modification of the Baum-Welch
algorithm~\cite{Rabiner:1990:RSR} which is used to learn the parameters for
hidden Markov models. In the following sections, we will provide a brief
description of the original Baum-Welch algorithm, and then describe our
modification.

\subsubsection{Baum-Welch for Traditional HMMs}
In the traditional HMM formulation, we have $\Lambda = (A, B, \pi)$ where
$A$ is a $N\times N$ matrix indicating the latent state transition
probabilities, $B$ is a $N \times V$ matrix indicating the probability of
generating an observation symbol from a given latent state, and $\pi$ is
the initial probability distribution over the latent states.

The Baum-Welch algorithm (also called the forward-backward algorithm)
defines two sets of variables $\alpha_t(i)$ called the forward variables
and $\beta_t(i)$ called the backward variables.

$\alpha_t(i) = P(o_1, \ldots, o_t, q_t = i \mid \Lambda)$ is the probability of
generating the sequence of observations $(o_1, o_2, \ldots, o_t)$ up to time
$t$ and arriving in state $i$ at that time. They are typically defined
using the following recursion:
\begin{itemize}
  \item $\alpha_1(i) = \pi_i b_i(o_1)$, the probability of starting in
    state $i$ ($\pi_i$) times the probability of generating the first
    observation $o_1$ from state $i$.
  \item $\alpha_{t+1}(i) = b_i(o_{t+1})\sum_{j=1}^N \alpha_t(j) a_{ji}$,
    the probability of generating the observation $o_{t+1}$ from state $i$
    times the probability that we arrive in state $i$ from any other
    previous state after generating all of the other observations.
\end{itemize}

Analogously, $\beta_t(i) = P(o_{t+1}, \ldots, o_{T} \mid q_t = i, \Lambda)$
is the probability of generating the \emph{rest of the sequence} given that
we are in state $i$ at time $t$. They are also defined using a recursion:
\begin{itemize}
  \item $\beta_T(i) = 1$
  \item $\beta_t(i) = \sum_{j=1}^N \beta_{t+1}(j) a_{ij} b_j(o_{t+1})$, the
    probability of transitioning to any state $j$ and generating the
    observation $o_{t+1}$ times the probability of generating the rest of
    the sequence given that we transitioned to state $j$.
\end{itemize}

Given the $\alpha$s and the $\beta$s, we can compute the posterior
probability of being in a given state $i$ at time $t$ and the posterior
probability of going through a transition from state $i$ to state $j$ at
time $t$ as
\begin{equation}
  \gamma_t(i) = \frac{\alpha_t(i)\beta_t(i)}{\sum_{j=1}^N
  \alpha_t(j)\beta_t(j)}
\end{equation}
and
\begin{equation}
  \xi_t(i,j) = \frac{\alpha_t(i)a_{ij} b_j(o_{t+1}) \beta_{t+1}(j)}
  {\sum_{j=1}^N \alpha_t(j)\beta_t(j)}
\end{equation}
respectively. These probabilities are enough to update our model parameters
$\Lambda$ as
\begin{align}
  \pi_i &= \gamma_1(i),\\
  a_{ij} &= \frac{\sum_{t=1}^T \xi_t(i,j)}{\sum_{j=1}^N \sum_{t=1}^T
  \xi_t(i,j)}, \text{ and }\\
  b_i(v) &= \frac{\sum_{t=1, o_t = v}^T \gamma_t(i)}
  {\sum_{t=1}^T \gamma_t(i)}.
\end{align}

% TODO: update this to compute \Lambda across multiple observation
% sequences

\subsubsection{Baum-Welch for HMMs with Sequence Observations}
The major deviation of our model from the traditional HMM is that our
observations are themselves sequences. Thus, in all of the equations
presented in the previous section, we treat an observation $o_t$ at time
$t$ as consisting itself of a list of actions $a_1, \ldots, a_M$. Thus,
instead of defining $b_i(v)$ as the probability that an observation
\emph{token} $v$ is generated from state $i$, we have
\begin{equation}
  b_i(a_1, \ldots, a_M) = P(a_1) \prod_{i=2}^M P(a_i \mid a_{i-1})
\end{equation}
where $b_i(\mathbf{o})$ now defines the probability of an action sequence
using a first-order (non-hidden) Markov model. Fortunately, the recursions
for the $\alpha$ and $\beta$ variables remain the same, as do the
definitions of $\gamma_t(i)$ and $\xi_t(i,j)$. The only change, then, is in
the updating equations, where we replace the update for $b_i$ by a pair of
updates

% TODO: add the pair of updates

### 2016-10-24 sec-model.tex
Clean up the model definition section.

@@
 Given a MOOC log, we can define a set $\mathbf{A}$ of actions that a
-student can take at any given time. For each student $\ell \in \mathbf{L}$,
-we then extract a list of action sequences $\mathbf{S}_\ell$ that he or she
-produced as observed in the log. Each sequence can be divided flexibly; in
-this paper we chose to denote the end of a sequence as occurring when no
-further actions occur within a 10 hour window of time (and thus our
-sequences roughly correspond to one day's worth of activity). Each action
-sequence $s_i \in \mathbf{S}_\ell$ is a list of actions from $\mathbf{A}$.
+student can take at any given time. For example, an action $a \in
+\mathbf{A}$ might be ``viewing lecture'', ``taking quiz'', or ``viewing
+forum''. For each student in the course $\ell \in \mathbf{L}$, we then
+extract a list of action sequences $\mathbf{S}_\ell$ that he or she
+produced as observed in the log, where each sequence $\mathbf{s} \in
+\mathbf{S}_\ell$ is itself a list of actions $(a_1, a_2, \ldots, a_M)$ with
+each $a_i \in \mathbf{A}$.  Each sequence can be divided flexibly; in this
+paper we chose to denote the end of a sequence as occurring when no further
+actions occur within a 10 hour window of time (and thus our sequences
+roughly correspond to one day's worth of activity).
 
-Each sequence $s_i$ is associated with a latent state $u_i$, and the
-actions within the sequence $(a_1, a_2, \ldots, a_{|s_i|})$ are modeled as
-a first-order Markov process, where each action is drawn from a
-distribution conditioned upon the previous action. (There is also an initial
-distribution that governs the probability of starting the sequence with a
-particular action $a_1 \in A$.)
+Each sequence $\mathbf{s} \in \mathbf{S}_\ell$ is associated with a latent
+state $u \in \{1,\ldots,K\}$ (where $K$ is a fixed constant picked in
+advance). The actions within the sequence $\mathbf{s} = (a_1, a_2, \ldots,
+a_M)$ are then modeled as a first-order Markov process conditioned upon $u$
+where each action is drawn from a distribution conditioned upon the
+previous action (except for the first which is sampled from an initial
+starting distribution). We can write the parameters for the first-order
+Markov model associated with latent state $u$ as $\lambda^{(u)} =
+(\pi^{(u)}, A^{(u)})$ where $\pi^{(u)}$ indicates the initial probability
+vector of length $|\mathbf{A}|$ and $A^{(u)}$ is an $|\mathbf{A}| \times
+|\mathbf{A}|$ matrix indicating the transition probabilities between each
+pair of actions from $\mathbf{A}$.
 
-For a given list of action sequences $\mathbf{S}_\ell$, we have the
-likelihood function
+Thus, the probability of a sequence $\mathbf{s}$ of length $M$ given its
+latent state $u$ is
+\begin{equation}
+  P(\mathbf{s} \mid \lambda^{(u)}) = P(a_1 \mid \pi^{(u)}) \prod_{i=2}^M P(a_i \mid
+  a_{i-1}, A^{(u)})
+\end{equation}
+where $P(a \mid \pi^{(u)}) = \pi^{(u)}_a$ is the probability of starting
+with action $a$ and $P(a_i \mid a_{i-1}, A^{(u)}) = A^{(u)}_{a_{i-1}, a_i}$
+is the transition probability of moving from action $a_{i-1}$ to $a_{i}$.
+
+We can compute the likelihood of a list of action sequences
+$\mathbf{S}_\ell$ of length $N$ for a student $\ell$ as
 \begin{equation}
   P(\mathbf{S}_\ell \mid \Lambda)
-  = \prod_{s_i \in \mathbf{S}_\ell}
-  \sum_{U} P(u_i \mid u_{i-1})
-  \prod_{a_t \in s_i}
-  P(a_t \mid a_{t-1}, u_i)
+  = \prod_{i=1}^N \sum_U P(U \mid \Lambda) P(\mathbf{s}_i \mid
+  \lambda^{(u_i)})
 \end{equation}
-where $\Lambda$ is the set of parameters for our model, and the summation
-over $U$ refers to summing over all possible latent state sequences of the
-same length as $\mathbf{S}_\ell$. (For $i=1$ and for $t=1$, we define the
-transition probability as drawing from the initial distribution; this can
-be thought of as padding the sequences with a dummy start symbol $u_0$ and
-$a_0$, respectively). This can be seen as a modification of the traditional
-hidden Markov model~\cite{Rabiner:1990:RSR} where instead of discrete
-observations (one for each latent state transition) we have observations
-that take the form of entire sequences $(a_1, a_2, \ldots a_T)$ whose
+where $\Lambda$ is the set of all model parameters, and the summation over
+$U$ indicates a sum over all possible latent state sequences of length
+$|\mathbf{s}_i|$. In our model, we let $\Lambda = (\pi, A, \lambda^{(1)},
+\ldots, \lambda^{(K)})$ where $\pi$ and $A$ are the parameters of a first-order
+Markov model over the latent states and each $\lambda^{(i)}$ consists of
+the parameters for the first-order Markov model over action sequences for
+latent state $i$. Thus $\pi$ (without superscripts) is an initial
+probability vector of length $K$ and $A$ (without superscripts) is a $K
+\times K$ transition probability matrix, analogous to the case with the
+individual first-order Markov model parameters $\lambda^{(i)}$ for each
+latent state.
+
+This can be seen as a modification of the traditional hidden Markov
+model~\cite{Rabiner:1990:RSR} where instead of discrete observations (one
+for each latent state transition) we have observations that take the form
+of entire sequences $\mathbf{s}_i = (a_1, a_2, \ldots a_M)$ whose
 probabilities are computed using another (non-hidden) Markov model
@@
 \subsubsection{Baum-Welch for Traditional HMMs}
-In the traditional HMM formulation, we have $\Lambda = (A, B, \pi)$ where
+In the traditional HMM formulation, we have $\Lambda = (\pi, A, B)$ where
+$\pi$ is the initial probability distribution over the latent states,
 $A$ is a $N\times N$ matrix indicating the latent state transition
-probabilities, $B$ is a $N \times V$ matrix indicating the probability of
-generating an observation symbol from a given latent state, and $\pi$ is
-the initial probability distribution over the latent states.
-
-The Baum-Welch algorithm (also called the forward-backward algorithm)
-defines two sets of variables $\alpha_t(i)$ called the forward variables
-and $\beta_t(i)$ called the backward variables.
+probabilities, and $B$ is a $N \times V$ matrix indicating the probability
+of generating a discrete observation symbol from a given latent state. The
+Baum-Welch algorithm (also called the forward-backward algorithm) defines
+two sets of variables $\alpha_t(i)$ called the forward variables and
+$\beta_t(i)$ called the backward variables.
 
@@
     observation $o_1$ from state $i$.
-  \item $\alpha_{t+1}(i) = b_i(o_{t+1})\sum_{j=1}^N \alpha_t(j) a_{ji}$,
+  \item $\alpha_{t+1}(i) = b_i(o_{t+1})\sum_{j=1}^N \alpha_t(j) A_{ji}$,
     the probability of generating the observation $o_{t+1}$ from state $i$
@@
   \item $\beta_T(i) = 1$
-  \item $\beta_t(i) = \sum_{j=1}^N \beta_{t+1}(j) a_{ij} b_j(o_{t+1})$, the
+  \item $\beta_t(i) = \sum_{j=1}^N \beta_{t+1}(j) A_{ij} b_j(o_{t+1})$, the
     probability of transitioning to any state $j$ and generating the
@@
 
-Given the $\alpha$s and the $\beta$s, we can compute the posterior
-probability of being in a given state $i$ at time $t$ and the posterior
-probability of going through a transition from state $i$ to state $j$ at
-time $t$ as
+Given the $\alpha$s and the $\beta$s, we can compute $\gamma_t(i)$, the posterior
+probability of being in a given state $i$ at time $t$, and $\xi_t(i,j)$,
+the posterior probability of going through a transition from state $i$ to
+state $j$ at time $t$ as
 \begin{equation}
@@
 \begin{equation}
-  \xi_t(i,j) = \frac{\alpha_t(i)a_{ij} b_j(o_{t+1}) \beta_{t+1}(j)}
+  \xi_t(i,j) = \frac{\alpha_t(i) A_{ij} b_j(o_{t+1}) \beta_{t+1}(j)}
   {\sum_{j=1}^N \alpha_t(j)\beta_t(j)}
 \end{equation}
-respectively. These probabilities are enough to update our model parameters
+respectively. This is the E-step of the EM algorithm.
+
+Given $\gamma_t(i)$ and $\xi_t(i,j)$ for each sequence $\mathbf{o}_k$ in
+some training data of length $M$, we can update our model parameters
 $\Lambda$ as
 \begin{align}
-  \pi_i &= \gamma_1(i),\\
-  a_{ij} &= \frac{\sum_{t=1}^T \xi_t(i,j)}{\sum_{j=1}^N \sum_{t=1}^T
-  \xi_t(i,j)}, \text{ and }\\
-  b_i(v) &= \frac{\sum_{t=1, o_t = v}^T \gamma_t(i)}
-  {\sum_{t=1}^T \gamma_t(i)}.
+  \pi_i &= \frac{\sum_{k=1}^M \gamma^{(\mathbf{o}_k)}_1(i)}{M},\\
+  A_{ij} &= \frac{\sum_{k=1}^M \sum_{t=1}^T \xi^{(\mathbf{o}_k)}_t(i,j)}
+  {\sum_{k=1}^M \sum_{j=1}^N \sum_{t=1}^T \xi^{(\mathbf{o}_k)}_t(i,j)},
+  \text{ and }\\
+  b_i(v) &= \frac{\sum_{k=1}^M \sum_{t=1, o_{k,t} = v}^T
+  \gamma^{(\mathbf{o}_k)}_t(i)}
+  {\sum_{k=1}^M \sum_{t=1}^T \gamma^{(\mathbf{o}_{k,t})}_t(i)}.
 \end{align}
-
-% TODO: update this to compute \Lambda across multiple observation
-% sequences
+This is the M-step of the EM algorithm.
 
@@
 observations are themselves sequences. Thus, in all of the equations
-presented in the previous section, we treat an observation $o_t$ at time
-$t$ as consisting itself of a list of actions $a_1, \ldots, a_M$. Thus,
-instead of defining $b_i(v)$ as the probability that an observation
-\emph{token} $v$ is generated from state $i$, we have
+presented in the previous section, we treat an observation $o_t =
+\mathbf{s}_t$ at time $t$ as consisting itself of a list of actions $(a_1,
+\ldots, a_M)$. Thus, instead of defining $b_i(v)$ as the probability that a
+discrete observation \emph{token} $v$ is generated from state $i$, we have
 \begin{equation}
-  b_i(a_1, \ldots, a_M) = P(a_1) \prod_{i=2}^M P(a_i \mid a_{i-1})
+  b_i(\mathbf{s}_t) = P(\mathbf{s}_t \mid \lambda^{(i)})
+  = P(a_1 \mid \pi^{(i)}) \prod_{k=2}^M P(a_k \mid
+  a_{k-1}, A^{(i)})
 \end{equation}
-where $b_i(\mathbf{o})$ now defines the probability of an action sequence
-using a first-order (non-hidden) Markov model. Fortunately, the recursions
-for the $\alpha$ and $\beta$ variables remain the same, as do the
-definitions of $\gamma_t(i)$ and $\xi_t(i,j)$. The only change, then, is in
-the updating equations, where we replace the update for $b_i$ by a pair of
-updates
+where $b_i(\mathbf{s})$ now defines the probability of an action sequence
+using a first-order (non-hidden) Markov model.
 
-% TODO: add the pair of updates
+Fortunately, the recursions for the $\alpha$ and $\beta$ variables remain
+the same, as do the definitions of $\gamma_t(i)$ and $\xi_t(i,j)$ in the
+E-step. The only change is in the updating equations in the M-step,
+where we replace the update for $b_i(v)$ by a pair of updates: one for
+$\pi^{(i)}_a$ one for $A^{(i)}_{ab}$. We denote the $m$-th action in the
+$t$-th sequence for the $k$-th student as $o_{k,t,m}$. The two updates can
+be written as
+\begin{align}
+  \pi^{(i)}_{a}
+  &= \frac{\sum_{k=1}^M \sum_{t=1,a_1 = a}^T \gamma^{(\mathbf{o}_k)}_t(i)}
+  {\sum_{k=1}^M \sum_{t=1}^T \gamma^{(\mathbf{o}_k)}_t(i)}, \text{ and }\\
+  A^{(i)}_{ab}
+  &= \frac{\sum_{k=1}^M \sum_{t=1}^T
+  \sum_{m=2,a_{m-1}=a \land a_m = b}^{|\mathbf{o}_{k,t}|}
+  \gamma^{(\mathbf{o}_k)}_t(i)}
+  {\sum_{k=1}^M \sum_{t=1}^T \sum_{m=2}^{|\mathbf{o}_{k,t}|}
+  \gamma^{(\mathbf{o}_k)}_t(i)}
+\end{align}
+where $a_m = o_{t,k,m}$ (for notational conciseness).

### 2016-10-24 sec-related-work.tex
Add citations for HMMs to related work section.

@@
 \section{Related Work}
+
+Our model is based heavily upon the prior art of Hidden Markov
+Models~\cite{Rabiner:1990:RSR} for sequence labeling tasks. These types of
+models are widely applicable and have been used for tasks such as speech
+recognition~\cite{Huang:1990:HMM} and part-of-speech
+tagging~\cite{Jurafsky:2009:SLP}. We employ maximum likelihood estimation
+for estimating our model parameters using the EM algorithm
+framework~\cite{Dempster:1977:JRSS} by modifying the Baum-Welch algorithm
+for traditional HMMs~\cite{Rabiner:1990:RSR}.
 

### 2016-10-25 sec-intro.tex
Add a first-draft introduction.

\section{Introduction}
The proliferation of massive open online courses (MOOCs) has resulted in a
profound impact on education. As more and more learners turn to MOOCs to
educate themselves on various topics, more and more behavioral data is
being collected as part of the system on which the MOOC is offered. The
data present in these logs has the power to aid us in understanding the
behavior of students who take our MOOCs, which is mostly undetectable for
instructors of these MOOCs today due to its vast scale. As a result, the
rich data available through these MOOC logs is highly underutilized today.

What stands in the way? Clearly, intelligent systems to create concise and
digestible summaries of the massive amount of interaction data collected
are needed in order to empower the instructors of these courses. If we can
understand how users are interacting with our MOOCs, we are much more
likely to be able to make changes to these courses that positively impact
learners. While we can easily observe the changes in behavior of students
in real classrooms, MOOCs present a challenge due to their hands-off
nature. We view this paper as attempting to bridge this gap. Specifically,
in this paper we explore unsupervised learning methods for automatically
detecting student learning behavior patterns or profiles from large
collections of click logs associated with MOOCs.

Our work is motivated by the following beliefs:
\begin{enumerate}
  \item Student behavior is complicated and cannot necessarily be captured
      sufficiently by rule-based methods such as those explored by
      \citet{Kizilcec:2013:LAK}. We instead propose to treat
      student behavior patterns as being generated via a sequence of
      \emph{latent states}. This allows us to automatically capture
      patterns that we might not have been able to articulate clearly a
      priori via a series of rules, and also allows us to model the
      inherent uncertainty in assigning a student's behavior to a
      particular pattern or group.
  \item Student behavior can vary over time. Previous models that treat students
      as exhibiting only one behavioral pattern over
      time~\cite{Faucon:2016:EDM} miss out on the opportunity to understand
      student behavior dynamics in a course.
  \item Analysis of student behavior can and should be performed at varying
      levels of granularity. This requires us to aggregate data over time
      with \emph{different levels of resolution}; existing models tend to come
      with a particular assumption about the resolution of time they
      consider~\cite{Faucon:2016:EDM, Kizilcec:2013:LAK, Shih:2010:EDM}.
\end{enumerate}

Thus, we want a \emph{latent variable approach} to mining student behavior
patterns that is \emph{probabilistic} for inference, and that should be
\emph{flexible to model state changes over different time resolutions}. In
this paper, we propose to discover latent student behavior patterns via
unsupervised learning on large collections of student behavior observation
sequences via a two-layer hidden Markov model.

### 2016-10-25 sec-related-work.tex
Add Bayesian Knowledge Tracing related work; reorganize that section.

@@
 
+\subsection{Hidden Markov Models}
 Our model is based heavily upon the prior art of Hidden Markov
@@
 
-\citet{Faucon:2016:EDM} propose a semi-Markov model for simulating MOOC
-students. They produce behavior profiles that characterize groups of
-students in the form of semi-Markov models like our proposed model does,
-but they differ from our work in two major ways. First, they assume that a
-student can belong to only one behavior profile across the entire course
-rather than allowing this profile to change over time.  Because we do not
-have this restriction, our model is also able to learn the transition
-probabilities between the different behavior profiles we discover.
-
 \citet{Shih:2010:EDM} investigate the use of HMM-based clustering
@@
 we investigate macro-level MOOC behavior here as opposed to looking at
-individual problem solving behavior. \citet{Ypma:2002:Springer} use
-mixtures of HMMs to categorize web pages and cluster users by investigating
-web log data, which is quite similar to the clickstream log data we obtain
-from MOOCs.
+individual problem solving behavior.
 
@@
 (for example, month-by-month) granularity.
+
+\subsection{Bayesian Knowledge Tracing}
+Bayesian Knowledge Tracing~\cite{Corbett:1994:UMUAI} in its basic form uses
+a hidden Markov model to model the probability that a learner knows a
+certain skill at a given time. Modifications to this algorithm include
+contextual estimation of the ``slip'' and ``guess'' probabilities of the
+model~\cite{Baker:2008:ITS} and most recently a re-framing as a neural
+network problem~\cite{Piech:2015:NIPS}.
+
+\subsection{Mixtures of Markov Models}
+\citet{Ypma:2002:Springer} use mixtures of HMMs to categorize web pages and
+cluster users by investigating web log data, which is quite similar to the
+clickstream log data we obtain from MOOCs.
+
+\citet{Faucon:2016:EDM} propose a semi-Markov model for simulating MOOC
+students. They produce behavior profiles that characterize groups of
+students in the form of semi-Markov models like our proposed model does,
+but they differ from our work in two major ways. First, they assume that a
+student can belong to only one behavior profile across the entire course
+rather than allowing this profile to change over time.  Because we do not
+have this restriction, our model is also able to learn the transition
+probabilities between the different behavior profiles we discover.

### 2016-10-25 sec-results.tex
Add results and conclusion sections.

\newcommand{\textretrieval}{\preprintonly{textretrieval-001}%
    \finalonly{textretrieval-001}%
    \submissiononly{moocname1-00X}}

\newcommand{\sustain}{\preprintonly{sustain-001}%
    \finalonly{sustain-001}%
    \submissiononly{moocname2-00X}}

\newcommand{\UIUC}{\preprintonly{UIUC}%
    \finalonly{UIUC}%
    \submissiononly{(redacted University)}}

\section{Results}
To qualitatively evaluate our model, we look at the latent state
representations we learn by fitting the model to the log data associated
with two different Coursera MOOCs: \textretrieval and \sustain.

### 2016-10-25 sec-results.tex
Fix spacing issue with use of macro for MOOC names.

@@
 \section{Results}
-To qualitatively evaluate our model, we look at the latent state
-representations we learn by fitting the model to the log data associated
-with two different Coursera MOOCs: \textretrieval and \sustain.
+To qualitatively evaluate our model, we can look at the latent state
+representations we learn by fitting the model to some MOOC log data. We can
+also inspect the transition matrix between the latent states discovered by
+the model.
+
+Specifically, we look at the MOOC logs associated with two different
+Coursera MOOCs offered by \UIUC{}: \textretrieval{} and \sustain{}.  The
+\textretrieval{} MOOC represents a highly technical computer science course,
+where the \sustain{} MOOC is more representative of a humanities course. We
+picked these two MOOCs because of their vastly different content domains.

### 2016-10-25 sec-conclusion.tex
Add comments to results + conclusion section skeletons.

\section{Conclusion}
% summarize the main contributions of the paper; summarize the experimental
% results; argue again for our novelty; discuss future directions

### 2016-10-25 sec-intro.tex

@@
 in real classrooms, MOOCs present a challenge due to their hands-off
-nature. We view this paper as attempting to bridge this gap. Specifically,
-in this paper we explore unsupervised learning methods for automatically
-detecting student learning behavior patterns or profiles from large
+nature and sometimes irregular schedule due to being a full-time worker.   
+We view this paper as attempting to bridge this gap. Specifically,
+in this paper we propose unsupervised learning methods for automatically
+discovering and characterizing student learning behavior patterns or profiles from large
 collections of click logs associated with MOOCs.
 
-Our work is motivated by the following beliefs:
+Our work is motivated by the following observations:
 \begin{enumerate}
@@
       \citet{Kizilcec:2013:LAK}. We instead propose to treat
-      student behavior patterns as being generated via a sequence of
+      student behavior patterns as being characterized (represented) via a sequence of
       \emph{latent states}. This allows us to automatically capture
@@
       time~\cite{Faucon:2016:EDM} miss out on the opportunity to understand
-      student behavior dynamics in a course.
+      student behavior dynamics in a course. We propose a latent space model to flexibly model the dynamics. 
   \item Analysis of student behavior can and should be performed at varying
@@
       with a particular assumption about the resolution of time they
-      consider~\cite{Faucon:2016:EDM, Kizilcec:2013:LAK, Shih:2010:EDM}.
+      consider~\cite{Faucon:2016:EDM, Kizilcec:2013:LAK, Shih:2010:EDM}. We propose a more flexible model to accommodate different levels of resolution. 
 \end{enumerate}
 
-Thus, we want a \emph{latent variable approach} to mining student behavior
-patterns that is \emph{probabilistic} for inference, and that should be
-\emph{flexible to model state changes over different time resolutions}. In
-this paper, we propose to discover latent student behavior patterns via
+Thus, what we propose is a \emph{latent variable approach} to mining student behavior
+patterns that is \emph{probabilistic} for inference and 
+\emph{flexible to model state changes over different time resolutions}. 
+More specifically, we propose a novel two-layer hidden Markov model (TL-HMM) to
+ discover latent student behavior patterns via
 unsupervised learning on large collections of student behavior observation
-sequences via a two-layer hidden Markov model.
+sequences. Evaluation results on a MOOC data set on Coursera demonstrate that TL-HMM can effectively 
+discover a variety of interesting interpretable student behavior patterns at different levels of resolution, many of which are beyond what existing approaches can discover.
+TL-HMM further enables easy use of the discovered patterns for discriminative analysis such as prediction of learning outcome.
+Since our proposed methods are unsupervised, they can potentially be applied to any MOOC data without requiring any manual work to facilitate understanding of student behaviors and their variations, opening up many possibilities for developing intelligent tutoring systems that can adaptive to student behavior.

### 2016-10-25 sec-results.tex
Add comments to results + conclusion section skeletons.
Start writing results section; minor tweaks elsewhere.

@@
 picked these two MOOCs because of their vastly different content domains.
+Table~\ref{table:datasets} summarizes the two datasets we extracted from
+the MOOCs.
+
+\begin{table}
+  \begin{center}
+    \begin{tabular}{rrrr}
+      \textbf{MOOC} & \textbf{Students} & \textbf{Sequences} & \textbf{Avg.
+      $|\mathbf{s}|$}\\\hline
+      \textretrieval{} & 18,941 & 85,240 & 7.31\\
+      \sustain{} & 85,240 & 231,881 & 15.4
+    \end{tabular}
+    \caption{Statistics about the sequences extracted from the two MOOCs.}
+    \label{table:datasets}
+  \end{center}
+\end{table}
+
+\subsection{Latent State Representations}
+First, we fit a 6-state TL-HMM to the \textretrieval{} sequence dataset and
+show some of the latent state representations we find. We used the
+following ten actions as our action set $\mathbf{A}$:
+\begin{enumerate*}[label=(\arabic*)]
+  \item quiz start,
+  \item quiz submit,
+  \item wiki (course material),
+  \item forum list (view the list of all forums),
+  \item forum thread list (view the list of all threads in a specific
+    forum),
+  \item forum thread view (view the list of posts within a specific
+    thread),
+  \item forum search (a search query issued against the forum),
+  \item forum post thread (a new thread was posted),
+  \item forum post reply (a new post was created within an existing
+    thread), and
+  \item view lecture (defined as either streaming or downloading a lecture
+    video).
+\end{enumerate*}
+
+To visualize these Markov models that represent our latent states, we plot
+them as a directed graph where we set the size of a node to be proportional
+to its personalized Pagerank score~\cite{Page:1999:PageRank, Jeh:2003:WWW}
+where the personalization vector is the initial state distribution for the
+Markov model. We let the thickness of a directed edge $(u, v)$ reflect the
+probability of taking that edge given that a random walk is currently at
+note $u$ (as indicated by the transition matrix). The plots were created
+using python-igraph\footnote{\url{http://igraph.org/python/}}.
+
+\begin{figure*}
+  \centering
+  \begin{subfigure}[t]{0.5\textwidth}
+    \includegraphics[width=\textwidth]{figures/text-6state/state0.png}
+    \caption{An example ``quiz taking'' state.}
+    \label{fig:practice-quiz-state}
+  \end{subfigure}%
+  \begin{subfigure}[t]{0.5\textwidth}
+    \includegraphics[width=\textwidth]{figures/text-6state/state1.png}
+    \caption{An example ``lecture viewing'' state.}
+    \label{fig:lecture-viewing-state}
+  \end{subfigure}
+  \caption{Two example states found by the 6-state TL-HMM.}
+  \label{fig:states}
+\end{figure*}
+
+Figure~\ref{fig:states} includes two such representations we learned. The
+first corresponds to a ``quiz taking'' state whereas the second corresponds
+to a ``lecture viewing'' state. Clearly, our unsupervised method can
+uncover states that do indeed correspond to student behavior modes that we
+would expect to find a priori.
+
+We also argue that it is important that the latent state representation be
+a Markov model rather than just a discrete distribution over actions in
+$\mathbf{A}$ (as would be the case for a traditional single-layer HMM). We
+can observe why if we take a closer look at each of the two latent state
+representations in Figure~\ref{fig:states} and look at their forum
+component (bottom right). We can see that the relative probability of the
+forum activities is roughly the same between these states, but the
+\emph{transitions} are quite different. In
+Figure~\ref{fig:practice-quiz-state} we have a relatively low probability
+of walking from the ``forum thread view'' action back to the ``forum thread
+list'' action, but in Figure~\ref{fig:lecture-viewing-state} we actually
+observe a very strong link in this direction. This difference highlights an
+important distinction between these two latent states: in the first you are
+more likely to visit the forum \emph{looking for a particular post}, where
+in the second you are more likely to visit the forum \emph{to browse
+existing posts}. Thus, capturing the action transition matrix within a
+latent state is important for capturing detailed insights involving bigrams
+of actions.
+
+% highlight interesting latent states discovered; make arguments that
+% motivate why it's useful to have a Markov model as the latent state
+% representation (capturing transitions between actions is useful and
+% more meaningful than just using a discrete distribution as the latent
+% state representation); show how the learned state representations differ
+% between the two MOOCs
+
+\subsection{Varying the Number of Latent States}
+% show what happens as we vary k (particularly interesting to see 2 -> 3
+% hidden state change, as the forum "topic" appears suddenly); argue that
+% this allows the model to capture behavior patterns at different levels of
+% granularity
+
+\subsection{Transitions Between Latent States}
+% using k=6, show that we can uncover interesting patterns in how students
+% transition between the different latent states; argue that this discovery
+% is interesting and unique to our formulation and could not be discovered
+% before

### 2016-10-25 sec-related-work.tex
Revised related work section.

--- 2016-10-24 sec-related-work.tex
@@
 
+%\subsection{Hidden Markov Models for Education Data Analysis}
 Our model is based heavily upon the prior art of Hidden Markov
@@
 recognition~\cite{Huang:1990:HMM} and part-of-speech
-tagging~\cite{Jurafsky:2009:SLP}. We employ maximum likelihood estimation
-for estimating our model parameters using the EM algorithm
-framework~\cite{Dempster:1977:JRSS} by modifying the Baum-Welch algorithm
-for traditional HMMs~\cite{Rabiner:1990:RSR}.
+tagging~\cite{Jurafsky:2009:SLP}.
+A major challenge in applying HMMs successful to solve a problem is to design an appropriate architecture of the model, which always varies according to specific applications.
 
-\citet{Faucon:2016:EDM} propose a semi-Markov model for simulating MOOC
-students. They produce behavior profiles that characterize groups of
-students in the form of semi-Markov models like our proposed model does,
-but they differ from our work in two major ways. First, they assume that a
-student can belong to only one behavior profile across the entire course
-rather than allowing this profile to change over time.  Because we do not
-have this restriction, our model is also able to learn the transition
-probabilities between the different behavior profiles we discover.
+HMMs or similar ideas have been previously 
+applied to model education data (e.g., ~\cite{Shih:2010:EDM,Kizilcec:2013:LAK}), but the previous models are not well tuned toward the student behavior task and thus cannot
+adequately address all the aspects of complexity of student learning behavior. 
+A main technical contribution of this paper is to propose a more general
+HMM that can better adapt to the variations of student behavior via its variable resolution
+and nested HMM structure, and thus enable discovery of more sophisticated behavior patterns and provide more detailed characterization of student behavior than the previous models.
 
-\citet{Shih:2010:EDM} investigate the use of HMM-based clustering
-techniques for automatic discovery of student learning strategies when
-solving a particular problem. This is similar to our approach in that the
-description of behavior profiles is a Markov-model, but differs in focus:
-we investigate macro-level MOOC behavior here as opposed to looking at
-individual problem solving behavior. \citet{Ypma:2002:Springer} use
-mixtures of HMMs to categorize web pages and cluster users by investigating
-web log data, which is quite similar to the clickstream log data we obtain
-from MOOCs.
-
-\citet{Kizilcec:2013:LAK} assign students  to states following a rule-based
+For example, the work \citet{Kizilcec:2013:LAK} assigned students to states following a rule-based
 approach based upon when the student submitted the assignment for a
@@
 are defined, allowing for both finer (for example, day-by-day) or courser
-(for example, month-by-month) granularity.
+(for example, month-by-month) granularity. The work \citet{Shih:2010:EDM} investigated the use of HMM-based clustering
+techniques for automatic discovery of student learning strategies when
+solving a particular problem. This is similar to our approach in that the
+description of behavior profiles is a Markov-model, but cannot further characterize
+each latent state with another informative HMM. Thus their work can be regarded
+to modelng ``micro'' behavior, whereas our model can model both ``micro'' and ``macro'' behavior. 
+%
+ %differs in focus:
+%we investigate macro-level MOOC behavior here as opposed to looking at
+%individual problem solving behavior.
+
+The work \citet{Faucon:2016:EDM} proposed a semi-Markov model for simulating MOOC
+students. They produce behavior profiles that characterize groups of
+students in the form of semi-Markov models like our proposed model does,
+but they differ from our work in two major ways. First, they assume that a
+student can belong to only one behavior profile across the entire course
+rather than allowing this profile to change over time.  Because we do not
+have this restriction, our model is also able to learn the transition
+probabilities between the different behavior profiles we discover.
+
+%We employ maximum likelihood estimation
+%for estimating our model parameters using the EM algorithm
+%framework~\cite{Dempster:1977:JRSS} by modifying the Baum-Welch algorithm
+%for traditional HMMs~\cite{Rabiner:1990:RSR}.
+
+There a few additional related studies worth mentioning. %\subsection{Bayesian Knowledge Tracing}
+Bayesian Knowledge Tracing~\cite{Corbett:1994:UMUAI} in its basic form uses
+a hidden Markov model to model the probability that a learner knows a
+certain skill at a given time. Modifications to this algorithm include
+contextual estimation of the ``slip'' and ``guess'' probabilities of the
+model~\cite{Baker:2008:ITS} and most recently a re-framing as a neural
+network problem~\cite{Piech:2015:NIPS}.
+%\subsection{Mixtures of Markov Models}
+The work \citet{Ypma:2002:Springer} uses mixtures of HMMs to categorize web pages and
+cluster users by investigating web log data, which is quite similar to the
+clickstream log data we obtain from MOOCs.

### 2016-10-25 sec-results.tex
Add cross-course comparison of two states; section on varying K.

@@
 
-% highlight interesting latent states discovered; make arguments that
-% motivate why it's useful to have a Markov model as the latent state
-% representation (capturing transitions between actions is useful and
-% more meaningful than just using a discrete distribution as the latent
-% state representation); show how the learned state representations differ
-% between the two MOOCs
+\begin{figure*}
+  \centering
+  \begin{subfigure}[t]{0.5\textwidth}
+    \includegraphics[width=\textwidth]{figures/text-6state/state4.png}
+    \caption{A state from \textretrieval{}.}
+  \end{subfigure}%
+  \begin{subfigure}[t]{0.5\textwidth}
+    \includegraphics[width=\textwidth]{figures/sustain-6state/state4.png}
+    \caption{A state from \sustain{}.}
+  \end{subfigure}
+  \caption{Two similar example states found by the 6-state TL-HMM.}
+  \label{fig:cross-course}
+\end{figure*}
+
+We can also use our model for cross-course behavior analysis. In
+Figure~\ref{fig:cross-course} we show two latent state representations
+learned by a 6-state TL-HMM, one from \textretrieval{} and one from
+\sustain{}. These two states were chosen as they are the most similar
+across the two courses. However, if we look at the transitions we can see
+some important differences. First, in the state from \textretrieval{}, we
+can see that the probability of returning to the course wiki after viewing
+a lecture is considerably lower than that probability in the state from
+\sustain{}. We also notice that the self-loop for staying in a lecture
+activity in \textretrieval{} is significantly higher probability than it is
+in the state from \sustain{}. This gives us some insight into the lecture
+viewing behavior of students in these two MOOCs which can possibly be a
+reflection of the course's structure. In \textretrieval{}, students are
+likely to view lecture videos in succession directly without first visiting
+the course wiki, whereas in \sustain{} students are much more likely to
+first return the course wiki before watching the next lecture video. This
+observation would be lost if we did not consider the transitions between
+the actions within the latent states.
 
 \subsection{Varying the Number of Latent States}
-% show what happens as we vary k (particularly interesting to see 2 -> 3
-% hidden state change, as the forum "topic" appears suddenly); argue that
-% this allows the model to capture behavior patterns at different levels of
-% granularity
+Our TL-HMM model has a parameter $K$ that sets the number of latent states
+to be learned. We believe that this can allow our model to flexibly
+discover patterns of different granularity, and we can show this by varying
+$K$ for a particular course and observing how the latent state
+representations evolve.
+
+\begin{figure*}
+  \centering
+  \begin{tabular}{cccc}
+    \textbf{State 1} & \textbf{State 2} & \textbf{State 3} & \textbf{State
+    4}\\\hline
+    \includegraphics[width=0.22\textwidth]{figures/text-2state/state0.png}
+    &
+    \includegraphics[width=0.22\textwidth]{figures/text-2state/state1.png}\\
+    \includegraphics[width=0.22\textwidth]{figures/text-3state/state0.png}
+    &
+    \includegraphics[width=0.22\textwidth]{figures/text-3state/state1.png}
+    &
+    \includegraphics[width=0.22\textwidth]{figures/text-3state/state2.png}\\
+    \includegraphics[width=0.22\textwidth]{figures/text-4state/state0.png}
+    &
+    \includegraphics[width=0.22\textwidth]{figures/text-4state/state1.png}
+    &
+    \includegraphics[width=0.22\textwidth]{figures/text-4state/state2.png}
+    &
+    \includegraphics[width=0.22\textwidth]{figures/text-4state/state3.png}
+  \end{tabular}
+  \caption{The evolution of states for increasing $K$ for the
+  \protect\textretrieval{} MOOC.} % I don't know why \protect is needed
+  \label{fig:text-state-evolution}
+\end{figure*}
+
+In Figure~\ref{fig:text-state-evolution} we see the evolution of the latent
+state representations found for the \textretrieval{} MOOC\footnote{Due to
+space constraints, the graphs are quite small, but the states are in the
+same positions as they were in previous figures to aid interpretation.}.
+With $K=2$ we uncover a video watching pattern (state 2) and a course
+material browsing pattern (state 1). However, we can see that when $K=3$ we
+are able to uncover a new pattern involving forum behavior in state 3
+(notice the node sizes on the bottom right). As we increase $K$ to four, we
+can see that state 2 splits into state 2 and state 4. These states appear
+quite similar at a glance, but there are still some key differences. First,
+we can see that state 2 now has a non-negligible forum component, whereas
+state 4 has hardly any weight on the forum.
+
+\begin{figure*}
+  \centering
+  \begin{tabular}{cccc}
+    \textbf{State 1} & \textbf{State 2} & \textbf{State 3} & \textbf{State
+    4}\\\hline
+    \includegraphics[width=0.22\textwidth]{figures/sustain-2state/state0.png}
+    &
+    \includegraphics[width=0.22\textwidth]{figures/sustain-2state/state1.png}\\
+    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state0.png}
+    &
+    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state1.png}
+    &
+    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state2.png}\\
+    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state0.png}
+    &
+    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state1.png}
+    &
+    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state2.png}
+    &
+    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state3.png}
+  \end{tabular}
+  \caption{The evolution of states for increasing $K$ for the
+  \protect\sustain{} MOOC.} % I don't know why \protect is needed
+  \label{fig:sustain-state-evolution}
+\end{figure*}
+
+We observe similar behavior in Figure~\ref{fig:sustain-state-evolution}
+as we increase $K$ when fitting our TL-HMM on the \sustain{} MOOC. Again,
+in the transition between $K=2$ and $K=3$ we discover forum behavior
+patterns as a latent state. However, in the transition between $K=3$ to
+$K=4$ we instead see a refining of that discovered forum state, where
+state 3 captures asymmetric transition probabilities between ``forum thread
+view'' and ``forum thread list''. These states could be seen as redundant,
+in which case a more appropriate setting of $K=3$ may be appropriate for
+the \sustain{} dataset.
 

### 2016-10-26 sec-results.tex
Add missing figures and latent state transition results.

@@
   \begin{tabular}{cccc}
-    \textbf{State 1} & \textbf{State 2} & \textbf{State 3} & \textbf{State
-    4}\\\hline
+    \textbf{State 0} & \textbf{State 1} & \textbf{State 2} & \textbf{State
+    3}\\\hline
     \includegraphics[width=0.22\textwidth]{figures/text-2state/state0.png}
@@
 same positions as they were in previous figures to aid interpretation.}.
-With $K=2$ we uncover a video watching pattern (state 2) and a course
-material browsing pattern (state 1). However, we can see that when $K=3$ we
+With $K=2$ we uncover a video watching pattern (state 1) and a course
+material browsing pattern (state 0). However, we can see that when $K=3$ we
 are able to uncover a new pattern involving forum behavior in state 3
 (notice the node sizes on the bottom right). As we increase $K$ to four, we
-can see that state 2 splits into state 2 and state 4. These states appear
+can see that state 1 splits into state 1 and state 3. These states appear
 quite similar at a glance, but there are still some key differences. First,
-we can see that state 2 now has a non-negligible forum component, whereas
-state 4 has hardly any weight on the forum.
+we can see that state 1 now has a non-negligible forum component, whereas
+state 3 has hardly any weight on the forum.
 
@@
   \begin{tabular}{cccc}
-    \textbf{State 1} & \textbf{State 2} & \textbf{State 3} & \textbf{State
-    4}\\\hline
+    \textbf{State 0} & \textbf{State 1} & \textbf{State 2} & \textbf{State
+    3}\\\hline
     \includegraphics[width=0.22\textwidth]{figures/sustain-2state/state0.png}
@@
 $K=4$ we instead see a refining of that discovered forum state, where
-state 3 captures asymmetric transition probabilities between ``forum thread
+state 2 captures asymmetric transition probabilities between ``forum thread
 view'' and ``forum thread list''. These states could be seen as redundant,
@@
 \subsection{Transitions Between Latent States}
-% using k=6, show that we can uncover interesting patterns in how students
-% transition between the different latent states; argue that this discovery
-% is interesting and unique to our formulation and could not be discovered
-% before
+
+\begin{figure*}
+  \centering
+  \begin{subfigure}[t]{0.5\textwidth}
+    \includegraphics[width=\textwidth,trim={0 3cm 0 2cm}]{figures/trans-comp/trans-avg.png}
+    \caption{\label{fig:trans-avg}}
+  \end{subfigure}%
+  \begin{subfigure}[t]{0.5\textwidth}
+    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{figures/trans-comp/trans-perfect.png}
+    \caption{\label{fig:trans-perfect}}
+  \end{subfigure}
+  \caption{The latent state transition diagrams for a 4-state TL-HMM fit to
+  \protect\textretrieval{} for all students (a) compared to only ``perfect''
+  students (b).}
+  \label{fig:trans-comp}
+\end{figure*}
+
+A unique property of our model is its ability to capture transitions
+between the \emph{behavior patterns themselves} that are captured by the
+latent states. In Figure~\ref{fig:trans-avg} we show the latent state
+transition diagram for a 3-state TL-HMM fit on \textretrieval{}. We can
+immediately observe two things: \begin{enumerate*}[label=(\arabic*)]
+  \item each latent state has a very high ``staying'' probability, and
+  \item the prevalence of each latent state matches our intuition.
+\end{enumerate*}
+In particular, we can see that the forum browsing state (state 2) has
+relatively lower probability than the other states as we might expect. It
+also makes sense that state 0 has rather high probability as this state
+likely captures all sequences where a student logged in to the platform and
+then did nothing else (likely checking for updates). If we look at state 1
+and state 3, their relative probabilities match our intuition as
+well. State 1 seems to capture a more engaged browsing session, where there
+is non-negligible probability associated with different activities such as
+quiz taking and forum browsing and, importantly, these activities have high
+probability symmetric edges (so students are taking quizzes one after the
+other, or viewing forum threads in succession). By contrast, state 3 seems
+to capture a more passive student, with negligible probability mass
+associated with forum activity (with low symmetry in the edges). The link
+between ``quiz submit'' and ``quiz start'' (indicating quiz repetition) is
+also significantly lower than state 1.
+
+Thus, we might expect to see students that perform well in the course
+preferring states 1 and 2 over states 0 and 3. To verify this, we took the
+model we learned on the full training data and retrofit it to training data
+consisting only of sequences produced by students in \textretrieval{} that
+had perfect marks. To prevent the latent state meanings from drifting, we
+forced the model parameters associated with their Markov model
+representations to be fixed, in effect only learning initial and transition
+probabilities for the top layer of our TL-HMM.\ We show the updated latent
+state transition diagram in Figure~\ref{fig:trans-perfect}. While the
+differences are small, we do in fact see what we suspected: states 1 and 2
+have higher probability in the ``perfect'' student group compared to the
+model learned on all students, and states 0 and 3 have lower probability.

### 2016-10-26 sec-results.tex
Fix trans-perfect.png figure and update results.

@@
 probabilities for the top layer of our TL-HMM.\ We show the updated latent
-state transition diagram in Figure~\ref{fig:trans-perfect}. While the
-differences are small, we do in fact see what we suspected: states 1 and 2
-have higher probability in the ``perfect'' student group compared to the
-model learned on all students, and states 0 and 3 have lower probability.
+state transition diagram in Figure~\ref{fig:trans-perfect}. We can clearly
+see that the probability of state 2 has increased dramatically, while the
+probability of states 0 and 3 has decreased. State 1 had its probability
+increase, but only very slightly.

### 2016-10-26 sec-related-work.tex
Minor wording changes in related work section.

--- 2016-10-24 sec-related-work.tex
@@
 recognition~\cite{Huang:1990:HMM} and part-of-speech
-tagging~\cite{Jurafsky:2009:SLP}. We employ maximum likelihood estimation
-for estimating our model parameters using the EM algorithm
-framework~\cite{Dempster:1977:JRSS} by modifying the Baum-Welch algorithm
-for traditional HMMs~\cite{Rabiner:1990:RSR}.
+tagging~\cite{Jurafsky:2009:SLP}.
+A major challenge in applying HMMs successful to solve a problem is to design an appropriate architecture of the model, which always varies according to specific applications.
 
-\citet{Faucon:2016:EDM} propose a semi-Markov model for simulating MOOC
-students. They produce behavior profiles that characterize groups of
-students in the form of semi-Markov models like our proposed model does,
-but they differ from our work in two major ways. First, they assume that a
-student can belong to only one behavior profile across the entire course
-rather than allowing this profile to change over time.  Because we do not
-have this restriction, our model is also able to learn the transition
-probabilities between the different behavior profiles we discover.
+HMMs or similar ideas have been previously 
+applied to model education data~\cite{Shih:2010:EDM,Kizilcec:2013:LAK}, but the previous models are not well tuned toward the student behavior task and thus cannot
+adequately address all the aspects of complexity of student learning behavior. 
+A main technical contribution of this paper is to propose a more general
+HMM that can better adapt to the variations of student behavior via its variable resolution
+and nested HMM structure, and thus enable discovery of more sophisticated behavior patterns and provide more detailed characterization of student behavior than the previous models.
 
-\citet{Shih:2010:EDM} investigate the use of HMM-based clustering
-techniques for automatic discovery of student learning strategies when
-solving a particular problem. This is similar to our approach in that the
-description of behavior profiles is a Markov-model, but differs in focus:
-we investigate macro-level MOOC behavior here as opposed to looking at
-individual problem solving behavior. \citet{Ypma:2002:Springer} use
-mixtures of HMMs to categorize web pages and cluster users by investigating
-web log data, which is quite similar to the clickstream log data we obtain
-from MOOCs.
-
-\citet{Kizilcec:2013:LAK} assign students  to states following a rule-based
+For example, \citet{Kizilcec:2013:LAK} assigned students to states following a rule-based
 approach based upon when the student submitted the assignment for a
@@
 are defined, allowing for both finer (for example, day-by-day) or courser
-(for example, month-by-month) granularity.
+(for example, month-by-month) granularity. \citet{Shih:2010:EDM} investigated the use of HMM-based clustering
+techniques for automatic discovery of student learning strategies when
+solving a particular problem. This is similar to our approach in that the
+description of behavior profiles is a Markov model, but cannot further characterize
+each latent state with another informative HMM. Thus their work can be regarded
+to modeling ``micro'' behavior, whereas our model can model both ``micro'' and ``macro'' behavior.
+
+\citet{Faucon:2016:EDM} proposed a semi-Markov model for simulating MOOC
+students. They produce behavior profiles that characterize groups of
+students in the form of semi-Markov models like our proposed model does,
+but they differ from our work in two major ways. First, they assume that a
+student can belong to only one behavior profile across the entire course
+rather than allowing this profile to change over time.  Because we do not
+have this restriction, our model is also able to learn the transition
+probabilities between the different behavior profiles we discover.
+
+There a few additional related studies worth mentioning.
+Bayesian Knowledge Tracing~\cite{Corbett:1994:UMUAI} in its basic form uses
+a hidden Markov model to model the probability that a learner knows a
+certain skill at a given time. Modifications to this algorithm include
+contextual estimation of the ``slip'' and ``guess'' probabilities of the
+model~\cite{Baker:2008:ITS} and most recently a re-framing as a neural
+network problem~\cite{Piech:2015:NIPS}.
+\citet{Ypma:2002:Springer} use mixtures of HMMs to categorize web pages and
+cluster users by investigating web log data, which is quite similar to the
+clickstream log data we obtain from MOOCs.

### 2016-10-26 sec-conclusion.tex
Write a first-draft conclusion.

@@
 % results; argue again for our novelty; discuss future directions
+We proposed a two-layer hidden Markov model for MOOC student behavior
+modeling in an unsupervised learning framework on large collections of
+actions extracted from MOOC log data. This model is different from existing
+methods in that it treats behavior patterns as a sequence of \emph{latent
+states}, rather than assigning these states in a rule-based manner. It
+allows for students to exhibit different behaviors over time, and allows
+analysis at different levels of granularity.
+
+We showed that such a model does in fact capture meaningful behavior
+patterns and produces descriptions of these behavior patterns that are easy
+to interpret. We argued that it is important to capture student behavior
+patterns with more sophisticated models than simple discrete distributions
+over actions in order to capture information present in bigrams of actions
+(or above). By varying the parameter $K$ controlling the number of latent
+states inferred, we showed that the model is flexible and can capture
+patterns of differing levels of specificity in this way.  Finally, we
+investigated whether we can detect differences in student behavior patterns
+as they correlate with course performance. Specifically, we demonstrated
+that high-performing students produce substantially different HMM
+transition diagrams.
+
+\subsection{Future Work}
+
+Our model can be used to produce an ordered list of latent states for each
+student over time. Given these labeled sequences, we could correlate
+course-wide drifts in these latent states with events in a course. For
+example, we might be able to automatically discover difficult or confusing
+parts of a course by noticing spikes in the distribution of students over
+latent states over time.
+
+Our model does not explicitly model drop-out like
+\citet{Kizilcec:2013:LAK}, but doing so is an obvious extension. Our model
+should be able to provide predictions of when a student is ``at risk'' for
+dropping out under such an extension.
+
+Currently, the model learns a transition matrix over the latent states that
+is \emph{shared} across all students. It would be interesting to instead
+learn a different latent state transition matrix for each individual
+student, but keep the second-level Markov models shared. This would provide
+the model with more flexibility which may be desirable itself, but would
+also naturally result in a description of a student (via his or her HMM
+transitions) that could be incorporated into existing supervised learning
+techniques that try to predict student outcomes.

### 2016-10-26 sec-model.tex
Add citation for MeTA.
Redact MeTA's name for blind reviewing in submission mode.
Finished a preliminary pass on the model section (mainly to add a section to give an intuiive introduction to the model and explain the motivation for designing the model in this way)

@@
-\section{Model}
+\section{A Two-Layer HMM for MOOC Log Analysis}
+
+\subsection{Basic idea and rationale}
+
+Our general idea is to use a probabilistic generative model to model 
+the student activities as recorded in a MOOC log, which means we will 
+assume that all the observed student activities are samples drawn (i.e., ``generated'') from 
+a parameterized probabilistic model. We can then estimate the 
+parameter values of the probabilistic model by fitting the model to a specific
+MOOC log data set. The estimated parameter values could then be treated as 
+the latent ``knowledge'' discovered from the data. Because such a generative model 
+attempts to fit {\em all} the data, it enables us to discover interesting patterns
+that can explain the {\em overall} behavior of a student or the {\em common} behavior patterns shared by many students.
+
+HMM is a specific probabilistic generative model with a ``built-in'' state transition system
+that would control the data to be generated by the model, thus it is especially 
+suitable for modeling sequence data. At any moment, the HMM would be in 
+one of $k$ states $U=\{u_1, ...,u_k\}$, and at the next moment, the HMM would move to 
+another state stochastically according to a transition matrix that specifies the probability of
+going to state $u_i$ when the HMM is currently in state $u_j$, i.e., $p(u_i|u_j)$. 
+When the HMM is in state $u$, the HMM can generate an observable data point $x$ 
+according to an output probabilistic model $p(x|u)$. Thus if we ``run'' an HMM for 
+$N$ time points denoted by $t=1, ..., N$, the HMM could ``generate'' a sequence of 
+observations $x_1 ... x_N$, where each $x_i$ is an output symbol by 
+going through a sequence of {\em hidden states} $w_1 ... w_N$ where $w_i \in U$ is a state. 
+The association of such a latent sequence of state transitions with the observed symbols makes
+it possible to use HMM to ``decode'' the latent behavior of students behind the surface behavior we directly observe in the log data, allowing for understanding student behavior more deeply than a model with no latent state variables.
+
+In many ways, the generation process behind an HMM is meant to simulate the actual behavior of a student in the following sense: First, students may be in different ``task state'' (or ``behavior state'')  in the process of study.
+For example, one task state may be to learn about a topic by mostly watching lecture videos, another task state may be to work on quizzes, and yet another may be to participate in forum discussions.
+While in each of these different states, the student would tend to exhibit different surface ``micro'' behaviors.
+For example, in the lecture study state, the student would tend to have many video-watching related behaviors and occasionally forum activities, while in the quiz-taking state (in order to pass each module), the student would tend to show many quiz-related ``micro'' activities as well as asking questions or checking disussions on the forum.
+Note that due to the complexity of the student behavior, it is very difficult to accurately {\em prescribe} the specific surface ``micro'' behavior patterns for each state, especially without  prior knowledge about the students.
+For example, forum activities are likely interleaved with other activities   in every task state and the interleaving pattern can be somewhat irregular with potentially many variations.
+A major motivation for using an HMM is because 1) it uses a probabilistic model (i.e., the output probability distribution $p(x|u)$ conditioned on each state) to directly capture the inevitable uncertainty in the association of surface ``micro'' activities with their corresponding latent task/behavior state, which is often our main taraget to discover and characterize, and 2) it does not make any assumption about which latent task/behavior state must be associated with which observed activities or how a student would move from one state to another, but instead let our data ``tell'' us what kind of associations are most likely and what kind of transitions are most often, and which states tend to be more long-lasting for any particular set of students.
+
+However, if we are to use an ordinary HMM to analyze our data, we would treat each observed ``micro'' activity (e.g., video watching, or forum post reading) as an output symbol, and thus the output distribution $p(x|u)$ for each discovered latent state would be a simple distribution over all kinds of observable micro activities recorded in our log data (e.g., 50\% lecture watching, 8\% quiz taking, 7\% quiz submission, 2\% course wiki reading, ...).
+While such a distribution is meaningful and can already help us intepret the corresponding latent state, it only gives us a rather superficial characterization of student behavior.
+Ideally, we want $p(x|u)$ to characterize the directly observable behavior in more detail to further capture the relations and dependencies of these micro activities.
+To this end, we would treat an {\em entire sequence} of micro activities (e.g., one session of activities) as an observed ``symbol'' from a latent state, and further model the generation of such a sequence with another Markov model where we treat each micro activity as an {\em observable} state, and model the transitions between these activity states in very much the same way as the state transitions in HMM.
+Combining this ``surface'' Markov model with the ``deep'' hidden Markov model gives us a fairly general and powerful two-layer HMM (TL-HMM) that can simultaneously learn ``deeply'' the latent task/behavior states and their transitions as well as the corresponding ``surface'' activity transition patterns associated with each latent state to facilitate interpretation and analysis of the discoverd latent state patterns.
+Next, we present this model more formally and discuss how to estimate its parameters.
+
+\subsection{Formal definition of TL-HMM}
 
@@
 \end{align}
-where $a_m = o_{t,k,m}$ (for notational conciseness).
+where $a_m = o_{t,k,m}$ (for notational conciseness). Our modified EM
+algorithm for HMMs with sequence observations is provided as part of the
+\meta/ toolkit~\citemeta{}.

### 2016-10-26 sec-model.tex
Minor updates to model intuition section.

--- 2016-10-24 sec-model.tex
@@
-\section{Model}
+\section{A Two-Layer HMM for MOOC Log Analysis}
+
+\subsection{Basic idea and rationale}
+
+Our general idea is to use a probabilistic generative model to model 
+the student activities as recorded in a MOOC log, which means we will 
+assume that all the observed student activities are samples drawn (i.e., ``generated'') from 
+a parameterized probabilistic model. We can then estimate the 
+parameter values of the probabilistic model by fitting the model to a specific
+MOOC log data set. The estimated parameter values could then be treated as 
+the latent ``knowledge'' discovered from the data. Because such a generative model 
+attempts to fit {\em all} the data, it enables us to discover interesting patterns
+that can explain the {\em overall} behavior of a student or the {\em common} behavior patterns shared by many students.
+
+An HMM  is a specific probabilistic generative model with a ``built-in'' state transition system
+that would control the data to be generated by the model, thus it is especially 
+suitable for modeling sequence data~\cite{Rabiner:1990:RSR, Huang:1990:HMM}. At any moment, the HMM would be in 
+one of $k$ states $U=\{u_1, ...,u_k\}$, and at the next moment, the HMM would move to 
+another state stochastically according to a transition matrix that specifies the probability of
+going to state $u_i$ when the HMM is currently in state $u_j$, i.e., $p(u_i|u_j)$. 
+When the HMM is in state $u$, the HMM can generate an observable data point $x$ 
+according to an output probabilistic model $p(x|u)$. Thus if we ``run'' an HMM for 
+$N$ time points denoted by $t=1, ..., N$, the HMM could ``generate'' a sequence of 
+observations $x_1 ... x_N$, where each $x_i$ is an output symbol by 
+going through a sequence of {\em hidden states} $w_1 ... w_N$ where $w_i \in U$ is a state. 
+The association of such a latent sequence of state transitions with the observed symbols makes
+it possible to use HMM to ``decode'' the latent behavior of students behind the surface behavior we directly observe in the log data, allowing for understanding student behavior more deeply than a model with no latent state variables.
+
+In many ways, the generation process behind an HMM is meant to simulate the
+actual behavior of a student. We may say that students transition through different ``task states'' (or ``behavior states'')  in the process of study. 
+One such task state may be to learn about a topic by mostly watching
+lecture videos, another task state may be to work on quizzes, and yet
+another may be to participate in forum discussions. While in each of these
+different states, the student would tend to exhibit different surface
+``micro'' behaviors. For example, in the lecture study state, the student
+would tend to have many video-watching related behaviors and occasionally
+forum activities, while in the quiz-taking state (in order to pass each
+module), the student would tend to show many quiz-related ``micro''
+activities as well as asking questions or checking discussions on the
+forum. Note that due to the complexity of the student behavior, it is very
+difficult to accurately {\em prescribe} the specific surface ``micro''
+behavior patterns for each state in advance, especially without  prior knowledge about the students.
+For example, forum activities are likely interleaved with other activities   in every task state and the interleaving pattern can be somewhat irregular with potentially many variations.
+The major motivations for using an HMM are that 
+\begin{enumerate*}[label=(\arabic*)]
+    \item it uses a probabilistic model (i.e., the output probability distribution $p(x|u)$ conditioned on each state) to directly capture the inevitable uncertainty in the association of surface ``micro'' activities with their corresponding latent task/behavior state, which is often our main target to discover and characterize, and
+    \item it does not make any assumption about which latent task/behavior
+      state must be associated with which observed activities or how a
+      student would move from one state to another, but instead allows our
+      data to ``tell'' us what kind of associations are most likely,
+      what kind of transitions are most probable, and which states tend to be more long-lasting for any particular set of students. 
+\end{enumerate*}
+
+However, if we use an ordinary HMM to analyze our data, we would treat each
+observed ``micro'' activity (e.g., video watching, or forum post reading)
+as an output symbol, and thus the output distribution $p(x|u)$ for each
+discovered latent state would be a simple distribution over all kinds of
+observable micro activities recorded in our log data (e.g., 50\% lecture
+watching, 8\% quiz taking, 7\% quiz submission, 2\% course wiki reading,
+...). While such a distribution is meaningful and can already help us
+interpret the corresponding latent state, it only gives us a rather
+superficial characterization of student behavior.
+
+Ideally, we want $p(x|u)$ to characterize the directly observable ``micro''
+behavior in more detail to further capture the relations and dependencies
+of these micro activities. To this end, we would treat an {\em entire
+sequence} of micro activities (e.g., one session of activities) as an
+observed ``symbol'' from a latent state, and further model the generation
+of such a sequence with another Markov model where we treat each micro
+activity as an {\em observable} state, and model the transitions between
+these activity states in very much the same way as the state transitions in
+HMM.
+
+Adding this second layer would allow us to characterize a latent task
+state in much more detail, as it would reveal not only what activities are
+most common to a task state, but also the transition patterns between these
+``micro'' activities (e.g., it can reveal frequent back-and-forth
+transitions between quiz-taking and quiz-submission, which would suggest a
+concentrated period of taking quizzes). Combining this ``surface'' Markov
+model over the ``micro'' actions with the ``deep'' hidden Markov model over
+the latent task states gives us a fairly general and powerful two-layer HMM
+(TL-HMM) that can simultaneously learn ``deeply'' the latent task/behavior
+states and their transitions as well as the corresponding ``micro''
+activity transition patterns associated with each latent state to
+facilitate interpretation and analysis of the discovered latent state
+patterns. Next, we present this model more formally and discuss how to
+estimate its parameters to uncover these latent patterns in an unsupervised
+manner.
+
+\subsection{Formal definition of TL-HMM}
 

### 2016-10-26 sec-conclusion.tex
Revised conclusions

@@
 states}, rather than assigning these states in a rule-based manner. It
-allows for students to exhibit different behaviors over time, and allows
+captures the variable behaviors of students over time, and allows
 analysis at different levels of granularity.
@@
 over actions in order to capture information present in bigrams of actions
-(or above). By varying the parameter $K$ controlling the number of latent
+(or above). By varying the number of latent
 states inferred, we showed that the model is flexible and can capture
@@
 that high-performing students produce substantially different HMM
-transition diagrams.
+transition diagrams that tend to show longer concentration span in quiz-taking
+and more active  forum participation as compared with the average students.
 
@@
 
-Our model can be used to produce an ordered list of latent states for each
+Although we only experimented with our model on two MOOCs, the model
+is completely general and can be easily applied to analyze the log of any other
+course to enable deep understanding of student behaviors as well as the coorrelations
+of such behaviors as other variables such as grades. We plan to develop a MOOC log analysis
+system based on the proposed model to both facilitate education research and 
+help instructors improve course design.
+
+Specifically, our model can be used to produce an ordered list of latent states for each
 student over time. Given these labeled sequences, we could correlate
@@
 transitions) that could be incorporated into existing supervised learning
-techniques that try to predict student outcomes.
+techniques that try to predict student outcomes for understanding
+which of the latent behavior patterns discovered by TL-HMM are most predictive
+of the performance of student learning.

### 2016-10-26 sec-model.tex
Rewrite equation 2 more clearly (and correctly).

@@
 We can compute the likelihood of a list of action sequences
-$\mathbf{S}_\ell$ of length $N$ for a student $\ell$ as
+$\mathbf{S}_\ell$ of length $N$ for a student $\ell$ by marginalizing over
+all possible latent state sequences $U=(v_1, \ldots, v_N)$ as
 \begin{equation}
+  \begin{split}
   P(\mathbf{S}_\ell \mid \Lambda)
-  = \prod_{i=1}^N \sum_U P(U \mid \Lambda) P(\mathbf{s}_i \mid
-  \lambda^{(u_i)})
+  &= \sum_U \left(P(v_1 \mid \Lambda) P(\mathbf{s}_1 \mid \lambda^{(v_1)})
+  \phantom{\prod_{i=2}^N}\right.\\
+  &\quad \left.\times \prod_{i=2}^N P(v_i \mid v_{i-1}, \Lambda)
+  P(\mathbf{s}_i \mid \lambda^{(v_i)})\right)
+  \end{split}
 \end{equation}
-where $\Lambda$ is the set of all model parameters, and the summation over
-$U$ indicates a sum over all possible latent state sequences of length
-$|\mathbf{s}_i|$. In our model, we let $\Lambda = (\pi, A, \lambda^{(1)},
-\ldots, \lambda^{(K)})$ where $\pi$ and $A$ are the parameters of a first-order
-Markov model over the latent states and each $\lambda^{(i)}$ consists of
-the parameters for the first-order Markov model over action sequences for
-latent state $i$. Thus $\pi$ (without superscripts) is an initial
-probability vector of length $K$ and $A$ (without superscripts) is a $K
-\times K$ transition probability matrix, analogous to the case with the
-individual first-order Markov model parameters $\lambda^{(i)}$ for each
-latent state.
+where $\Lambda$ is the set of all model parameters.  In our model, we let
+$\Lambda = (\pi, A, \lambda^{(1)}, \ldots, \lambda^{(K)})$ where $\pi$ and
+$A$ are the parameters of a first-order Markov model over the latent states
+and each $\lambda^{(i)}$ consists of the parameters for the first-order
+Markov model over action sequences for latent state $i$. Thus $\pi$
+(without superscripts) is an initial probability vector of length $K$ and
+$A$ (without superscripts) is a $K \times K$ transition probability matrix,
+analogous to the case with the individual first-order Markov model
+parameters $\lambda^{(i)}$ for each latent state.
 

### 2016-10-26 sec-results.tex
Add results for ``low'' student group; incorporate some of Cheng's comments.

@@
 
-\section{Results}
-To qualitatively evaluate our model, we can look at the latent state
-representations we learn by fitting the model to some MOOC log data. We can
-also inspect the transition matrix between the latent states discovered by
-the model.
+\section{Experiment Results}
+As an analysis tool, the TL-HMM model provides with us the following two
+patterns to characterize student behavior:
+\begin{enumerate*}[label=(\arabic*)]
+  \item latent state representations, and
+  \item latent state transitions.
+\end{enumerate*}
+Thus to evaluate the proposed model, we conduct experiments to
+qualitatively analyze both types of patterns discovered from empirical MOOC
+log data.
 
@@
   \centering
-  \begin{subfigure}[t]{0.5\textwidth}
+  \begin{subfigure}[t]{0.33\textwidth}
     \includegraphics[width=\textwidth,trim={0 3cm 0 2cm}]{figures/trans-comp/trans-avg.png}
@@
   \end{subfigure}%
-  \begin{subfigure}[t]{0.5\textwidth}
+  \begin{subfigure}[t]{0.33\textwidth}
     \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{figures/trans-comp/trans-perfect.png}
     \caption{\label{fig:trans-perfect}}
+  \end{subfigure}%
+  \begin{subfigure}[t]{0.33\textwidth}
+    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{figures/trans-comp/trans-low.png}
+    \caption{\label{fig:trans-low}}
   \end{subfigure}
@@
   \protect\textretrieval{} for all students (a) compared to only ``perfect''
-  students (b).}
+  students (b) and only ``low'' students (c).}
   \label{fig:trans-comp}
@@
 increase, but only very slightly.
+
+In Figure~\ref{fig:trans-low} we plot the latent state transition diagram
+for a second group of ``low'' students. These students were selected so
+that they attempted all required quizzes in the course, but such that their
+average quiz score was $\leq 70\%$. Here, we see that \emph{state 1} has a
+large increase in size, where we might have expected state 3 to grow
+instead.  However, there is an alternative explanation for this phenomenon:
+Since state 1 seems to indicate a highly engaged student, it is a perfectly
+reasonable explanation for the ``low'' student group as they are going to
+be working hard to try to fill in the gaps in their knowledge. By contrast,
+the ``perfect'' student group likely has many members who can take the quiz
+more passively and get perfect marks, perhaps because they already know
+much of the material being presented, or are just naturally strong and do
+not require much background review to perform well. This also explains why
+state 1 did not increase in size for the ``perfect'' group like we were
+anticipating. Our model lends itself well to discovering this potentially
+counter-intuitive insight directly from data.

### 2016-10-26 sec-model.tex
Add a few more citations, minor cleanups.

@@
 
-\subsection{Formal definition of TL-HMM}
+\subsection{Formal Definition of TL-HMM}
 
@@
 algorithm~\cite{Rabiner:1990:RSR} which is used to learn the parameters for
-hidden Markov models. In the following sections, we will provide a brief
-description of the original Baum-Welch algorithm, and then describe our
-modification.
+hidden Markov models in an unsupervised setting. In the following sections,
+we will provide a brief description of the original Baum-Welch algorithm,
+and then describe our modification.
 
@@
 where $a_m = o_{t,k,m}$ (for notational conciseness). Our modified EM
-algorithm for HMMs with sequence observations is provided as part of the
-\meta/ toolkit~\citemeta{}.
+algorithm for TL-HMMs is provided as part of the \meta/
+toolkit~\citemeta{}.

### 2016-10-26 sec-related-work.tex
Add a few more citations, minor cleanups.

@@
 recognition~\cite{Huang:1990:HMM} and part-of-speech
-tagging~\cite{Jurafsky:2009:SLP}.
-A major challenge in applying HMMs successful to solve a problem is to design an appropriate architecture of the model, which always varies according to specific applications.
+tagging~\cite{Jurafsky:2009:SLP}, and are a member of the more general
+family of probabilistic graphical models~\cite{Koller:2009:PGM}.
+A major challenge in applying HMMs and other graphical models successfully to
+solve a problem is to design an appropriate architecture of the model, which
+always varies according to specific applications.
 

### 2016-10-26 sec-results.tex
Add a few more citations, minor cleanups.

@@
 \begin{enumerate*}[label=(\arabic*)]
-  \item latent state representations, and
-  \item latent state transitions.
+  \item the latent state representations, and
+  \item the latent state transitions.
 \end{enumerate*}
@@
   \begin{tabular}{cccc}
-    \textbf{State 0} & \textbf{State 1} & \textbf{State 2} & \textbf{State
-    3}\\\hline
+    \textbf{State 0 (Wiki)} & \textbf{State 1 (Lecture)}
+    & \textbf{State 2 (Thread View)} & \textbf{State 3 (Lecture/Wiki)}\\\hline
     \includegraphics[width=0.22\textwidth]{figures/text-2state/state0.png}
@@
   \caption{The evolution of states for increasing $K$ for the
-  \protect\textretrieval{} MOOC.} % I don't know why \protect is needed
+  \protect\textretrieval{} MOOC.\ State ``names'' indicate the highest
+  probability action(s) within the state.} % I don't know why \protect is needed
   \label{fig:text-state-evolution}
@@
   \begin{tabular}{cccc}
-    \textbf{State 0} & \textbf{State 1} & \textbf{State 2} & \textbf{State
-    3}\\\hline
+    \textbf{State 0 (Wiki)}
+    & \textbf{State 1 (Lecture/Wiki)}
+    & \textbf{State 2 (Thread List)}
+    & \textbf{State 3 (Thread View)}\\\hline
     \includegraphics[width=0.22\textwidth]{figures/sustain-2state/state0.png}
@@
     &
-    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state1.png}
+    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state2.png}
     &
-    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state2.png}\\
+    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state1.png}\\
     \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state0.png}
+    &
+    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state3.png}
     &
@@
     \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state2.png}
-    &
-    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state3.png}
   \end{tabular}
   \caption{The evolution of states for increasing $K$ for the
-  \protect\sustain{} MOOC.} % I don't know why \protect is needed
+  \protect\sustain{} MOOC.\ State ``names'' indicate the highest
+  probability action(s) within a state.} % I don't know why \protect is needed
   \label{fig:sustain-state-evolution}
@@
 We observe similar behavior in Figure~\ref{fig:sustain-state-evolution}
-as we increase $K$ when fitting our TL-HMM on the \sustain{} MOOC. Again,
+as we increase $K$ when fitting our TL-HMM on the \sustain{} MOOC.\ Again,
 in the transition between $K=2$ and $K=3$ we discover forum behavior
@@
 $K=4$ we instead see a refining of that discovered forum state, where
-state 2 captures asymmetric transition probabilities between ``forum thread
+state 3 captures asymmetric transition probabilities between ``forum thread
 view'' and ``forum thread list''. These states could be seen as redundant,
-in which case a more appropriate setting of $K=3$ may be appropriate for
-the \sustain{} dataset.
+in which case a  setting of $K=3$ may be more appropriate for the
+\sustain{} dataset.
 
@@
 state transition diagram in Figure~\ref{fig:trans-perfect}. We can clearly
-see that the probability of state 2 has increased dramatically, while the
-probability of states 0 and 3 has decreased. State 1 had its probability
-increase, but only very slightly.
+see that the probability of state 2 has increased dramatically, consistent
+with previous observations of the positive correlation between forum activity and
+grades~\cite{Huang:2014:LAS}, while the probability of states 0 and 3 has
+decreased. State 1 had its probability increase, but only very slightly.
 

### 2017-01-14 sec-results.tex
Switch to JEDM theme for a EDM 2017 Journal Track submission.
Update all figures for adjusted font size.

@@
-\newcommand{\textretrieval}{\preprintonly{textretrieval-001}%
-    \finalonly{textretrieval-001}%
-    \submissiononly{moocname1-00X}}
+\newcommand{\textretrieval}{textretrieval-001}
 
-\newcommand{\sustain}{\preprintonly{sustain-001}%
-    \finalonly{sustain-001}%
-    \submissiononly{moocname2-00X}}
+\newcommand{\sustain}{sustain-001}
 
-\newcommand{\UIUC}{\preprintonly{UIUC}%
-    \finalonly{UIUC}%
-    \submissiononly{(redacted University)}}
+\newcommand{\UIUC}{UIUC}
 
@@
   \end{subfigure}
-  \caption{Two example states found by the 6-state TL-HMM.}
+  \caption{Two example states found by the 6-state TL-HMM. (The naming of
+  these states reflects our own interpretation.)}
   \label{fig:states}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+  \begin{subfigure}[t]{0.5\textwidth}
+    \includegraphics[width=\textwidth]{figures/text-6state/state4.png}
+    \caption{A state from \textretrieval{}.}
+  \end{subfigure}%
+  \begin{subfigure}[t]{0.5\textwidth}
+    \includegraphics[width=\textwidth]{figures/sustain-6state/state4.png}
+    \caption{A state from \sustain{}.}
+  \end{subfigure}
+  \caption{Two similar example states found by the 6-state TL-HMM.}
+  \label{fig:cross-course}
 \end{figure*}
@@
 
-\begin{figure*}
-  \centering
-  \begin{subfigure}[t]{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figures/text-6state/state4.png}
-    \caption{A state from \textretrieval{}.}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figures/sustain-6state/state4.png}
-    \caption{A state from \sustain{}.}
-  \end{subfigure}
-  \caption{Two similar example states found by the 6-state TL-HMM.}
-  \label{fig:cross-course}
-\end{figure*}
-
 We can also use our model for cross-course behavior analysis. In
@@
 
+\afterpage{%
+\begin{landscape}
 \begin{figure*}
@@
     & \textbf{State 2 (Thread View)} & \textbf{State 3 (Lecture/Wiki)}\\\hline
-    \includegraphics[width=0.22\textwidth]{figures/text-2state/state0.png}
+    \includegraphics[width=0.30\textwidth]{figures/text-2state/state0.png}
     &
-    \includegraphics[width=0.22\textwidth]{figures/text-2state/state1.png}\\
-    \includegraphics[width=0.22\textwidth]{figures/text-3state/state0.png}
+    \includegraphics[width=0.30\textwidth]{figures/text-2state/state1.png}\\
+    \includegraphics[width=0.30\textwidth]{figures/text-3state/state0.png}
     &
-    \includegraphics[width=0.22\textwidth]{figures/text-3state/state1.png}
+    \includegraphics[width=0.30\textwidth]{figures/text-3state/state1.png}
     &
-    \includegraphics[width=0.22\textwidth]{figures/text-3state/state2.png}\\
-    \includegraphics[width=0.22\textwidth]{figures/text-4state/state0.png}
+    \includegraphics[width=0.30\textwidth]{figures/text-3state/state2.png}\\
+    \includegraphics[width=0.30\textwidth]{figures/text-4state/state0.png}
     &
-    \includegraphics[width=0.22\textwidth]{figures/text-4state/state1.png}
+    \includegraphics[width=0.30\textwidth]{figures/text-4state/state1.png}
     &
-    \includegraphics[width=0.22\textwidth]{figures/text-4state/state2.png}
+    \includegraphics[width=0.30\textwidth]{figures/text-4state/state2.png}
     &
-    \includegraphics[width=0.22\textwidth]{figures/text-4state/state3.png}
+    \includegraphics[width=0.30\textwidth]{figures/text-4state/state3.png}
   \end{tabular}
@@
 \end{figure*}
+\end{landscape}
+}
 
-In Figure~\ref{fig:text-state-evolution} we see the evolution of the latent
-state representations found for the \textretrieval{} MOOC\footnote{Due to
-space constraints, the graphs are quite small, but the states are in the
-same positions as they were in previous figures to aid interpretation.}.
-With $K=2$ we uncover a video watching pattern (state 1) and a course
-material browsing pattern (state 0). However, we can see that when $K=3$ we
-are able to uncover a new pattern involving forum behavior in state 3
-(notice the node sizes on the bottom right). As we increase $K$ to four, we
-can see that state 1 splits into state 1 and state 3. These states appear
-quite similar at a glance, but there are still some key differences. First,
-we can see that state 1 now has a non-negligible forum component, whereas
-state 3 has hardly any weight on the forum.
-
+\afterpage{%
+\begin{landscape}
 \begin{figure*}
@@
     & \textbf{State 3 (Thread View)}\\\hline
-    \includegraphics[width=0.22\textwidth]{figures/sustain-2state/state0.png}
+    \includegraphics[width=0.30\textwidth]{figures/sustain-2state/state0.png}
     &
-    \includegraphics[width=0.22\textwidth]{figures/sustain-2state/state1.png}\\
-    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state0.png}
+    \includegraphics[width=0.30\textwidth]{figures/sustain-2state/state1.png}\\
+    \includegraphics[width=0.30\textwidth]{figures/sustain-3state/state0.png}
     &
-    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state2.png}
+    \includegraphics[width=0.30\textwidth]{figures/sustain-3state/state2.png}
     &
-    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state1.png}\\
-    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state0.png}
+    \includegraphics[width=0.30\textwidth]{figures/sustain-3state/state1.png}\\
+    \includegraphics[width=0.30\textwidth]{figures/sustain-4state/state0.png}
     &
-    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state3.png}
+    \includegraphics[width=0.30\textwidth]{figures/sustain-4state/state3.png}
     &
-    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state1.png}
+    \includegraphics[width=0.30\textwidth]{figures/sustain-4state/state1.png}
     &
-    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state2.png}
+    \includegraphics[width=0.30\textwidth]{figures/sustain-4state/state2.png}
   \end{tabular}
@@
 \end{figure*}
+\end{landscape}
+}
+
+In Figure~\ref{fig:text-state-evolution} we see\footnote{Due to space
+constraints, the graphs are quite small, but the states are in the same
+positions as they were in previous figures to aid interpretation.} the
+evolution of the latent state representations found for the
+\textretrieval{} MOOC.  With $K=2$ we uncover a video watching pattern
+(state 1) and a course material browsing pattern (state 0). However, we can
+see that when $K=3$ we are able to uncover a new pattern involving forum
+behavior in state 3 (notice the node sizes on the bottom right). As we
+increase $K$ to four, we can see that state 1 splits into state 1 and state
+3. These states appear quite similar at a glance, but there are still some
+key differences. First, we can see that state 1 now has a non-negligible
+forum component, whereas state 3 has hardly any weight on the forum.
 
@@
 \subsection{Transitions Between Latent States}
-
-\begin{figure*}
-  \centering
-  \begin{subfigure}[t]{0.33\textwidth}
-    \includegraphics[width=\textwidth,trim={0 3cm 0 2cm}]{figures/trans-comp/trans-avg.png}
-    \caption{\label{fig:trans-avg}}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.33\textwidth}
-    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{figures/trans-comp/trans-perfect.png}
-    \caption{\label{fig:trans-perfect}}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.33\textwidth}
-    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{figures/trans-comp/trans-low.png}
-    \caption{\label{fig:trans-low}}
-  \end{subfigure}
-  \caption{The latent state transition diagrams for a 4-state TL-HMM fit to
-  \protect\textretrieval{} for all students (a) compared to only ``perfect''
-  students (b) and only ``low'' students (c).}
-  \label{fig:trans-comp}
-\end{figure*}
 
@@
 also significantly lower than state 1.
+
+\begin{figure*}
+  \centering
+  \begin{subfigure}[t]{0.33\textwidth}
+    \includegraphics[width=\textwidth,trim={0 3cm 0 2cm}]{figures/trans-comp/trans-avg.png}
+    \caption{\label{fig:trans-avg}}
+  \end{subfigure}%
+  \begin{subfigure}[t]{0.33\textwidth}
+    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{figures/trans-comp/trans-perfect.png}
+    \caption{\label{fig:trans-perfect}}
+  \end{subfigure}%
+  \begin{subfigure}[t]{0.33\textwidth}
+    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{figures/trans-comp/trans-low.png}
+    \caption{\label{fig:trans-low}}
+  \end{subfigure}
+  \caption{The latent state transition diagrams for a 4-state TL-HMM fit to
+  \protect\textretrieval{} for all students (a) compared to only ``perfect''
+  students (b) and only ``low'' students (c).}
+  \label{fig:trans-comp}
+\end{figure*}
 

### 2017-01-14 sec-intro.tex
Start writing results section; minor tweaks elsewhere.
Added a bit to the motivation so that we now have 3 lines of motivation: 1) latent states; 2) latent state transitions; and 3) multi-resolution. In our results, we can organize the analysis in similar ways.
Switch to JEDM theme for a EDM 2017 Journal Track submission.
Add additional citation to intro/related work.

@@
       sufficiently by rule-based methods such as those explored by
-      \citet{Kizilcec:2013:LAK}. We instead propose to treat
-      student behavior patterns as being characterized (represented) via a sequence of
-      \emph{latent states}. This allows us to automatically capture
-      patterns that we might not have been able to articulate clearly a
-      priori via a series of rules, and also allows us to model the
-      inherent uncertainty in assigning a student's behavior to a
-      particular pattern or group. 
+      \citet{Kizilcec:2013:LAK} and \citet{Davis:2016:EDM}. We instead
+      propose to treat student behavior patterns as being characterized
+      (represented) via a sequence of \emph{latent states}. This allows us
+      to automatically capture patterns that we might not have been able to
+      articulate clearly a priori via a series of rules, and also allows us
+      to model the inherent uncertainty in assigning a student's behavior
+      to a particular pattern or group.
   \item Student behavior can vary over time. Previous models that treat students
@@
       time~\cite{Faucon:2016:EDM} miss out on the opportunity to understand
-      student behavior dynamics in a course. We propose a latent space model to flexibly model the dynamics. 
+      student behavior dynamics in a course. We propose a latent space
+      model with {\em latent state transitions} to flexibly model the
+      dynamics.
   \item Analysis of student behavior can and should be performed at varying
@@
       with a particular assumption about the resolution of time they
-      consider~\cite{Faucon:2016:EDM, Kizilcec:2013:LAK, Shih:2010:EDM}. We propose a more flexible model to accommodate different levels of resolution. 
+      consider~\cite{Faucon:2016:EDM, Kizilcec:2013:LAK, Shih:2010:EDM}. We
+      propose a more flexible model to accommodate different levels of
+      resolution.
 \end{enumerate}

### 2017-01-14 sec-related-work.tex
Add additional citation to intro/related work.

@@
 
-HMMs or similar ideas have been previously 
-applied to model education data~\cite{Shih:2010:EDM,Kizilcec:2013:LAK}, but the previous models are not well tuned toward the student behavior task and thus cannot
-adequately address all the aspects of complexity of student learning behavior. 
-A main technical contribution of this paper is to propose a more general
-HMM that can better adapt to the variations of student behavior via its variable resolution
-and nested HMM structure, and thus enable discovery of more sophisticated behavior patterns and provide more detailed characterization of student behavior than the previous models.
+HMMs or similar ideas have been previously applied to model education
+data~\cite{Shih:2010:EDM,Kizilcec:2013:LAK,Davis:2016:EDM}, but the
+previous models are not well tuned toward the student behavior task and
+thus cannot adequately address all the aspects of complexity of student
+learning behavior.  A main technical contribution of this paper is to
+propose a more general HMM that can better adapt to the variations of
+student behavior via its variable resolution and nested HMM structure, and
+thus enable discovery of more sophisticated behavior patterns and provide
+more detailed characterization of student behavior than the previous
+models.
 
-For example, \citet{Kizilcec:2013:LAK} assigned students to states following a rule-based
-approach based upon when the student submitted the assignment for a
-particular week in the course. They investigated how students transitioned
-between these states as the course progressed, and used the sequence of
-states a student exhibited as a representation for performing $k$-means
-clustering of students into related groups. This differs from our method
-substantially: we assign students to states using a probabilistic
-framework to account for uncertainty in this state assignment, and jointly
-learn representations for these states, which are treated as being
-\emph{latent} as opposed to pre-definied using some rule (or set of rules).
-Furthermore, our model provides more flexibility in how the time segments
-are defined, allowing for both finer (for example, day-by-day) or courser
-(for example, month-by-month) granularity. \citet{Shih:2010:EDM} investigated the use of HMM-based clustering
-techniques for automatic discovery of student learning strategies when
-solving a particular problem. This is similar to our approach in that the
-description of behavior profiles is a Markov model, but cannot further characterize
-each latent state with another informative HMM. Thus their work can be regarded
-to modeling ``micro'' behavior, whereas our model can model both ``micro'' and ``macro'' behavior.
+For example, \citet{Kizilcec:2013:LAK} assigned students to states
+following a rule-based approach based upon when the student submitted the
+assignment for a particular week in the course. They investigated how
+students transitioned between these states as the course progressed, and
+used the sequence of states a student exhibited as a representation for
+performing $k$-means clustering of students into related groups. This
+differs from our method substantially: we assign students to states using a
+probabilistic framework to account for uncertainty in this state
+assignment, and jointly learn \emph{representations} for these states,
+which are treated as being \emph{latent} as opposed to pre-definied using
+some rule (or set of rules).  Furthermore, our model provides more
+flexibility in how the time segments are defined, allowing for both finer
+(for example, day-by-day) or courser (for example, month-by-month)
+granularity. \citet{Shih:2010:EDM} investigated the use of HMM-based
+clustering techniques for automatic discovery of student learning
+strategies when solving a particular problem.  This is similar to our
+approach in that the description of behavior profiles is a Markov model,
+but cannot further characterize each latent state with another informative
+HMM. Thus their work can be regarded to modeling ``micro'' behavior,
+whereas our model can model both ``micro'' and ``macro'' behavior.
+
+\citet{Davis:2016:EDM} investigate frequent student behavior pattern chains
+with a set of actions that is defined similarly to ours. However, their
+method for finding the common behavioral patterns involves a manual
+clustering step to identify behavioral ``motifs'', which is then followed
+by an automatic (rule-based) assignment of all sequences to these motifs.
+Our method, by contrast, attempts to do this automatically: the latent
+state representations obtained by our model attempt to capture similar
+meanings to their behavior motifs in a completely automated fashion. They
+also automatically generate and investigate Markov models for different
+MOOCs, but do so by considering \emph{all} student action sequences as
+belonging to a \emph{single Markov model}. In our approach, we allow each
+student behavior sequence to belong to one of $K$ different Markov models
+(and further model the transition probabilities between these latent state
+Markov models between each sequence a student generates). Thus, their
+Markov models presented are a special case of our model when $K = 1$.
 

### 2017-01-14 sec-results.tex
Add citation in results section; add explicit probability values.

@@
 probability of taking that edge given that a random walk is currently at
-note $u$ (as indicated by the transition matrix). The plots were created
-using python-igraph\footnote{\url{http://igraph.org/python/}}.
+note $u$ (as indicated by the transition matrix)\footnote{We do not plot
+the transition probabilities directly within the figure to ease
+readability; we instead will mention relevant transition probabilities in
+the text as we discuss the plots. The plots were created using
+python-igraph: \url{http://igraph.org/python/}.}.
 
@@
 
-Figure~\ref{fig:states} includes two such representations we learned. The
-first corresponds to a ``quiz taking'' state whereas the second corresponds
-to a ``lecture viewing'' state. Clearly, our unsupervised method can
-uncover states that do indeed correspond to student behavior modes that we
-would expect to find a priori.
+Figure~\ref{fig:states} includes two such representations we learned. Under
+our interpretation, the first corresponds to a ``quiz taking'' state (it
+has higher ``quiz: start'' and ``quiz: submit'' state probabilities than
+the other five states) whereas the second corresponds to a ``lecture
+viewing'' state. Our unsupervised method can uncover states that do indeed
+correspond to student behavior modes that we would expect to find a priori.
 
@@
 of walking from the ``forum thread view'' action back to the ``forum thread
-list'' action, but in Figure~\ref{fig:lecture-viewing-state} we actually
-observe a very strong link in this direction. This difference highlights an
-important distinction between these two latent states: in the first you are
-more likely to visit the forum \emph{looking for a particular post}, where
-in the second you are more likely to visit the forum \emph{to browse
-existing posts}. Thus, capturing the action transition matrix within a
-latent state is important for capturing detailed insights involving bigrams
-of actions.
+list'' action (see the bottom rightmost arc; transition probability $p
+\approx 0.17$), but in Figure~\ref{fig:lecture-viewing-state} we actually
+observe a much stronger link in this direction (transition probability $p
+\approx 0.63$). This difference highlights an important distinction between
+these two latent states: in the first you are more likely to visit the
+forum \emph{looking for a particular post}, where in the second you are
+more likely to visit the forum \emph{to browse existing posts}. Thus,
+capturing the action transition matrix within a latent state is important
+for capturing detailed insights involving bigrams of actions.
 
@@
 can see that the probability of returning to the course wiki after viewing
-a lecture is considerably lower than that probability in the state from
-\sustain{}. We also notice that the self-loop for staying in a lecture
-activity in \textretrieval{} is significantly higher probability than it is
-in the state from \sustain{}. This gives us some insight into the lecture
-viewing behavior of students in these two MOOCs which can possibly be a
-reflection of the course's structure. In \textretrieval{}, students are
-likely to view lecture videos in succession directly without first visiting
-the course wiki, whereas in \sustain{} students are much more likely to
-first return the course wiki before watching the next lecture video. This
-observation would be lost if we did not consider the transitions between
-the actions within the latent states.
+a lecture (transition probability $p \approx 0.24$) is considerably lower
+than that probability in the state from \sustain{} (transition probability
+$p \approx 0.57$). We also notice that the self-loop for staying in a
+lecture activity in \textretrieval{} ( transition probability $p \approx
+0.70$) is significantly higher probability than it is in the state from
+\sustain{} (transition probability $p \approx 0.34$).  This gives us some
+insight into the lecture viewing behavior of students in these two MOOCs
+which can possibly be a reflection of the course's structure (which, as
+demonstrated in \citet{Davis:2016:EDM}, can differ substantially across
+different MOOCs).  In \textretrieval{}, students are likely to view lecture
+videos in succession directly without first visiting the course wiki,
+whereas in \sustain{} students are much more likely to first return the
+course wiki before watching the next lecture video. This observation would
+be lost if we did not consider the transitions between the actions within
+the latent states.
 
@@
   \caption{The evolution of states for increasing $K$ for the
-  \protect\textretrieval{} MOOC.\ State ``names'' indicate the highest
+  \protect\textretrieval{} MOOC.\ Each row corresponds to the next value of
+  $K$, starting from $K=2$. State ``names'' indicate the highest
   probability action(s) within the state.} % I don't know why \protect is needed
@@
   \caption{The evolution of states for increasing $K$ for the
-  \protect\sustain{} MOOC.\ State ``names'' indicate the highest
+  \protect\sustain{} MOOC.\ Each row corresponds to the next value of $K$,
+  starting from $K=2$. State ``names'' indicate the highest
   probability action(s) within a state.} % I don't know why \protect is needed
@@
 
-In Figure~\ref{fig:text-state-evolution} we see\footnote{Due to space
-constraints, the graphs are quite small, but the states are in the same
-positions as they were in previous figures to aid interpretation.} the
-evolution of the latent state representations found for the
-\textretrieval{} MOOC.  With $K=2$ we uncover a video watching pattern
-(state 1) and a course material browsing pattern (state 0). However, we can
-see that when $K=3$ we are able to uncover a new pattern involving forum
-behavior in state 3 (notice the node sizes on the bottom right). As we
-increase $K$ to four, we can see that state 1 splits into state 1 and state
-3. These states appear quite similar at a glance, but there are still some
-key differences. First, we can see that state 1 now has a non-negligible
-forum component, whereas state 3 has hardly any weight on the forum.
+In Figure~\ref{fig:text-state-evolution} we see\footnote{The graphs are
+quite small, but the states are in the same positions as they were in
+previous figures.} the evolution of the latent state
+representations found for the \textretrieval{} MOOC.  With $K=2$ we uncover
+a video watching pattern (state 1) and a course material browsing pattern
+(state 0). However, we can see that when $K=3$ we are able to uncover a new
+pattern involving forum behavior in state 3 (notice the node sizes on the
+bottom right). As we increase $K$ to four, we can see that state 1 splits
+into state 1 and state 3. These states appear quite similar at a glance,
+but there are still some key differences. First, we can see that state 1
+now has a non-negligible forum component, whereas state 3 has hardly any
+weight on the forum.
 
@@
 large increase in size, where we might have expected state 3 to grow
-instead.  However, there is an alternative explanation for this phenomenon:
+instead.  However, there is an alternative explanation for this phenomenon.
 Since state 1 seems to indicate a highly engaged student, it is a perfectly
@@
 state 1 did not increase in size for the ``perfect'' group like we were
-anticipating. Our model lends itself well to discovering this potentially
-counter-intuitive insight directly from data.
+anticipating. \citet{Kizilcec:2017:CandE} observe similar phenomena with
+the courses they studied where they find that certificate earning is
+negatively correlated with help seeking behavior. Our model lends itself
+well to discovering this potentially counter-intuitive insight directly
+from data.

### 2017-01-14 sec-results.tex
Remove extraneous space.

@@
 $p \approx 0.57$). We also notice that the self-loop for staying in a
-lecture activity in \textretrieval{} ( transition probability $p \approx
+lecture activity in \textretrieval{} (transition probability $p \approx
 0.70$) is significantly higher probability than it is in the state from

### 2017-01-15 sec-results.tex
Add results about mean rank comparisons.

@@
 from data.
+
+To quantify this finding, we perform the following experiment. First, we
+select all students from \textretrieval{} who completed all of the quizzes
+$\mathbf{L}_q \subset \mathbf{L}$. This gives us 1,985 students along with
+their average quiz score. We then create a ranked list of the students in
+$\mathbf{L}_q$ by sorting them by their ``preference'' for a specific
+latent state
+\begin{equation}
+  p_\ell(i) = \frac{\sum_{t=1}^T \gamma_t^{(\mathbf{o}_\ell)}(i)}
+  {\sum_{j=1}^{K} \sum_{t=1}^T \gamma_t^{(\mathbf{o}_\ell)}(j)}
+\end{equation}
+where $\mathbf{o}_\ell$ is the list of action sequences for student $\ell$
+and $\gamma$ is defined as before and computed using the Baum-Welch
+algorithm. We can now compare the average rank in this list for both
+the ``perfect'' student group and the ``low'' student group: a useful state
+for distinguishing the two groups should result in a ranked list with
+statistically significant differences in average rank between the two
+groups. Our results are summarized in Table~\ref{table:mean-rank}. Indeed,
+we discover that states 1 and 2 are useful for teasing apart these two
+groups: state 1 ranks students in the ``low'' group significantly higher
+than those in the ``perfect'' group, and state 2 does the opposite and
+prefers students in the ``perfect'' group to those in the ``low'' group.
+
+\begin{table}
+  \centering
+  \begin{tabular}{r|rrrr}
+    \textbf{Group} & \textbf{State 0} & \textbf{State 1}^\dagger &
+    \textbf{State 2}^\dagger & \textbf{State 3}\\\hline
+    \textbf{Perfect} & 975.3  & 1001.5 & 999.0  & 1056.5\\
+    \textbf{Low}     & 1024.9 & 816.4  & 1230.5 & 1161.2\\
+  \end{tabular}
+  \caption{Average rank for ``perfect'' and ``low'' student groups in the
+  ranked lists associated with the four latent states found by a TL-HMM.
+  $\dagger$ indicates statistically significant different mean ranks at $p
+  < 0.01$ according to an unpaired $t$-test.}
+  \label{table:mean-rank}
+\end{table}

### 2017-01-15 sec-results.tex
Add experiment results for mean rank with 2 -> 2 transition ranking.

@@
 
+\begin{table}
+  \centering
+  \begin{tabular}{r|rrrrr}
+    \textbf{Group} & \textbf{State 0} & \textbf{State 1}^\dagger &
+    \textbf{State 2}^\dagger & \textbf{State 3} & \textbf{State 2
+    $\rightarrow$ 2}^\dagger\\\hline
+    \textbf{Perfect} & 975.3  & 1001.5 & 999.0  & 1056.5 & 939.6\\
+    \textbf{Low}     & 1024.9 & 816.4  & 1230.5 & 1161.2 & 1187.4\\
+  \end{tabular}
+  \caption{Average rank for ``perfect'' and ``low'' student groups in the
+  ranked lists associated with the four latent states found by a TL-HMM.
+  $\dagger$ indicates statistically significant different mean ranks at $p
+  < 0.01$ according to an unpaired $t$-test.}
+  \label{table:mean-rank}
+\end{table}
+
 To quantify this finding, we perform the following experiment. First, we
@@
 groups. Our results are summarized in Table~\ref{table:mean-rank}. Indeed,
-we discover that states 1 and 2 are useful for teasing apart these two
-groups: state 1 ranks students in the ``low'' group significantly higher
+we discover that states 1 and 2 are correlated with the ``perfect'' or
+``low'' groups: state 1 ranks students in the ``low'' group significantly higher
 than those in the ``perfect'' group, and state 2 does the opposite and
@@
 
-\begin{table}
-  \centering
-  \begin{tabular}{r|rrrr}
-    \textbf{Group} & \textbf{State 0} & \textbf{State 1}^\dagger &
-    \textbf{State 2}^\dagger & \textbf{State 3}\\\hline
-    \textbf{Perfect} & 975.3  & 1001.5 & 999.0  & 1056.5\\
-    \textbf{Low}     & 1024.9 & 816.4  & 1230.5 & 1161.2\\
-  \end{tabular}
-  \caption{Average rank for ``perfect'' and ``low'' student groups in the
-  ranked lists associated with the four latent states found by a TL-HMM.
-  $\dagger$ indicates statistically significant different mean ranks at $p
-  < 0.01$ according to an unpaired $t$-test.}
-  \label{table:mean-rank}
-\end{table}
+Returning to Figure~\ref{fig:trans-comp}, we an also see a difference in
+the transitions between the latent states. In particular, look at the
+transitions in Figure~\ref{fig:trans-perfect} and
+Figure~\ref{fig:trans-low} that leave state 2. In the ``perfect'' group,
+nearly all of this probability mass is allocated for the self-loop ($p
+\approx 0.93$). In the ``low'' group, this self loop is less strong ($p
+\approx 0.80$; most easily seen by noting that the edges leaving state 2
+are darker than for the ``perfect'' group). We can perform a similar
+experiment to above by producing a ranked list of students $\ell \in
+\mathbf{L}_q$ by their staying probability for state 2 (that is, given that
+a student $\ell$ is already in state 2, how likely are they to remain there
+in the next action sequence?). This can be computed as
+\begin{equation}
+  p_\ell(2, 2) = \frac{\sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,2)}
+  {\sum_{i=1}^K \sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,i)}
+\end{equation}
+where $\xi$ is defined as before and computed using the Baum-Welch
+algorithm. The last column of Table~\ref{table:mean-rank} indicates that
+this transition feature also correlates with the achievement group and
+results in significant differences in mean rank.

### 2017-01-15 sec-model.tex
Switch to JEDM theme for a EDM 2017 Journal Track submission.
Fix extraneous alignment character.
Address Cheng's comments.

@@
 
-\subsection{Formal Definition of TL-HMM}
+\subsection{Formal Definition of the TL-HMM}
 
@@
 with action $a$ and $P(a_i \mid a_{i-1}, A^{(u)}) = A^{(u)}_{a_{i-1}, a_i}$
-is the transition probability of moving from action $a_{i-1}$ to $a_{i}$.
+is the transition probability of moving from action $a_{i-1}$ to $a_{i}$
+given that the model is currently in latent state $u$.
 
@@
 $\mathbf{S}_\ell$ of length $N$ for a student $\ell$ by marginalizing over
-all possible latent state sequences $U=(v_1, \ldots, v_N)$ as
+all possible latent state sequences $(v_1, \ldots, v_N) \in U$ as
 \begin{equation}
-  \begin{split}
   P(\mathbf{S}_\ell \mid \Lambda)
-  &= \sum_U \left(P(v_1 \mid \Lambda) P(\mathbf{s}_1 \mid \lambda^{(v_1)})
-  \phantom{\prod_{i=2}^N}\right.\\
-  &\quad \left.\times \prod_{i=2}^N P(v_i \mid v_{i-1}, \Lambda)
-  P(\mathbf{s}_i \mid \lambda^{(v_i)})\right)
-  \end{split}
+  = \sum_{(v_1, \ldots, v_N) \in U} \left(P(v_1 \mid \Lambda)
+  P(\mathbf{s}_1 \mid \lambda^{(v_1)}) \times \prod_{i=2}^N P(v_i \mid
+  v_{i-1}, \Lambda) P(\mathbf{s}_i \mid \lambda^{(v_i)})\right)
 \end{equation}

### 2017-01-15 sec-results.tex
Address Cheng's comments.

@@
 \subsection{Latent State Representations}
-First, we fit a 6-state TL-HMM to the \textretrieval{} sequence dataset and
-show some of the latent state representations we find. We used the
-following ten actions as our action set $\mathbf{A}$:
+The TL-HMM is meant to be a tool for exploratory analysis of student
+behavior. As such, the number of states should be empirically set based on
+the goal of analysis. A higher number of states will lead to a
+finer-grained modeling of student behavior. In our experiments, we explored
+using between 2--6 states. First, we fit a 6-state TL-HMM to the
+\textretrieval{} sequence dataset and show some of the latent state
+representations we find. We used the following ten actions as our action
+set $\mathbf{A}$:
 \begin{enumerate*}[label=(\arabic*)]

### 2017-02-22 short-paper/clickstream-hmm-short.tex
Add L@S17 short paper.

\documentclass{sigchi-ext}

% Please be sure that you have the dependencies (i.e., additional
% LaTeX packages) to compile this example.
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage[scaled=.92]{helvet} % for proper fonts
\usepackage{graphicx} % for EPS use the graphics package instead
\usepackage{balance}  % for useful for balancing the last columns
\usepackage{booktabs} % for pretty table rules
\usepackage{ccicons}  % for Creative Commons citation icons
\usepackage{ragged2e} % for tighter hyphenation

% Some optional stuff you might like/need.
% \usepackage{marginnote}
% \usepackage[shortlabels]{enumitem}
% \usepackage{paralist}
\usepackage[utf8]{inputenc} % for a UTF8 editor only
\usepackage{amsmath}
\usepackage{float}
\usepackage[numbers]{natbib}

\usepackage{subcaption}
\usepackage{pdflscape}
\usepackage{afterpage}

\renewcommand\cite{\citep}
\def\meta/{\textsc{MeTA}}
\newcommand{\textretrieval}{textretrieval-001}
\newcommand{\sustain}{sustain-001}
\newcommand{\UIUC}{UIUC}

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP --
% \copyrightinfo{Permission to make digital or hard copies of all or
% part of this work for personal or classroom use is granted without
% fee provided that copies are not made or distributed for profit or
% commercial advantage and that copies bear this notice and the full
% citation on the first page. Copyrights for components of this work
% owned by others than ACM must be honored. Abstracting with credit is
% permitted. To copy otherwise, or republish, to post on servers or to
% redistribute to lists, requires prior specific permission and/or a
% fee. Request permissions from permissions@acm.org.\\
% {\emph{CHI'14}}, April 26--May 1, 2014, Toronto, Canada. \\
% Copyright \copyright~2014 ACM ISBN/14/04...\$15.00. \\
% DOI string from ACM form confirmation}
%% EXAMPLE END

\copyrightinfo{Permission to make digital or hard copies of all or part of
  this work for personal or classroom use is granted without fee provided
  that copies are not made or distributed for profit or commercial
  advantage and that copies bear this notice and the full citation on the
  first page. Copyrights for components of this work owned by others than
  the author(s) must be honored. Abstracting with credit is permitted. To
  copy otherwise, or republish, to post on servers or to redistribute to
  lists, requires prior specific permission and/or a fee. Request
  permissions from Permissions@acm.org.\\
  \emph{L@S 2017}, April 20 - 21, 2017, Cambridge, MA, USA.\\
  Copyright is held by the owner/author(s). Publication rights licensed to
  ACM.\\
  ACM 978-1-4503-4450-0/17/04 \$15.00\\
  DOI: \url{http://dx.doi.org/10.1145/3051457.3053986}}

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Modeling MOOC Student Behavior with Two-Layer Hidden Markov
Models}
\def\plainauthor{Chase Geigle, ChengXiang Zhai}
\def\emptyauthor{}
\def\plainkeywords{MOOC log analysis; studetn behavior modeling; Markov models; hidden Markov models}
\def\plaingeneralterms{\plainkeywords}

\title{\plaintitle}

\numberofauthors{2}
% Notice how author names are alternately typesetted to appear ordered
% in 2-column format; i.e., the first 4 autors on the first column and
% the other 4 auhors on the second column. Actually, it's up to you to
% strictly adhere to this author notation.
\author{%
  \alignauthor{%
    \textbf{Chase Geigle}\\
    \affaddr{University of Illinois at Urbana-Champaign} \\
    \affaddr{201 N Goodwin Ave}\\
    \affaddr{Urbana, IL 61801} \\
    \email{geigle1@illinois.edu} } \vfil \alignauthor{%
    \textbf{ChengXiang Zhai}\\
    \affaddr{University of Illinois at Urbana-Champaign} \\
    \affaddr{201 N Goodwin Ave}\\
    \affaddr{Urbana, IL 61801} \\
    \email{czhai@illinois.edu} } }

% Make sure hyperref comes last of your loaded packages, to give it a
% fighting chance of not being over-written, since its job is to
% redefine many LaTeX commands.
\definecolor{linkColor}{RGB}{6,125,233}
\hypersetup{%
  pdftitle={\plaintitle},
%  pdfauthor={\plainauthor},
  pdfauthor={\emptyauthor},
  pdfkeywords={\plainkeywords},
  bookmarksnumbered,
  pdfstartview={FitH},
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=black,
  urlcolor=linkColor,
  breaklinks=true,
}

% \reversemarginpar%

\newcommand{\citemeta}{\cite{Massung:2016:ACL}}

\begin{document}

\maketitle

% Uncomment to disable hyphenation (not recommended)
% https://twitter.com/anjirokhan/status/546046683331973120
\RaggedRight{}

% Do not change the page size or page settings.
\begin{abstract}
  Massive open online courses (MOOCs) provide educators with an abundance
  of data describing how students interact with the platform, but this data
  is highly underutilized today. This is in part due to the lack of
  sophisticated tools to provide interpretable and actionable summaries of
  huge amounts of MOOC activity present in log data. In this paper, we
  propose a method for automatically discovering student behavior patterns
  by leveraging the click log data that can be obtained from the MOOC
  platform itself in a completely unsupervised manner.
\end{abstract}

\keywords{\plainkeywords}

\begin{CCSXML}
  <ccs2012>
    <concept>
      <concept_id>10002950.10003648.10003670.10003683</concept_id>
      <concept_desc>Mathematics of computing~Kalman filters and hidden Markov models</concept_desc>
      <concept_significance>500</concept_significance>
    </concept>
    <concept>
      <concept_id>10010147.10010257.10010258.10010260.10010267</concept_id>
      <concept_desc>Computing methodologies~Mixture modeling</concept_desc>
      <concept_significance>500</concept_significance>
    </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Mathematics of computing~Kalman filters and hidden Markov models}
\ccsdesc[500]{Computing methodologies~Mixture modeling}

\printccsdesc{}

\input{sec-intro}
\input{sec-model}
\input{sec-results}
\input{sec-conclusion}

\section{Acknowledgments}

This material is based upon work supported by the NSF GRFP under Grant
Number DGE-1144245.

\balance{}

\bibliographystyle{acm-sigchi-modern}
\bibliography{bib}

\end{document}

### 2017-02-22 short-paper/sec-conclusion.tex
Add L@S17 short paper.

\section{Conclusions and Future Directions}
We proposed a two-layer hidden Markov model for MOOC student behavior
modeling on large collections of actions extracted from MOOC log data. This
model uniquely treats behavior patterns as a sequence of \emph{latent states},
rather than assigning these states in a rule-based manner, captures the
variable behaviors of students over time, and allows analysis at different
levels of granularity.

Although we only experimented with our model on one MOOC, the model is
completely general and can be easily applied to analyze the log of any
other course to enable deep understanding of student behaviors as well as
the correlations of such behaviors as other variables such as grades.
Our model can be used to produce an ordered list of latent states for each
student over time from which we could correlate course-wide drifts with
events in a course, which could for example allow the automatic discovery
of difficult segments of a course.

### 2017-02-22 short-paper/sec-intro.tex
Add L@S17 short paper.

--- 2016-10-25 sec-intro.tex
@@
 
-What stands in the way? Clearly, intelligent systems to create concise and
-digestible summaries of the massive amount of interaction data collected
-are needed in order to empower the instructors of these courses. If we can
-understand how users are interacting with our MOOCs, we are much more
-likely to be able to make changes to these courses that positively impact
-learners. While we can easily observe the changes in behavior of students
-in real classrooms, MOOCs present a challenge due to their hands-off
-nature. We view this paper as attempting to bridge this gap. Specifically,
-in this paper we explore unsupervised learning methods for automatically
-detecting student learning behavior patterns or profiles from large
-collections of click logs associated with MOOCs.
-
-Our work is motivated by the following beliefs:
-\begin{enumerate}
-  \item Student behavior is complicated and cannot necessarily be captured
-      sufficiently by rule-based methods such as those explored by
-      \citet{Kizilcec:2013:LAK}. We instead propose to treat
-      student behavior patterns as being generated via a sequence of
-      \emph{latent states}. This allows us to automatically capture
-      patterns that we might not have been able to articulate clearly a
-      priori via a series of rules, and also allows us to model the
-      inherent uncertainty in assigning a student's behavior to a
-      particular pattern or group.
-  \item Student behavior can vary over time. Previous models that treat students
-      as exhibiting only one behavioral pattern over
-      time~\cite{Faucon:2016:EDM} miss out on the opportunity to understand
-      student behavior dynamics in a course.
-  \item Analysis of student behavior can and should be performed at varying
-      levels of granularity. This requires us to aggregate data over time
-      with \emph{different levels of resolution}; existing models tend to come
-      with a particular assumption about the resolution of time they
-      consider~\cite{Faucon:2016:EDM, Kizilcec:2013:LAK, Shih:2010:EDM}.
-\end{enumerate}
-
-Thus, we want a \emph{latent variable approach} to mining student behavior
-patterns that is \emph{probabilistic} for inference, and that should be
-\emph{flexible to model state changes over different time resolutions}. In
-this paper, we propose to discover latent student behavior patterns via
-unsupervised learning on large collections of student behavior observation
-sequences via a two-layer hidden Markov model.
+In this paper, we propose a novel two-layer hidden Markov model
+(TL-HMM) to discover latent student behavior patterns via unsupervised
+learning on large collections of student behavior observation sequences.
+This \emph{latent variable approach} to mining student behavior
+patterns that is \emph{probabilitic} for inference and \emph{flexible to
+model state changes over different time resolutions} gives our model
+a unique perspective compared to previous work that uses rule-based
+approaches~\citep{Kizilcec:2013:LAK,Davis:2016:EDM} and work that comes
+with a particular assumption about the resolution of times
+considered~\citep{Faucon:2016:EDM, Kizilcec:2013:LAK}.

### 2017-02-22 short-paper/sec-model.tex
Add L@S17 short paper.

\section{A Two-Layer HMM for MOOC Log Analysis}

\begin{margintable}
  \vspace{-4.7in}
  \begin{center}
    \begin{tabular}{rr}
      \multicolumn{2}{c}{\textbf{\textretrieval{}}}\\\hline
      Students & 18,941\\
      Sequences & 85,240\\
      Avg. Len. & 7.31
    \end{tabular}
    \caption{Statistics about the sequences extracted from the MOOC.}
    \label{table:datasets}
  \end{center}
\end{margintable}

\marginpar{%
  \vspace{-1.6in}
  Each sequence is comprised of the following actions:
  \begin{enumerate}[itemsep=0,align=left,leftmargin=*]
    \item quiz start,
    \item quiz submit,
    \item wiki (course material),
    \item forum list (view the list of all forums),
    \item forum thread list (view the list of all threads in a specific
      forum),
    \item forum thread view (view the list of posts within a specific
      thread),
    \item forum search (a search query issued against the forum),
    \item forum post thread (a new thread was posted),
    \item forum post reply (a new post was created within an existing
      thread), and
    \item view lecture (defined as either streaming or downloading a lecture
      video).
  \end{enumerate}
}

Our general idea is to use a probabilistic generative model to model the
student activities as recorded in a MOOC log, which means we will assume
that all the observed student activities are samples drawn (i.e.,
``generated'') from a parameterized probabilistic model. We can then
estimate the parameter values of the probabilistic model by fitting the
model to a specific MOOC log data set. The estimated parameter values could
then be treated as the latent ``knowledge'' discovered from the data.
Because such a generative model attempts to fit {\em all} the data, it
enables us to discover interesting patterns that can explain the {\em
overall} behavior of a student or the {\em common} behavior patterns shared
by many students.

An HMM is a specific probabilistic generative model with a ``built-in''
state transition system that would control the data to be generated by the
model, thus it is especially suitable for modeling sequence
data. We may say that students transition through different ``task states''
(or ``behavior states'')  in the process of study.  One such task state may
be to learn about a topic by mostly watching lecture videos, another task
state may be to work on quizzes, and yet another may be to participate in
forum discussions. While in each of these different states, the student
would tend to exhibit different surface ``micro'' behaviors.  For example,
in the lecture study state, the student would tend to have many
video-watching related behaviors and occasionally forum activities, while
in the quiz-taking state, the student would tend to show many quiz-related
``micro'' activities as well as asking questions or checking discussions on
the forum.

The most natural application of an HMM to analyze our data would treat each
latent state as having a discrete distribution over the surface ``micro''
behaviors, but this only gives a rather superficial characterization of the
student behavior. Instead, we want to directly characterize the transitions
between the ``micro'' actions, so we treat an {\em entire
sequence} of micro activities (e.g., one session of activities) as an
observed ``symbol'' from a latent state, and further model the generation
of such a sequence with another Markov model where we treat each micro
activity as an {\em observable} state, and model the transitions between
these activity states in very much the same way as the state transitions in
HMM. Adding this second layer would allow us to characterize a latent task
state in much more detail, as it would reveal not only what activities are
most common to a task state, but also the transition patterns between these
activities.

### 2017-02-22 short-paper/sec-results.tex
Add L@S17 short paper.

--- 2017-01-15 sec-results.tex
@@
-\newcommand{\textretrieval}{textretrieval-001}
+\section{Experiment Results}
 
-\newcommand{\sustain}{sustain-001}
-
-\newcommand{\UIUC}{UIUC}
-
-\section{Experiment Results}
 As an analysis tool, the TL-HMM model provides with us the following two
@@
 qualitatively analyze both types of patterns discovered from empirical MOOC
-log data.
+log data. Specifically, we look at the MOOC logs associated with the
+\textretrieval{} Coursera MOOC offered by \UIUC{}.
+Table~\ref{table:datasets} provides statistics about the sequences we
+extracted from this MOOC and the action space we used for those sequences.
 
-Specifically, we look at the MOOC logs associated with two different
-Coursera MOOCs offered by \UIUC{}: \textretrieval{} and \sustain{}.  The
-\textretrieval{} MOOC represents a highly technical computer science course,
-where the \sustain{} MOOC is more representative of a humanities course. We
-picked these two MOOCs because of their vastly different content domains.
-Table~\ref{table:datasets} summarizes the two datasets we extracted from
-the MOOCs.
+\begin{marginfigure}
+  %\vspace{-2in}
+  \centering
+  \begin{subfigure}[t]{\textwidth}
+    \includegraphics[width=0.95\textwidth,trim={0 3cm 0 2cm}]{../figures/trans-comp/trans-avg.png}
+    \caption{\vspace{20pt}\label{fig:trans-avg}}
+  \end{subfigure}
 
-\begin{table}
-  \begin{center}
-    \begin{tabular}{rrrr}
-      \textbf{MOOC} & \textbf{Students} & \textbf{Sequences} & \textbf{Avg.
-      $|\mathbf{s}|$}\\\hline
-      \textretrieval{} & 18,941 & 85,240 & 7.31\\
-      \sustain{} & 85,240 & 231,881 & 15.4
-    \end{tabular}
-    \caption{Statistics about the sequences extracted from the two MOOCs.}
-    \label{table:datasets}
-  \end{center}
-\end{table}
+  \begin{subfigure}[t]{\textwidth}
+    \includegraphics[width=0.95\textwidth,trim={0, 3cm 0 2cm}]{../figures/trans-comp/trans-perfect.png}
+    \caption{\vspace{20pt}\label{fig:trans-perfect}}
+  \end{subfigure}
 
-\subsection{Latent State Representations}
-First, we fit a 6-state TL-HMM to the \textretrieval{} sequence dataset and
-show some of the latent state representations we find. We used the
-following ten actions as our action set $\mathbf{A}$:
-\begin{enumerate*}[label=(\arabic*)]
-  \item quiz start,
-  \item quiz submit,
-  \item wiki (course material),
-  \item forum list (view the list of all forums),
-  \item forum thread list (view the list of all threads in a specific
-    forum),
-  \item forum thread view (view the list of posts within a specific
-    thread),
-  \item forum search (a search query issued against the forum),
-  \item forum post thread (a new thread was posted),
-  \item forum post reply (a new post was created within an existing
-    thread), and
-  \item view lecture (defined as either streaming or downloading a lecture
-    video).
-\end{enumerate*}
+  \begin{subfigure}[t]{\textwidth}
+    \includegraphics[width=0.95\textwidth,trim={0, 3cm 0 2cm}]{../figures/trans-comp/trans-low.png}
+    \caption{\label{fig:trans-low}}
+  \end{subfigure}
 
-To visualize these Markov models that represent our latent states, we plot
-them as a directed graph where we set the size of a node to be proportional
-to its personalized Pagerank score~\cite{Page:1999:PageRank, Jeh:2003:WWW}
-where the personalization vector is the initial state distribution for the
-Markov model. We let the thickness of a directed edge $(u, v)$ reflect the
-probability of taking that edge given that a random walk is currently at
-note $u$ (as indicated by the transition matrix)\footnote{We do not plot
-the transition probabilities directly within the figure to ease
-readability; we instead will mention relevant transition probabilities in
-the text as we discuss the plots. The plots were created using
-python-igraph: \url{http://igraph.org/python/}.}.
+  \caption{The latent state transition diagrams for a 4-state TL-HMM fit to
+  \protect\textretrieval{} for all students (a) compared to only ``perfect''
+  students (b) and only ``low'' students (c).}
+  \label{fig:trans-comp}
+\end{marginfigure}
 
@@
   \centering
-  \begin{subfigure}[t]{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figures/text-6state/state0.png}
-    \caption{An example ``quiz taking'' state.}
-    \label{fig:practice-quiz-state}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figures/text-6state/state1.png}
-    \caption{An example ``lecture viewing'' state.}
-    \label{fig:lecture-viewing-state}
+  \begin{subfigure}[t]{0.24\textwidth}
+    \includegraphics[width=\textwidth]{../figures/text-4state/state0.png}
+    \caption{\label{fig:state0}State 0}
   \end{subfigure}
-  \caption{Two example states found by the 6-state TL-HMM. (The naming of
-  these states reflects our own interpretation.)}
+  \begin{subfigure}[t]{0.24\textwidth}
+    \includegraphics[width=\textwidth]{../figures/text-4state/state1.png}
+    \caption{\label{fig:state1}State 1}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.24\textwidth}
+    \includegraphics[width=\textwidth]{../figures/text-4state/state2.png}
+    \caption{\label{fig:state2}State 2}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.24\textwidth}
+    \includegraphics[width=\textwidth]{../figures/text-4state/state3.png}
+    \caption{\label{fig:state3}State 3}
+  \end{subfigure}
+  \caption{Example states learned by a 4-state TL-HMM.}
   \label{fig:states}
@@
 
-\begin{figure*}
-  \centering
-  \begin{subfigure}[t]{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figures/text-6state/state4.png}
-    \caption{A state from \textretrieval{}.}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figures/sustain-6state/state4.png}
-    \caption{A state from \sustain{}.}
-  \end{subfigure}
-  \caption{Two similar example states found by the 6-state TL-HMM.}
-  \label{fig:cross-course}
-\end{figure*}
-
-Figure~\ref{fig:states} includes two such representations we learned. Under
-our interpretation, the first corresponds to a ``quiz taking'' state (it
-has higher ``quiz: start'' and ``quiz: submit'' state probabilities than
-the other five states) whereas the second corresponds to a ``lecture
-viewing'' state. Our unsupervised method can uncover states that do indeed
-correspond to student behavior modes that we would expect to find a priori.
-
-We also argue that it is important that the latent state representation be
-a Markov model rather than just a discrete distribution over actions in
-$\mathbf{A}$ (as would be the case for a traditional single-layer HMM). We
-can observe why if we take a closer look at each of the two latent state
-representations in Figure~\ref{fig:states} and look at their forum
-component (bottom right). We can see that the relative probability of the
-forum activities is roughly the same between these states, but the
-\emph{transitions} are quite different. In
-Figure~\ref{fig:practice-quiz-state} we have a relatively low probability
-of walking from the ``forum thread view'' action back to the ``forum thread
-list'' action (see the bottom rightmost arc; transition probability $p
-\approx 0.17$), but in Figure~\ref{fig:lecture-viewing-state} we actually
-observe a much stronger link in this direction (transition probability $p
-\approx 0.63$). This difference highlights an important distinction between
-these two latent states: in the first you are more likely to visit the
-forum \emph{looking for a particular post}, where in the second you are
-more likely to visit the forum \emph{to browse existing posts}. Thus,
-capturing the action transition matrix within a latent state is important
-for capturing detailed insights involving bigrams of actions.
-
-We can also use our model for cross-course behavior analysis. In
-Figure~\ref{fig:cross-course} we show two latent state representations
-learned by a 6-state TL-HMM, one from \textretrieval{} and one from
-\sustain{}. These two states were chosen as they are the most similar
-across the two courses. However, if we look at the transitions we can see
-some important differences. First, in the state from \textretrieval{}, we
-can see that the probability of returning to the course wiki after viewing
-a lecture (transition probability $p \approx 0.24$) is considerably lower
-than that probability in the state from \sustain{} (transition probability
-$p \approx 0.57$). We also notice that the self-loop for staying in a
-lecture activity in \textretrieval{} (transition probability $p \approx
-0.70$) is significantly higher probability than it is in the state from
-\sustain{} (transition probability $p \approx 0.34$).  This gives us some
-insight into the lecture viewing behavior of students in these two MOOCs
-which can possibly be a reflection of the course's structure (which, as
-demonstrated in \citet{Davis:2016:EDM}, can differ substantially across
-different MOOCs).  In \textretrieval{}, students are likely to view lecture
-videos in succession directly without first visiting the course wiki,
-whereas in \sustain{} students are much more likely to first return the
-course wiki before watching the next lecture video. This observation would
-be lost if we did not consider the transitions between the actions within
-the latent states.
-
-\subsection{Varying the Number of Latent States}
-Our TL-HMM model has a parameter $K$ that sets the number of latent states
-to be learned. We believe that this can allow our model to flexibly
-discover patterns of different granularity, and we can show this by varying
-$K$ for a particular course and observing how the latent state
-representations evolve.
-
-\afterpage{%
-\begin{landscape}
-\begin{figure*}
-  \centering
-  \begin{tabular}{cccc}
-    \textbf{State 0 (Wiki)} & \textbf{State 1 (Lecture)}
-    & \textbf{State 2 (Thread View)} & \textbf{State 3 (Lecture/Wiki)}\\\hline
-    \includegraphics[width=0.30\textwidth]{figures/text-2state/state0.png}
-    &
-    \includegraphics[width=0.30\textwidth]{figures/text-2state/state1.png}\\
-    \includegraphics[width=0.30\textwidth]{figures/text-3state/state0.png}
-    &
-    \includegraphics[width=0.30\textwidth]{figures/text-3state/state1.png}
-    &
-    \includegraphics[width=0.30\textwidth]{figures/text-3state/state2.png}\\
-    \includegraphics[width=0.30\textwidth]{figures/text-4state/state0.png}
-    &
-    \includegraphics[width=0.30\textwidth]{figures/text-4state/state1.png}
-    &
-    \includegraphics[width=0.30\textwidth]{figures/text-4state/state2.png}
-    &
-    \includegraphics[width=0.30\textwidth]{figures/text-4state/state3.png}
-  \end{tabular}
-  \caption{The evolution of states for increasing $K$ for the
-  \protect\textretrieval{} MOOC.\ Each row corresponds to the next value of
-  $K$, starting from $K=2$. State ``names'' indicate the highest
-  probability action(s) within the state.} % I don't know why \protect is needed
-  \label{fig:text-state-evolution}
-\end{figure*}
-\end{landscape}
-}
-
-\afterpage{%
-\begin{landscape}
-\begin{figure*}
-  \centering
-  \begin{tabular}{cccc}
-    \textbf{State 0 (Wiki)}
-    & \textbf{State 1 (Lecture/Wiki)}
-    & \textbf{State 2 (Thread List)}
-    & \textbf{State 3 (Thread View)}\\\hline
-    \includegraphics[width=0.30\textwidth]{figures/sustain-2state/state0.png}
-    &
-    \includegraphics[width=0.30\textwidth]{figures/sustain-2state/state1.png}\\
-    \includegraphics[width=0.30\textwidth]{figures/sustain-3state/state0.png}
-    &
-    \includegraphics[width=0.30\textwidth]{figures/sustain-3state/state2.png}
-    &
-    \includegraphics[width=0.30\textwidth]{figures/sustain-3state/state1.png}\\
-    \includegraphics[width=0.30\textwidth]{figures/sustain-4state/state0.png}
-    &
-    \includegraphics[width=0.30\textwidth]{figures/sustain-4state/state3.png}
-    &
-    \includegraphics[width=0.30\textwidth]{figures/sustain-4state/state1.png}
-    &
-    \includegraphics[width=0.30\textwidth]{figures/sustain-4state/state2.png}
-  \end{tabular}
-  \caption{The evolution of states for increasing $K$ for the
-  \protect\sustain{} MOOC.\ Each row corresponds to the next value of $K$,
-  starting from $K=2$. State ``names'' indicate the highest
-  probability action(s) within a state.} % I don't know why \protect is needed
-  \label{fig:sustain-state-evolution}
-\end{figure*}
-\end{landscape}
-}
-
-In Figure~\ref{fig:text-state-evolution} we see\footnote{The graphs are
-quite small, but the states are in the same positions as they were in
-previous figures.} the evolution of the latent state
-representations found for the \textretrieval{} MOOC.  With $K=2$ we uncover
-a video watching pattern (state 1) and a course material browsing pattern
-(state 0). However, we can see that when $K=3$ we are able to uncover a new
-pattern involving forum behavior in state 3 (notice the node sizes on the
-bottom right). As we increase $K$ to four, we can see that state 1 splits
-into state 1 and state 3. These states appear quite similar at a glance,
-but there are still some key differences. First, we can see that state 1
-now has a non-negligible forum component, whereas state 3 has hardly any
-weight on the forum.
-
-We observe similar behavior in Figure~\ref{fig:sustain-state-evolution}
-as we increase $K$ when fitting our TL-HMM on the \sustain{} MOOC.\ Again,
-in the transition between $K=2$ and $K=3$ we discover forum behavior
-patterns as a latent state. However, in the transition between $K=3$ to
-$K=4$ we instead see a refining of that discovered forum state, where
-state 3 captures asymmetric transition probabilities between ``forum thread
-view'' and ``forum thread list''. These states could be seen as redundant,
-in which case a  setting of $K=3$ may be more appropriate for the
-\sustain{} dataset.
-
-\subsection{Transitions Between Latent States}
+To visualize the Markov models that represent our latent states, we plot
+them as a directed graph where we set the size of a node to be proportional
+to its probability of being visited during a random walk.  We let the
+thickness of a directed edge $(u, v)$ reflect the probability of taking
+that edge given that a random walk is currently at note $u$ (as indicated
+by the transition matrix). Figure~\ref{fig:states} shows the four latent
+states we learned by fitting a 4-state TL-HMM to the \textretrieval{}
+sequence dataset.
 
@@
 latent states. In Figure~\ref{fig:trans-avg} we show the latent state
-transition diagram for a 3-state TL-HMM fit on \textretrieval{}. We can
-immediately observe two things: \begin{enumerate*}[label=(\arabic*)]
-  \item each latent state has a very high ``staying'' probability, and
-  \item the prevalence of each latent state matches our intuition.
-\end{enumerate*}
-In particular, we can see that the forum browsing state (state 2) has
-relatively lower probability than the other states as we might expect. It
-also makes sense that state 0 has rather high probability as this state
-likely captures all sequences where a student logged in to the platform and
-then did nothing else (likely checking for updates). If we look at state 1
-and state 3, their relative probabilities match our intuition as
-well. State 1 seems to capture a more engaged browsing session, where there
-is non-negligible probability associated with different activities such as
-quiz taking and forum browsing and, importantly, these activities have high
-probability symmetric edges (so students are taking quizzes one after the
-other, or viewing forum threads in succession). By contrast, state 3 seems
-to capture a more passive student, with negligible probability mass
-associated with forum activity (with low symmetry in the edges). The link
-between ``quiz submit'' and ``quiz start'' (indicating quiz repetition) is
-also significantly lower than state 1.
-
-\begin{figure*}
-  \centering
-  \begin{subfigure}[t]{0.33\textwidth}
-    \includegraphics[width=\textwidth,trim={0 3cm 0 2cm}]{figures/trans-comp/trans-avg.png}
-    \caption{\label{fig:trans-avg}}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.33\textwidth}
-    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{figures/trans-comp/trans-perfect.png}
-    \caption{\label{fig:trans-perfect}}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.33\textwidth}
-    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{figures/trans-comp/trans-low.png}
-    \caption{\label{fig:trans-low}}
-  \end{subfigure}
-  \caption{The latent state transition diagrams for a 4-state TL-HMM fit to
-  \protect\textretrieval{} for all students (a) compared to only ``perfect''
-  students (b) and only ``low'' students (c).}
-  \label{fig:trans-comp}
-\end{figure*}
+transition diagram for the 4-state TL-HMM we learned.
+We can see that the ``forum browsing'' state 2 (figure~\ref{fig:state2})
+has relatively lower probability than the other states as we might expect.
+It also makes sense that state 0 (figure~\ref{fig:state0}) has rather high
+probability as this state likely captures all sequences where a student
+logged in to the platform and then did nothing else (likely checking for
+updates). If we look at state 1 (figure~\ref{fig:state1}) and state 3
+(figure~\ref{fig:state3}), their relative probabilities match our intuition
+as well. State 1 seems to capture a more engaged browsing session, where
+there is non-negligible probability associated with different activities
+such as quiz taking and forum browsing and, importantly, these activities
+have high probability symmetric edges (so students are taking quizzes one
+after the other, or viewing forum threads in succession). By contrast,
+state 3 seems to capture a more passive student, with negligible
+probability mass associated with forum activity (with low symmetry in the
+edges). The link between ``quiz submit'' and ``quiz start'' (indicating
+quiz repetition) is also significantly lower than state 1.
 
@@
 model we learned on the full training data and retrofit it to training data
-consisting only of sequences produced by students in \textretrieval{} that
-had perfect marks. To prevent the latent state meanings from drifting, we
-forced the model parameters associated with their Markov model
-representations to be fixed, in effect only learning initial and transition
-probabilities for the top layer of our TL-HMM.\ We show the updated latent
-state transition diagram in Figure~\ref{fig:trans-perfect}. We can clearly
-see that the probability of state 2 has increased dramatically, consistent
-with previous observations of the positive correlation between forum activity and
-grades~\cite{Huang:2014:LAS}, while the probability of states 0 and 3 has
+consisting only of sequences produced by students that had perfect marks.
+To prevent the latent state meanings from drifting, we forced the model
+parameters associated with their Markov model representations to be fixed,
+in effect only learning initial and transition probabilities for the top
+layer of our TL-HMM.\ We show the updated latent state transition diagram
+in Figure~\ref{fig:trans-perfect}. We can see that the probability of state
+2 has increased dramatically, while the probability of states 0 and 3 has
 decreased. State 1 had its probability increase, but only very slightly.
@@
 state 1 did not increase in size for the ``perfect'' group like we were
-anticipating. \citet{Kizilcec:2017:CandE} observe similar phenomena with
-the courses they studied where they find that certificate earning is
-negatively correlated with help seeking behavior. Our model lends itself
-well to discovering this potentially counter-intuitive insight directly
-from data.
-
-To quantify this finding, we perform the following experiment. First, we
-select all students from \textretrieval{} who completed all of the quizzes
-$\mathbf{L}_q \subset \mathbf{L}$. This gives us 1,985 students along with
-their average quiz score. We then create a ranked list of the students in
-$\mathbf{L}_q$ by sorting them by their ``preference'' for a specific
-latent state
-\begin{equation}
-  p_\ell(i) = \frac{\sum_{t=1}^T \gamma_t^{(\mathbf{o}_\ell)}(i)}
-  {\sum_{j=1}^{K} \sum_{t=1}^T \gamma_t^{(\mathbf{o}_\ell)}(j)}
-\end{equation}
-where $\mathbf{o}_\ell$ is the list of action sequences for student $\ell$
-and $\gamma$ is defined as before and computed using the Baum-Welch
-algorithm. We can now compare the average rank in this list for both
-the ``perfect'' student group and the ``low'' student group: a useful state
-for distinguishing the two groups should result in a ranked list with
-statistically significant differences in average rank between the two
-groups. Our results are summarized in Table~\ref{table:mean-rank}. Indeed,
-we discover that states 1 and 2 are useful for teasing apart these two
-groups: state 1 ranks students in the ``low'' group significantly higher
-than those in the ``perfect'' group, and state 2 does the opposite and
-prefers students in the ``perfect'' group to those in the ``low'' group.
-
-\begin{table}
-  \centering
-  \begin{tabular}{r|rrrr}
-    \textbf{Group} & \textbf{State 0} & \textbf{State 1}^\dagger &
-    \textbf{State 2}^\dagger & \textbf{State 3}\\\hline
-    \textbf{Perfect} & 975.3  & 1001.5 & 999.0  & 1056.5\\
-    \textbf{Low}     & 1024.9 & 816.4  & 1230.5 & 1161.2\\
-  \end{tabular}
-  \caption{Average rank for ``perfect'' and ``low'' student groups in the
-  ranked lists associated with the four latent states found by a TL-HMM.
-  $\dagger$ indicates statistically significant different mean ranks at $p
-  < 0.01$ according to an unpaired $t$-test.}
-  \label{table:mean-rank}
-\end{table}
+anticipating. Our model lends itself well to discovering this potentially
+counter-intuitive insight.

### 2017-02-22 short-paper/clickstream-hmm-short.tex
Fix typo in short paper's keywords.
Update short paper to use SIGCHI Proceedings format.

@@
-\documentclass{sigchi-ext}
+%% SIGCHI Proceedings Format (modern) Sample Document
+%% This sample file demonstrates the use of the `sigchi-modern' class file.
+%%
+%% The content of this document draws heavily on the `HCI Archive Format'
+%% template supplied by SIGCHI, but has been modified to demonstrate and
+%% document the LaTeX class.
 
-% Please be sure that you have the dependencies (i.e., additional
-% LaTeX packages) to compile this example.
-\usepackage[T1]{fontenc}
-\usepackage{textcomp}
-\usepackage[scaled=.92]{helvet} % for proper fonts
-\usepackage{graphicx} % for EPS use the graphics package instead
-\usepackage{balance}  % for useful for balancing the last columns
-\usepackage{booktabs} % for pretty table rules
-\usepackage{ccicons}  % for Creative Commons citation icons
-\usepackage{ragged2e} % for tighter hyphenation
+% The parameter given to the `\documentclass' command can be one of:
+%   `preprint', `submission', or `final'.
+% These control the presence of certain features for each stage of the authoring
+% process. For example, `submission' does not output the author block, and
+% `final' disables page numbers.
+%
+% The \preprintonly{}, \submissiononly{}, and \finalonly{} commands allow you
+% to wrap text that will only be output under the corresponding option.
 
-% Some optional stuff you might like/need.
-% \usepackage{marginnote}
-% \usepackage[shortlabels]{enumitem}
-% \usepackage{paralist}
-\usepackage[utf8]{inputenc} % for a UTF8 editor only
-\usepackage{amsmath}
-\usepackage{float}
+\PassOptionsToPackage{inline}{enumitem}
+\documentclass[final]{sigchi-modern}
+
+\def\meta/{\preprintonly{\textsc{MeTA}}\finalonly{\textsc{MeTA}}%
+\submissiononly{(redacted)}}
+
+\newcommand{\citemeta}{\preprintonly{\cite{Massung:2016:ACL}}%
+  \finalonly{\cite{Massung:2016:ACL}}%
+  \submissiononly{[redacted citation]}}
+
+% These packages aren't required, but do provide some helpful features that are
+% used in this sample.
+\usepackage{graphicx}  % Graphics for figures
+\usepackage{booktabs}  % Nice formatting for tables
+\usepackage{balance}   % Attempts to balance the columns on the last page
+
+% A nice way to help avoid overful lines handing out into the margins is the
+% `microtype' package, which lets certain characters extrude a *little* bit.
+\usepackage{microtype}
+% If the situation becomes dire, use:
+%\sloppy
+
+% You may use any bibliography package as long as the output matches that of
+% the prescribed format. `natbib' is a fairly nice default, and the provided
+% BibTeX style file has been tested with it.
 \usepackage[numbers]{natbib}
+\bibliographystyle{acm-sigchi-modern}
 
@@
 
-\renewcommand\cite{\citep}
-\def\meta/{\textsc{MeTA}}
-\newcommand{\textretrieval}{textretrieval-001}
-\newcommand{\sustain}{sustain-001}
-\newcommand{\UIUC}{UIUC}
+% Define these tokens as appropriate for the conference to generate the
+% necessary copyright notice for `final` mode.
+\confname{L@S 2017}
+\confdate{April 20-21}
+\confyear{2017}
+\conflocation{Cambridge, MA, USA}
+\procissn{978-1-4503-4450-0/17/04}
+\doi{10.1145/3051457.3053986}
+% \copylicense be set to one of: \acmcopyright, \authorlicense, or \openlicense
+%\copylicense{\authorlicense}
 
-%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP --
-% \copyrightinfo{Permission to make digital or hard copies of all or
-% part of this work for personal or classroom use is granted without
-% fee provided that copies are not made or distributed for profit or
-% commercial advantage and that copies bear this notice and the full
-% citation on the first page. Copyrights for components of this work
-% owned by others than ACM must be honored. Abstracting with credit is
-% permitted. To copy otherwise, or republish, to post on servers or to
-% redistribute to lists, requires prior specific permission and/or a
-% fee. Request permissions from permissions@acm.org.\\
-% {\emph{CHI'14}}, April 26--May 1, 2014, Toronto, Canada. \\
-% Copyright \copyright~2014 ACM ISBN/14/04...\$15.00. \\
-% DOI string from ACM form confirmation}
-%% EXAMPLE END
-
-\copyrightinfo{Permission to make digital or hard copies of all or part of
-  this work for personal or classroom use is granted without fee provided
-  that copies are not made or distributed for profit or commercial
-  advantage and that copies bear this notice and the full citation on the
-  first page. Copyrights for components of this work owned by others than
-  the author(s) must be honored. Abstracting with credit is permitted. To
-  copy otherwise, or republish, to post on servers or to redistribute to
-  lists, requires prior specific permission and/or a fee. Request
-  permissions from Permissions@acm.org.\\
-  \emph{L@S 2017}, April 20 - 21, 2017, Cambridge, MA, USA.\\
-  Copyright is held by the owner/author(s). Publication rights licensed to
-  ACM.\\
-  ACM 978-1-4503-4450-0/17/04 \$15.00\\
-  DOI: \url{http://dx.doi.org/10.1145/3051457.3053986}}
-
-% Paper metadata (use plain text, for PDF inclusion and later
-% re-using, if desired).  Use \emtpyauthor when submitting for review
-% so you remain anonymous.
-\def\plaintitle{Modeling MOOC Student Behavior with Two-Layer Hidden Markov
-Models}
-\def\plainauthor{Chase Geigle, ChengXiang Zhai}
-\def\emptyauthor{}
-\def\plainkeywords{MOOC log analysis; studetn behavior modeling; Markov models; hidden Markov models}
-\def\plaingeneralterms{\plainkeywords}
-
-\title{\plaintitle}
-
-\numberofauthors{2}
-% Notice how author names are alternately typesetted to appear ordered
-% in 2-column format; i.e., the first 4 autors on the first column and
-% the other 4 auhors on the second column. Actually, it's up to you to
-% strictly adhere to this author notation.
-\author{%
-  \alignauthor{%
-    \textbf{Chase Geigle}\\
-    \affaddr{University of Illinois at Urbana-Champaign} \\
-    \affaddr{201 N Goodwin Ave}\\
-    \affaddr{Urbana, IL 61801} \\
-    \email{geigle1@illinois.edu} } \vfil \alignauthor{%
-    \textbf{ChengXiang Zhai}\\
-    \affaddr{University of Illinois at Urbana-Champaign} \\
-    \affaddr{201 N Goodwin Ave}\\
-    \affaddr{Urbana, IL 61801} \\
-    \email{czhai@illinois.edu} } }
-
-% Make sure hyperref comes last of your loaded packages, to give it a
-% fighting chance of not being over-written, since its job is to
-% redefine many LaTeX commands.
-\definecolor{linkColor}{RGB}{6,125,233}
-\hypersetup{%
-  pdftitle={\plaintitle},
-%  pdfauthor={\plainauthor},
-  pdfauthor={\emptyauthor},
-  pdfkeywords={\plainkeywords},
-  bookmarksnumbered,
-  pdfstartview={FitH},
-  colorlinks,
-  citecolor=black,
-  filecolor=black,
-  linkcolor=black,
-  urlcolor=linkColor,
-  breaklinks=true,
-}
-
-% \reversemarginpar%
-
-\newcommand{\citemeta}{\cite{Massung:2016:ACL}}
-
+% End of preamble. Here comes the document.
 \begin{document}
 
+\title{Modeling MOOC Student Behavior With Two-Layer Hidden Markov Models}
+
+% Author information can be set automatically if you provide a series of
+% author names and affiliations:
+%   \author[1]{Author One}
+%   \author[2]{Author Two}
+%   \author[1,2]{Author Three}
+%   \affiliation[1]{Organisation One\\
+%                   Some Country}
+%   \affiliation[2]{Organisation Two\\
+%                   Some Country}
+%   \authorextra{\mailto{one@example.com}, \mailto{two@example.org},
+%                \mailto{three@example.net}}
+% Or, manually:
+%   \author{
+%     \authorname{Author One}\\
+%     \authoraffil{
+%       Organisation One\\
+%       Some Country\\
+%       \mailto{one@example.com}}}
+%   \author{
+%     \authorname{Author Two}\\
+%     \authoraffil{
+%       Organisation Two\\
+%       Some Country\\
+%       \mailto{two@example.org}}}
+% Refer to the body text of the document for more details.
+\author[1]{Chase Geigle}
+\author[1]{ChengXiang Zhai}
+
+% Set \authorwidth with the longest author name if you don't like the default
+% spacing:
+\settowidth{\authorwidth}{ChengXiang Zhai1}
+\settowidth{\affilwidth}{University of Illinois at Urbana-ChampaignAAA}
+
+\affiliation[1]{Department of Computer Science\\
+    University of Illinois at Urbana-Champaign\\
+    Urbana, Illinois, USA}
+
+\authorpostscript{\email{\{%
+  \href{mailto:geigle1@illinois.edu}{geigle1},
+  \href{mailto:czhai@illinois.edu}{czhai}\}@illinois.edu}}
+
+% If you format the author blocks manually, use \authorlist with a
+% comma-delimited list of the author names.
+% This is written into the PDF metadata for preprint and final modes.
+%\authorlist{Author One, Author Two, Author Three, Author Four}
+
+% Produces the title, author, and optional banner block.
 \maketitle
 
-% Uncomment to disable hyphenation (not recommended)
-% https://twitter.com/anjirokhan/status/546046683331973120
-\RaggedRight{}
-
-% Do not change the page size or page settings.
 \begin{abstract}
@@
   huge amounts of MOOC activity present in log data. In this paper, we
-  propose a method for automatically discovering student behavior patterns
-  by leveraging the click log data that can be obtained from the MOOC
-  platform itself in a completely unsupervised manner.
+  propose a new method for automatically discovering student behavior
+  patterns by leveraging the click log data that can be obtained from the
+  MOOC platform itself in a completely unsupervised manner.
 \end{abstract}
 
-\keywords{\plainkeywords}
+% A semicolon-separated list of keywords to describe the paper.
+\keywords{MOOC log analysis; student behavior modeling; Markov models;
+hidden Markov models}
+
+\newcommand{\acknowledgments}{This material is based upon work supported by
+the National Science Foundation under Grant Number DGE-1144245 (GRFP) and
+Grant Number 1629161.}
+
+% ACM classification block.
+% \category{}{}{}[] has three mandatory arguments and one optional argument; it
+% can be used multiple times.
+%\classification{%
+%    \category{}{}{}
+%}
 
@@
 
-\input{sec-intro}
+\input{../sec-intro}
 \input{sec-model}
@@
 \section{Acknowledgments}
+\preprintonly{\acknowledgments{}}
+\finalonly{\acknowledgments{}}
+\submissiononly{(Anonymized for blind reviewing.)}
 
-This material is based upon work supported by the NSF GRFP under Grant
-Number DGE-1144245.
-
-\balance{}
-
-\bibliographystyle{acm-sigchi-modern}
+\balance
 \bibliography{bib}

### 2017-02-22 short-paper/sec-conclusion.tex
Update short paper to use SIGCHI Proceedings format.

@@
-\section{Conclusions and Future Directions}
+\section{Conclusion}
+% summarize the main contributions of the paper; summarize the experimental
+% results; argue again for our novelty; discuss future directions
 We proposed a two-layer hidden Markov model for MOOC student behavior
-modeling on large collections of actions extracted from MOOC log data. This
-model uniquely treats behavior patterns as a sequence of \emph{latent states},
-rather than assigning these states in a rule-based manner, captures the
-variable behaviors of students over time, and allows analysis at different
-levels of granularity.
+modeling in an unsupervised learning framework on large collections of
+actions extracted from MOOC log data. This model is different from existing
+methods in that it treats behavior patterns as a sequence of \emph{latent
+states}, rather than assigning these states in a rule-based manner. It
+captures the variable behaviors of students over time, and allows
+analysis at different levels of granularity.
+
+We showed that such a model does in fact capture meaningful behavior
+patterns and produces descriptions of these behavior patterns that are easy
+to interpret. We argued that it is important to capture student behavior
+patterns with more sophisticated models than simple discrete distributions
+over actions in order to capture information present in bigrams of actions
+(or above). Finally, we investigated whether we can detect differences in
+student behavior patterns as they correlate with course performance.
+Specifically, we demonstrated that high-performing students produce
+substantially different HMM transition diagrams that tend to show longer
+concentration span in quiz-taking and more active forum participation as
+compared with the average students.
+
+\subsection{Future Work}
 
@@
 other course to enable deep understanding of student behaviors as well as
-the correlations of such behaviors as other variables such as grades.
-Our model can be used to produce an ordered list of latent states for each
-student over time from which we could correlate course-wide drifts with
-events in a course, which could for example allow the automatic discovery
-of difficult segments of a course.
+the coorrelations of such behaviors as other variables such as grades. We
+plan to develop a MOOC log analysis system based on the proposed model to
+both facilitate education research and help instructors improve course
+design.
+
+Specifically, our model can be used to produce an ordered list of latent
+states for each student over time. Given these labeled sequences, we could
+correlate course-wide drifts in these latent states with events in a
+course. For example, we might be able to automatically discover difficult
+or confusing parts of a course by noticing spikes in the distribution of
+students over latent states over time.
+
+%Our model does not explicitly model drop-out like
+%\citet{Kizilcec:2013:LAK}, but doing so is an obvious extension. Our model
+%should be able to provide predictions of when a student is ``at risk'' for
+%dropping out under such an extension.
+%
+%Currently, the model learns a transition matrix over the latent states that
+%is \emph{shared} across all students. It would be interesting to instead
+%learn a different latent state transition matrix for each individual
+%student, but keep the second-level Markov models shared. This would provide
+%the model with more flexibility which may be desirable itself, but would
+%also naturally result in a description of a student (via his or her HMM
+%transitions) that could be incorporated into existing supervised learning
+%techniques that try to predict student outcomes for understanding
+%which of the latent behavior patterns discovered by TL-HMM are most predictive
+%of the performance of student learning.

### 2017-02-22 short-paper/sec-model.tex
Update short paper to use SIGCHI Proceedings format.

@@
 
-\begin{margintable}
-  \vspace{-4.7in}
-  \begin{center}
-    \begin{tabular}{rr}
-      \multicolumn{2}{c}{\textbf{\textretrieval{}}}\\\hline
-      Students & 18,941\\
-      Sequences & 85,240\\
-      Avg. Len. & 7.31
-    \end{tabular}
-    \caption{Statistics about the sequences extracted from the MOOC.}
-    \label{table:datasets}
-  \end{center}
-\end{margintable}
+Our general idea is to use a probabilistic generative model to model 
+the student activities as recorded in a MOOC log, which means we will 
+assume that all the observed student activities are samples drawn (i.e., ``generated'') from 
+a parameterized probabilistic model. We can then estimate the 
+parameter values of the probabilistic model by fitting the model to a specific
+MOOC log data set. The estimated parameter values could then be treated as 
+the latent ``knowledge'' discovered from the data. Because such a generative model 
+attempts to fit {\em all} the data, it enables us to discover interesting patterns
+that can explain the {\em overall} behavior of a student or the {\em common} behavior patterns shared by many students.
 
-\marginpar{%
-  \vspace{-1.6in}
-  Each sequence is comprised of the following actions:
-  \begin{enumerate}[itemsep=0,align=left,leftmargin=*]
-    \item quiz start,
-    \item quiz submit,
-    \item wiki (course material),
-    \item forum list (view the list of all forums),
-    \item forum thread list (view the list of all threads in a specific
-      forum),
-    \item forum thread view (view the list of posts within a specific
-      thread),
-    \item forum search (a search query issued against the forum),
-    \item forum post thread (a new thread was posted),
-    \item forum post reply (a new post was created within an existing
-      thread), and
-    \item view lecture (defined as either streaming or downloading a lecture
-      video).
-  \end{enumerate}
-}
+An HMM  is a specific probabilistic generative model with a ``built-in'' state transition system
+that would control the data to be generated by the model, thus it is especially 
+suitable for modeling sequence data~\cite{Rabiner:1990:RSR}. At any moment, the HMM would be in 
+one of $k$ states $U=\{u_1, ...,u_k\}$, and at the next moment, the HMM would move to 
+another state stochastically according to a transition matrix that specifies the probability of
+going to state $u_i$ when the HMM is currently in state $u_j$, i.e., $p(u_i|u_j)$. 
+When the HMM is in state $u$, the HMM can generate an observable data point $x$ 
+according to an output probabilistic model $p(x|u)$. Thus if we ``run'' an HMM for 
+$N$ time points denoted by $t=1, ..., N$, the HMM could ``generate'' a sequence of 
+observations $x_1 ... x_N$, where each $x_i$ is an output symbol by 
+going through a sequence of {\em hidden states} $w_1 ... w_N$ where $w_i \in U$ is a state. 
+The association of such a latent sequence of state transitions with the observed symbols makes
+it possible to use HMM to ``decode'' the latent behavior of students behind the surface behavior we directly observe in the log data, allowing for understanding student behavior more deeply than a model with no latent state variables.
 
-Our general idea is to use a probabilistic generative model to model the
-student activities as recorded in a MOOC log, which means we will assume
-that all the observed student activities are samples drawn (i.e.,
-``generated'') from a parameterized probabilistic model. We can then
-estimate the parameter values of the probabilistic model by fitting the
-model to a specific MOOC log data set. The estimated parameter values could
-then be treated as the latent ``knowledge'' discovered from the data.
-Because such a generative model attempts to fit {\em all} the data, it
-enables us to discover interesting patterns that can explain the {\em
-overall} behavior of a student or the {\em common} behavior patterns shared
-by many students.
+In many ways, the generation process behind an HMM is meant to simulate the
+actual behavior of a student. We assume that students transition through different ``task states'' (or ``behavior states'')  in the process of study. 
+One such task state may be to learn about a topic by mostly watching
+lecture videos, another task state may be to work on quizzes, and yet
+another may be to participate in forum discussions. While in each of these
+different states, the student would tend to exhibit different surface
+``micro'' behaviors. For example, in the lecture study state, the student
+would tend to have many video-watching related behaviors and occasionally
+forum activities, while in the quiz-taking state (in order to pass each
+module), the student would tend to show many quiz-related ``micro''
+activities as well as asking questions or checking discussions on the
+forum. Note that due to the complexity of the student behavior, it is very
+difficult to accurately {\em prescribe} the specific surface ``micro''
+behavior patterns for each state in advance, especially without  prior knowledge about the students.
+For example, forum activities are likely interleaved with other activities   in every task state and the interleaving pattern can be somewhat irregular with potentially many variations.
+The major motivations for using an HMM are that 
+\begin{enumerate*}[label=(\arabic*)]
+    \item it uses a probabilistic model (i.e., the output probability distribution $p(x|u)$ conditioned on each state) to directly capture the inevitable uncertainty in the association of surface ``micro'' activities with their corresponding latent task/behavior state, which is often our main target to discover and characterize, and
+    \item it does not make any assumption about which latent task/behavior
+      state must be associated with which observed activities or how a
+      student would move from one state to another, but instead allows our
+      data to ``tell'' us what kind of associations are most likely,
+      what kind of transitions are most probable, and which states tend to be more long-lasting for any particular set of students. 
+\end{enumerate*}
 
-An HMM is a specific probabilistic generative model with a ``built-in''
-state transition system that would control the data to be generated by the
-model, thus it is especially suitable for modeling sequence
-data. We may say that students transition through different ``task states''
-(or ``behavior states'')  in the process of study.  One such task state may
-be to learn about a topic by mostly watching lecture videos, another task
-state may be to work on quizzes, and yet another may be to participate in
-forum discussions. While in each of these different states, the student
-would tend to exhibit different surface ``micro'' behaviors.  For example,
-in the lecture study state, the student would tend to have many
-video-watching related behaviors and occasionally forum activities, while
-in the quiz-taking state, the student would tend to show many quiz-related
-``micro'' activities as well as asking questions or checking discussions on
-the forum.
+However, if we use an ordinary HMM to analyze our data, we would treat each
+observed ``micro'' activity (e.g., video watching, or forum post reading)
+as an output symbol, and thus the output distribution $p(x|u)$ for each
+discovered latent state would be a simple distribution over all kinds of
+observable micro activities recorded in our log data (e.g., 50\% lecture
+watching, 8\% quiz taking, 7\% quiz submission, 2\% course wiki reading,
+...). While such a distribution is meaningful and can already help us
+interpret the corresponding latent state, it only gives us a rather
+superficial characterization of student behavior.
 
-The most natural application of an HMM to analyze our data would treat each
-latent state as having a discrete distribution over the surface ``micro''
-behaviors, but this only gives a rather superficial characterization of the
-student behavior. Instead, we want to directly characterize the transitions
-between the ``micro'' actions, so we treat an {\em entire
+Ideally, we want $p(x|u)$ to characterize the directly observable ``micro''
+behavior in more detail to further capture the relations and dependencies
+of these micro activities. To this end, we would treat an {\em entire
 sequence} of micro activities (e.g., one session of activities) as an
@@
 these activity states in very much the same way as the state transitions in
-HMM. Adding this second layer would allow us to characterize a latent task
+HMM.
+
+Adding this second layer would allow us to characterize a latent task
 state in much more detail, as it would reveal not only what activities are
 most common to a task state, but also the transition patterns between these
-activities.
+``micro'' activities (e.g., it can reveal frequent back-and-forth
+transitions between quiz-taking and quiz-submission, which would suggest a
+concentrated period of taking quizzes). Combining this ``surface'' Markov
+model over the ``micro'' actions with the ``deep'' hidden Markov model over
+the latent task states gives us a fairly general and powerful two-layer HMM
+(TL-HMM) that can simultaneously learn ``deeply'' the latent task/behavior
+states and their transitions as well as the corresponding ``micro''
+activity transition patterns associated with each latent state to
+facilitate interpretation and analysis of the discovered latent state
+patterns. Our implementation of the learning algorithm for TL-HMMs is
+included as part of the \meta/ toolkit~\citemeta{}.

### 2017-02-22 short-paper/sec-results.tex
Update short paper to use SIGCHI Proceedings format.

@@
+\newcommand{\textretrieval}{textretrieval-001}
+
+\newcommand{\sustain}{sustain-001}
+
+\newcommand{\UIUC}{UIUC}
+
 \section{Experiment Results}
-
 As an analysis tool, the TL-HMM model provides with us the following two
@@
 qualitatively analyze both types of patterns discovered from empirical MOOC
-log data. Specifically, we look at the MOOC logs associated with the
-\textretrieval{} Coursera MOOC offered by \UIUC{}.
-Table~\ref{table:datasets} provides statistics about the sequences we
-extracted from this MOOC and the action space we used for those sequences.
+log data.
 
-\begin{marginfigure}
-  %\vspace{-2in}
-  \centering
-  \begin{subfigure}[t]{\textwidth}
-    \includegraphics[width=0.95\textwidth,trim={0 3cm 0 2cm}]{../figures/trans-comp/trans-avg.png}
-    \caption{\vspace{20pt}\label{fig:trans-avg}}
-  \end{subfigure}
+Specifically, we looked at the MOOC logs associated with the \textretrieval{}
+Coursera MOOC offered by \UIUC{} and extracted a dataset consisting of
+18,941 students who produced 85,240 sequences with an average length of
+7.31. We used the following ten actions as our action set $\mathbf{A}$:
+\begin{enumerate*}[label=(\arabic*)]
+  \item quiz start,
+  \item quiz submit,
+  \item wiki (course material),
+  \item forum list (view the list of all forums),
+  \item forum thread list (view the list of all threads in a specific
+    forum),
+  \item forum thread view (view the list of posts within a specific
+    thread),
+  \item forum search (a search query issued against the forum),
+  \item forum post thread (a new thread was posted),
+  \item forum post reply (a new post was created within an existing
+    thread), and
+  \item view lecture (defined as either streaming or downloading a lecture
+    video).
+\end{enumerate*}
 
-  \begin{subfigure}[t]{\textwidth}
-    \includegraphics[width=0.95\textwidth,trim={0, 3cm 0 2cm}]{../figures/trans-comp/trans-perfect.png}
-    \caption{\vspace{20pt}\label{fig:trans-perfect}}
-  \end{subfigure}
-
-  \begin{subfigure}[t]{\textwidth}
-    \includegraphics[width=0.95\textwidth,trim={0, 3cm 0 2cm}]{../figures/trans-comp/trans-low.png}
-    \caption{\label{fig:trans-low}}
-  \end{subfigure}
-
-  \caption{The latent state transition diagrams for a 4-state TL-HMM fit to
-  \protect\textretrieval{} for all students (a) compared to only ``perfect''
-  students (b) and only ``low'' students (c).}
-  \label{fig:trans-comp}
-\end{marginfigure}
+\subsection{Latent State Representations}
+To visualize these Markov models that represent our latent states, we plot
+them as a directed graph where we set the size of a node to be proportional
+to its probability of being visited during a random walk.  We let the
+thickness of a directed edge $(u, v)$ reflect the probability of taking
+that edge given that a random walk is currently at note $u$ (as indicated
+by the transition matrix)\footnote{We do not plot the transition
+probabilities directly within the figure to ease readability; we instead
+will mention relevant transition probabilities in the text as we discuss
+the plots. The plots were created using python-igraph:
+\url{http://igraph.org/python/}.}.
 
@@
 
-To visualize the Markov models that represent our latent states, we plot
-them as a directed graph where we set the size of a node to be proportional
-to its probability of being visited during a random walk.  We let the
-thickness of a directed edge $(u, v)$ reflect the probability of taking
-that edge given that a random walk is currently at note $u$ (as indicated
-by the transition matrix). Figure~\ref{fig:states} shows the four latent
-states we learned by fitting a 4-state TL-HMM to the \textretrieval{}
-sequence dataset.
+Figure~\ref{fig:states} shows the latent state representations learned by
+fitting a 4-state TL-HMM to the \textretrieval{} sequence dataset. Our
+interpretations of the states is as follows:
+\begin{enumerate*}[label=(\alph*)]
+  \item \textbf{state 0} likely captures all sequences where a student
+    logged in to the platform and did nothing else (likely just checking
+    for updates);
+  \item \textbf{state 1} seems to capture a more engaged browsing session,
+    where there is non-negligible probability associated with different
+    activities such as quiz taking and forum browsing and, importantly,
+    these activities have high probability symmetric edges (so students are
+    taking quizzes one after the other, or viewing forum threads in
+    succession);
+  \item \textbf{state 2} captures a ``forum browsing'' state, with most
+    weight being placed on consecutive thread views; and
+  \item \textbf{state 3} seems to capture a more passive student, with
+    negligible probability mass associated with forum activity (with low
+    symmetry in the edges). The link between ``quiz submit'' and ``quiz
+    start'' (indicating quiz repetition) is also significantly lower than
+    state 1.
+\end{enumerate*}
 
+\subsection{Transitions Between Latent States}
 A unique property of our model is its ability to capture transitions
@@
 latent states. In Figure~\ref{fig:trans-avg} we show the latent state
-transition diagram for the 4-state TL-HMM we learned.
-We can see that the ``forum browsing'' state 2 (figure~\ref{fig:state2})
-has relatively lower probability than the other states as we might expect.
-It also makes sense that state 0 (figure~\ref{fig:state0}) has rather high
-probability as this state likely captures all sequences where a student
-logged in to the platform and then did nothing else (likely checking for
-updates). If we look at state 1 (figure~\ref{fig:state1}) and state 3
-(figure~\ref{fig:state3}), their relative probabilities match our intuition
-as well. State 1 seems to capture a more engaged browsing session, where
-there is non-negligible probability associated with different activities
-such as quiz taking and forum browsing and, importantly, these activities
-have high probability symmetric edges (so students are taking quizzes one
-after the other, or viewing forum threads in succession). By contrast,
-state 3 seems to capture a more passive student, with negligible
-probability mass associated with forum activity (with low symmetry in the
-edges). The link between ``quiz submit'' and ``quiz start'' (indicating
-quiz repetition) is also significantly lower than state 1.
+transition diagram for a 4-state TL-HMM fit on \textretrieval{}. We can
+immediately observe two things: \begin{enumerate*}[label=(\arabic*)]
+  \item each latent state has a very high ``staying'' probability, and
+  \item the prevalence of each latent state matches our intuition.
+\end{enumerate*}
+In particular, we can see that the forum browsing state (state 2) has
+relatively lower probability than the other states as we might expect. It
+also makes sense that state 0 (low activity) has rather high probability.
+If we look at state 1 and state 3, their relative probabilities match our
+intuition as well: there should be more students exhibiting more passive
+behaviors (state 3) than very active behaviors (state 1).
+
+\begin{figure*}
+  \vspace{12pt}
+  \centering
+  \begin{subfigure}[t]{0.25\textwidth}
+    \includegraphics[width=\textwidth,trim={0 3cm 0 2cm}]{../figures/trans-comp/trans-avg.png}
+    \caption{\label{fig:trans-avg}}
+  \end{subfigure}%
+  \begin{subfigure}[t]{0.25\textwidth}
+    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{../figures/trans-comp/trans-perfect.png}
+    \caption{\label{fig:trans-perfect}}
+  \end{subfigure}%
+  \begin{subfigure}[t]{0.25\textwidth}
+    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{../figures/trans-comp/trans-low.png}
+    \caption{\label{fig:trans-low}}
+  \end{subfigure}
+  \caption{The latent state transition diagrams for a 4-state TL-HMM fit to
+  \protect\textretrieval{} for all students (a) compared to only ``perfect''
+  students (b) and only ``low'' students (c).}
+  \label{fig:trans-comp}
+\end{figure*}
 
@@
 model we learned on the full training data and retrofit it to training data
-consisting only of sequences produced by students that had perfect marks.
-To prevent the latent state meanings from drifting, we forced the model
-parameters associated with their Markov model representations to be fixed,
-in effect only learning initial and transition probabilities for the top
-layer of our TL-HMM.\ We show the updated latent state transition diagram
-in Figure~\ref{fig:trans-perfect}. We can see that the probability of state
-2 has increased dramatically, while the probability of states 0 and 3 has
-decreased. State 1 had its probability increase, but only very slightly.
+consisting only of sequences produced by students in \textretrieval{} that
+had perfect marks. To prevent the latent state meanings from drifting, we
+forced the model parameters associated with their Markov model
+representations to be fixed, in effect only learning initial and transition
+probabilities for the top layer of our TL-HMM.\ We show the updated latent
+state transition diagram in Figure~\ref{fig:trans-perfect}. We can clearly
+see that the probability of state 2 has increased dramatically, consistent
+with previous observations of the positive correlation between forum
+activity and grades~\cite{Huang:2014:LAS}, while the probability of states
+0 and 3 has decreased. State 1 had its probability increase very slightly.
 
@@
 state 1 did not increase in size for the ``perfect'' group like we were
-anticipating. Our model lends itself well to discovering this potentially
-counter-intuitive insight.
+anticipating. \citet{Kizilcec:2017:CandE} observe similar phenomena with
+the courses they studied where they find that certificate earning is
+negatively correlated with help seeking behavior. Our model enables
+data-driven discovery of potentially counter-intuitive insights like this.
+
+%\begin{table}
+%  \centering
+%  \begin{tabular}{r|rrrrr}
+%    \textbf{Group} & \textbf{State 0} & \textbf{State
+%  1}\textsuperscript\textdagger &
+%    \textbf{State 2}\textsuperscript\textdagger & \textbf{State 3} & \textbf{State 2
+%    $\rightarrow$ 2}\textsuperscript\textdagger\\\hline
+%    \textbf{Perfect} & 975.3  & 1001.5 & 999.0  & 1056.5 & 939.6\\
+%    \textbf{Low}     & 1024.9 & 816.4  & 1230.5 & 1161.2 & 1187.4\\
+%  \end{tabular}
+%  \caption{Average rank for ``perfect'' and ``low'' student groups in the
+%  ranked lists associated with the four latent states found by a TL-HMM.
+%  $\dagger$ indicates statistically significant different mean ranks at $p
+%  < 0.01$ according to an unpaired $t$-test.}
+%  \label{table:mean-rank}
+%\end{table}
+%
+%To quantify this finding, we perform the following experiment. First, we
+%select all students from \textretrieval{} who completed all of the quizzes
+%$\mathbf{L}_q \subset \mathbf{L}$. This gives us 1,985 students along with
+%their average quiz score. We then create a ranked list of the students in
+%$\mathbf{L}_q$ by sorting them by their ``preference'' for a specific
+%latent state
+%\begin{equation}
+%  p_\ell(i) = \frac{\sum_{t=1}^T \gamma_t^{(\mathbf{o}_\ell)}(i)}
+%  {\sum_{j=1}^{K} \sum_{t=1}^T \gamma_t^{(\mathbf{o}_\ell)}(j)}
+%\end{equation}
+%where $\mathbf{o}_\ell$ is the list of action sequences for student $\ell$
+%and $\gamma$ is defined as before and computed using the Baum-Welch
+%algorithm. We can now compare the average rank in this list for both
+%the ``perfect'' student group and the ``low'' student group: a useful state
+%for distinguishing the two groups should result in a ranked list with
+%statistically significant differences in average rank between the two
+%groups. Our results are summarized in Table~\ref{table:mean-rank}. Indeed,
+%we discover that states 1 and 2 are correlated with the ``perfect'' or
+%``low'' groups: state 1 ranks students in the ``low'' group significantly higher
+%than those in the ``perfect'' group, and state 2 does the opposite and
+%prefers students in the ``perfect'' group to those in the ``low'' group.
+%
+%Returning to Figure~\ref{fig:trans-comp}, we an also see a difference in
+%the transitions between the latent states. In particular, look at the
+%transitions in Figure~\ref{fig:trans-perfect} and
+%Figure~\ref{fig:trans-low} that leave state 2. In the ``perfect'' group,
+%nearly all of this probability mass is allocated for the self-loop ($p
+%\approx 0.93$). In the ``low'' group, this self loop is less strong ($p
+%\approx 0.80$; most easily seen by noting that the edges leaving state 2
+%are darker than for the ``perfect'' group). We can perform a similar
+%experiment to above by producing a ranked list of students $\ell \in
+%\mathbf{L}_q$ by their staying probability for state 2 (that is, given that
+%a student $\ell$ is already in state 2, how likely are they to remain there
+%in the next action sequence?). This can be computed as
+%\begin{equation}
+%  p_\ell(2, 2) = \frac{\sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,2)}
+%  {\sum_{i=1}^K \sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,i)}
+%\end{equation}
+%where $\xi$ is defined as before and computed using the Baum-Welch
+%algorithm. The last column of Table~\ref{table:mean-rank} indicates that
+%this transition feature also correlates with the achievement group and
+%results in significant differences in mean rank.

### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md
Add initial draft response to round-1 reviewer C.

> 1.      How relevant is this submission to the scope of JEDM? (if
>         applicable: to the targeted special issue?)
>
> The topic itself, categorizing subsequences of actions in a learning
> management system, is quite relevant.
>
>
> 2.      How novel is the described research? Are the authors aware of
>         related work?
>
> It is somewhat novel, but the authors do not seem to be adequately aware of
> related work in the methodology contribution. They seem to be even less
> aware of substantive work in education.

Thank you. We agree with your assessment of relevance to JEDM, and concede
that our knowledge of related work in education is a potential blind spot
for us.

> 3.      What is the scientific contribution of this submission? Is it
>         clearly explained, in terms of how the paper advances the EDM field
>         or contributes to related fields?
>
> This paper presents a novel method for learning categories of action
> sequences from MOOC clickstream logs. Although the authors have done a bit
> of background research on approaches to MOOC activity sequences, the
> educational content knowledge is very thin. The approach is under-theorized
> and does not adequately address the “why are you doing this” question.

We agree that we are weak on the educational aspect of the related work. We
will attempt to revise this based on your later comments. We also agree
that we may need to more clearly articulate our answer to the "why"
question from a more education-oriented perspective. Our current
description is perhaps too technically motivated and insufficiently
addresses this particular perspective.

> The paper is on shaky ground because it attempts to do two things and
> does neither well. The first is develop a new method. The second is apply
> the method to learn about behavior patterns in MOOCs.

We respectfully disagree that we do neither well. We can do a better job of
framing our work, but we feel that the model we use in this work is novel
in the education domain and that the behavior patterns we uncover are
useful and important.

> The method itself is a twist on Markov switching models or on mixtures of
> Markov Chains and appears to be very related to the following works, which
> are not referenced:
>
> Hamilton, J. D. (1990). Analysis of time series subject to changes in
> regime. Journal of Econometrics, 45(1-2):39–70.
>
> Song, Y., Keromytis, A. D., & Stolfo, S. J. (2009, February). Spectrogram:
> A Mixture-of-Markov-Chains Model for Anomaly Detection in Web Traffic. In
> NDSS (Vol. 9, pp. 1-15).
>
> Gupta, R., Kumar, R., & Vassilvitskii, S. (2016). On Mixtures of Markov
> Chains. In Advances in Neural Information Processing Systems (pp.
> 3441-3449).
>
>
> The last reference is admittedly recent and used spectral methods. However,
> the important point is that these above references all offer more
> systematic descriptions of model properties, for example by studying the
> recoverability of parameters in a simulation study. (Note that the Gupta et
> al. reference showed that EM estimation was error-prone and needed to be
> run on the order of 1000 times to recover parameters as well as the
> spectral approach.)

We agree that our related work section is weak in the breadth sense and
perhaps focuses too much on MOOC-specific or CS-specific research and did
not cover enough related work in wider domains. We wholeheartedly thank you
for your suggested references, which we will incorporate into a
substantially revised related work section that more clearly articulates
how our model differs from similar models that have been applied in other
domains.

We will attempt to also do a better job of discussing the potential
limitations and drawbacks of our method, including those related to the
application of the EM algorithm for fitting our model, in a new section
added to the paper.

> 4.      Is the work technically sound? Are there enough methodological
>         details? Are claims convincingly substantiated, either through
>         theoretical argument or empirical data?
>
> The current manuscript does not adequately develop the methodology for
> review on its technical soundness, and therefore the application is
> premature. Much of the substantive conclusions are the results of
> Rorschach-test-y looking at graphs.

While there are many results that do indeed require qualitative assessment
of latent state representations and transitions, we disagree that this
makes the method technically unsound. Our goal is to uncover patterns that
exist in this massive clickstream log data, which without a model are
prohibitively expensive for instructors to discover.

> There are selective statistical tests performed which violate some basic
> principles of statistical practice.

We cannot adequately address this criticism without a clear articulation of
exactly where in the paper such mistakes were made, and what the proper
practice should be in those cases.

> The issue of temporal grain size/resolution seems important, but is very
> mysterious once the method is actually described. As it is sold, one would
> expect the model itself to determine when the (latent) state has switched.
> But it appears that the authors simply chose a cutoff to divide sequences
> into “one day’s worth of activity.” Why is that a reasonable choice? And
> why should tomorrow’s latent state depend on today’s state (which they do
> in a HMM)?

Choosing the cutoff points for what defines a browsing session is indeed
an important assumption that will impact the patterns that are uncovered.
However, rather than viewing this as a weakness, we view this as a strength
in the sense that this decision depends entirely on the granularity of the
patterns you wish to discover.

The model still does determine when the latent state switches, but you are
correct in the sense that when it is allowed to determine a switch occurs
is dependent on how the data is segmented and is not completely flexible in
the model itself. It _is_ flexible in the sense that you can change the
segmentation strategy you employ on the sequence data before running the
model, though.

We will attempt to do a better job of explaining this assumption, and why
we chose the segmentation strategy we did for our experiments. We can also
discuss when it might be appropriate to chose a different segmentation
depending on the kinds of patterns you wish to uncover.

> Also, the semantics/combinatorics is unclear. The authors make
> it sound like the observable “state” is a sequence, which would suggest a
> staggering cardinality of the observable variable (O(10^L) for sequences of
> length L). Rather, it seems that the observable is one of N possible Markov
> chains, corresponding to the cardinality of latent variable.

We do not believe there is an issue here. The observable variable is indeed
a sequence, whose cardinality is indeed exponentially large in the number
of available actions for that sequence. However, the way that we model the
_generation_ of these sequences, and thus the space of model parameters, is
as you describe. Each sequence is modeled as being generated from one of
the K Markov models: its initial state is chosen based on the initial state
distribution, and then each subsequent action is chosen by sampling a
transition from the current state to the next state.

I will attempt to make an analogy with text documents. The space of all
possible English journal articles is unfathomably large, but we can still
model the generation of these articles using simpler statistical models. If
we have a discrete distribution over words, we can model the generation of
a document by repeatedly sampling the next word in the document from this
discrete distribution. This model is obviously naive and doesn't capture
everything we know about how documents are generated, but it is
"sufficient" in the sense that for any text document, the contents of it
can in fact be modeled by repeated sampling of words from some
distribution. We do not require exponentially many parameters to model
document generation, and similarly here we do not require exponentially
many parameters to model sequence generation.

> While the idea of breaking up a longer sequence of observed actions into
> shorter subsequences which are each characterized by a latent state is
> interesting, the usefulness of this approach in the present paper is less
> clear. The authors overstate the degree to which this method will help to
> automatically infer student behaviors, as much of the interpretability of
> the results still rests on human readings of the Markov chains.

We will modify the paper to be more precise about our claims about what is
automatically discovered and what is manually discovered. We still feel,
however, that this method is useful in the sense that these patterns cannot
be discovered from the raw clickstream data manually by instructors without
the application of a model that can provide statistical summaries of the
data such as the ones provided by our model.

> Moreover, the discoveries do not go beyond obvious findings like
> “watching video” states and “reading forum thread” states.

While it is indeed sometimes challenging to discover what the meaning of a
state is, we argue that there are more non-obvious findings later in the
section, where we discover that a student's probability of being in one
particular state that indicates more diverse activity when taking quizzes
correlates negatively with grade outcomes.

> It’s hard to say this from reviewing the figures alone, but the states
> seem to be characterized almost entirely by the self-loops in the Markov
> chain. If that is the case, then surely simpler averaging methods are
> reasonable.

It is indeed the case that the latent state transitions are _mostly_
described by self-loops, but we make a claim in table 2 that the latent
state transition probabilities are also important in that the "low"
performing students tend to transition out of state 2 more than the
"perfect" performing students (essentially saying that students that get
perfect quiz scores are more likely to continue to participate heavily in
the forum between browsing sessions compared to students that performed
poorly on quizzes). You can also see this visibly if you compare Figure
5(b) and Figure 5(c)'s edge weight from transitioning from state 2 to state
1, which is significantly darker in 5(c) than 5(b).

> 5.      Do the authors describe the limitations of their approach in a
>         satisfactory manner?
>
> I would not say so. The authors seem to suggest that this method solves all
> of the problems of inferring behavior in MOOC learner data.

We did not intend to come across this way. We will add a new section to the
paper that discusses the limitations and drawbacks of our approach, and how
these might be addressed in practice. We will also better highlight what
gaps remain in this section and in our future work.

> 6.      How significant is the research? Will the paper be likely to have
>         an impact on the community?
>
> The method is not convincingly developed, and the results offer no insights
> that seem to merit this approach. If at least the method were better
> described, I think the community might find useful applications of it.

We appreciate this feedback. We will attempt to better describe the model
and more clearly articulate why this particular approach is important
compared to other approaches.

> 7.      Does the title of this paper clearly and sufficiently reflect its
>         contents?
>
> I would suggest that the authors consider an alternative to the name
> “two-layer hidden Markov model.” The model does not have two hidden layers,
> and the mixture/switching of Markov chains seems more apt. The abbreviation
> TL for two-layer also seems ill-advised, since three-layer would have the
> same result.

Perhaps "layers" is making it seem too close to neural network literature,
but we do feel that there are two different "levels" of pattern extraction,
one which models the within-session patterns and one that models the
between-session patterns.

Thanks for pointing out the potential ambiguity in our abbreviation. We
hadn't considered this since we only focused on two layers.

> 8.      Are the presentation, organization and length satisfactory?
>
> The introductory section should be shortened, since the details on prior
> work are delayed for too long.

Thank you for the suggestion.

> 9.      Can you suggest additions or amendments or an introductory
>         statement that will increase the value of this paper?
>
> The method needs significantly more rigorous development.  Stylistically,
> the authors should try to be less boastful about the utility of this
> method, which is much more dubious than they make out.

We will be more clear about exactly the contributions we are making that
are sufficiently supported by our existing experiment results.

> 10.  Can you suggest any reductions in the paper, or deletions of parts?
>
> I don’t think it is necessary to reproduce

> 11.  Are the illustrations and tables necessary and acceptable?
>
> Table 2 has serious flaws. The counts are not independent.

I don't know how to address this without a concrete suggestion.

> 12.  Are the key words and abstracts/summary informative?
>
> 13.  Is the paper (c) acceptable for publication with major revisions only
>
> 14.  To what extent does this paper fall within your area of expertise?  (
>      ) fully (X) well ( ) partially ( ) hardly
>
> 15.  Please list any other general comments or specific suggestions below.

Thank you again for your detailed comments. We hope that our revised
manuscript adequately addresses your concerns.

### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md
Minor tweaks to response to reviewer c.

@@
 in the education domain and that the behavior patterns we uncover are
-useful and important.
+useful, unique, and important.
 
@@
 exist in this massive clickstream log data, which without a model are
-prohibitively expensive for instructors to discover.
+prohibitively expensive for instructors to discover. We view our model as a
+component in a human-machine collaborative system in which the
+computational techniques employed by the machine are applied to do what
+they are good at (namely, uncovering statistical patterns/trends) and the
+human effort is leveraged to uncover actionable knowledge based upon these
+statistical discoveries.
 
@@
 exactly where in the paper such mistakes were made, and what the proper
-practice should be in those cases.
+practice should be in those cases. We would benefit greatly from more
+elaboration on this point.
 

### 2017-04-04 sec-limitations.tex
Add an explicit section to discuss the limitations of our model.

\section{Limitations and Potential Drawbacks}

There are a few limitations of our model that are important to highlight.
First, there are specific technical limitations due to the statistical
nature of the model and the particular methodology we propose for fitting
our model parameters. Second, there are limitations in the kinds of
patterns our model is able to discover in its current formulation, and the
ease with which instructors are able to extract knowledge from these
patterns. We address both of these lines below.

\subsection{Technical Limitations and Implementation Challenges}
In particular, because the model is complex and has many parameters,
to truly uncover relevant patterns in the data, a large amount of it must
be available to the training algorithm. The assumption that we have a large
amount of sequence data available for training on should generally hold for
most MOOC courses, but this assumption may be problematic if attempting to
apply our model to smaller online (or on-campus) courses.

Our model fits its parameters using maximum likelihood estimation using the
EM algorithm. The EM algorithm is a hill climbing algorithm that is
optimizing in a highly non-convex parameter space. Thus, it can only
guarantee reaching a local maximum in practice~\citep{Dempster:1977:JRSS}.
This may mean that the parameters found for the model may not be the
``best'' parameters in a global sense, which may lead to suboptimal latent
state representations and transition patterns. A commonly applied approach
to address this drawback is to run the model multiple times with different
starting points to allow the model to explore a larger portion of the
potential parameter space. One can then compare the log-likelihood of the
data between the models that were started from different initialization
points and select the one that has the highest value. This still does not
guarantee finding a global maximum, but it does help alleviate the
potential for finding a particularly bad local maximum. In practice, we
have found that our model model can be fit to the data fairly quickly on
commodity hardware, so running it multiple times to address this concern is
not computationally unfeasible.

A further complication in blindly applying the EM algorithm to our model is
the fact that the observation probabilities will be incredibly small. The
probability that a specific sequence is generated by a specific Markov
chain (one of our latent states) will decrease exponentially with its
length. While there are general approaches to avoiding numerical underflow
in hidden Markov models, applying the ``scaling'' method proposed by
\citet{Rabiner:1990:RSR} will still result in numeric stability issues in
our case due to the incredibly small sequence-generation probabilities. We
address this in our open-source implementation by computing the trellises
in log-space and using the log-sum-exp trick when we need to take
summations, which exploits the identify
\begin{equation}
    \log \sum_{i=1}^n e^{x_i} = a + \log \sum_{i=1}^n e^{x_i - a}
\end{equation}
where $a$ is typically set to $\max_i x_i$ to improve stability. We did not
encounter further stability issues once we applied these two tricks.

As is the case with traditional hidden Markov models, it is often important
to smooth the underlying model's distributions to ensure that there is a
non-zero probability of generating the observations. We employed a simple
additive smoothing in our implementation with a small additive constant for
all of our transition matrices to avoid this problem.

\subsection{Limitations of Discovered Patterns}

Our proposed model is flexible in the patterns it can discover in two main
ways. The first is that the granularity of the patterns can be adjusted by
changing the segmentation strategy one uses to divide the user action
stream into discrete ``sessions''. The other is in the number of latent
states $K$ that are used to describe the segmented action sequences. One
drawback of these two lines of flexibility is that there is not necessarily
a clear answer as to the ``correct'' approach for both in any given
scenario. Varying the segmentation strategy changes the granularity of the
patterns that are required to be explained by the latent states, which will
change their meaning. Varying the number of latent states increases the
flexibility of the model, but also may result in latent states that are not
substantially different from the other latent states and/or latent states
with very low probability. This flexibility forces a user of our model to
make some assumptions about the patterns the wish to find (granularity,
diversity), and the model itself does not necessarily provide clear
guidance as to what the best approach is.

Furthermore, we have made an implicit assumption that the segmentation
strategy is applied as a pre-processing step (and is most obviously a
deterministic process). The proposed segmentation strategies in this paper
do not specifically allow for the transitioning between the different
latent behavior patterns to occur over different windows of time for
different users: we have made a strong assumption that transitions between
latent states only occur at action sequence boundaries. One can model how
much time a user stays in a given state in terms of the number of sessions
they remain there before transitioning, but it may be better to directly
model the amount of time we expect a user to stay in the given state
directly. In other words, a more powerful model might be one in which the
segmentation and the latent behavior pattern discovery are jointly modeled
in a single probabilistic framework rather than being completely separate
pieces as we investigate in this work.

While the model can discover \emph{patterns} in the data in a fully
automated way, there is still clearly a burden on the user of our model to
interpret the patterns it has discovered to create actionable
\emph{knowledge} about the MOOC from which the data was extracted.
Extracting the patterns is a necessary step towards the creation of
knowledge, and we view our model as a component in a collaborative system
which leverages the machine to perform statistical modeling to extract
patterns which then enable a human actor to extract knowledge and take
actions on the basis of the data. The pattern discovery is an important
and absolutely necessary component, but it is not the end of the knowledge
discovery process.

### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md
Update the related work section based on reviewer comments.

@@
 We agree that we are weak on the educational aspect of the related work. We
-will attempt to revise this based on your later comments. We also agree
+have attempted to revise this based on your later comments. We also agree
 that we may need to more clearly articulate our answer to the "why"
-question from a more education-oriented perspective. Our current
+question from a more education-oriented perspective. Our original
 description is perhaps too technically motivated and insufficiently
@@
 
-We agree that our related work section is weak in the breadth sense and
+We agree that our related work section was weak in the breadth sense and
 perhaps focuses too much on MOOC-specific or CS-specific research and did
 not cover enough related work in wider domains. We wholeheartedly thank you
-for your suggested references, which we will incorporate into a
+for your suggested references, which we have incorporated into a
 substantially revised related work section that more clearly articulates
@@
 
-We will attempt to also do a better job of discussing the potential
+Specifically, we have now called out our model's relation to similar
+two-layered approaches applied in other domains, articulated how our model
+may be viewed as a special case of a more general Hierarchical Hidden
+Markov Model with certain hard constraints, and discussed in more detail
+how our model relates to the literature about mixtures of hidden Markov
+models.
+
+We have also attempted to also do a better job of discussing the potential
 limitations and drawbacks of our method, including those related to the

### 2017-04-04 sec-conclusion.tex
Update the related work section based on reviewer comments.

@@
 of the performance of student learning.
+
+There is more recent work on better learning algorithms for mixtures of
+Markov models~\citep{Gupta:2016:NIPS}. It would be worth exploring whether
+the advances proposed in this and similar work can be applied to our model
+to address some of the concerns surrounding our use of the EM algorithm for
+our parameter estimation.

### 2017-04-04 sec-related-work.tex
Address Cheng's comments.
Update the related work section based on reviewer comments.

@@
 models are widely applicable and have been used for tasks such as speech
-recognition~\cite{Huang:1990:HMM} and part-of-speech
-tagging~\cite{Jurafsky:2009:SLP}, and are a member of the more general
-family of probabilistic graphical models~\cite{Koller:2009:PGM}.
-A major challenge in applying HMMs and other graphical models successfully to
-solve a problem is to design an appropriate architecture of the model, which
-always varies according to specific applications.
+recognition~\cite{Huang:1990:HMM}, part-of-speech
+tagging~\cite{Jurafsky:2009:SLP}, and econometrics~\cite{Hamilton:1990:JoE}
+and are a member of the more general family of probabilistic graphical
+models~\cite{Koller:2009:PGM}.  A major challenge in applying HMMs and
+other graphical models successfully to solve a problem is to design an
+appropriate architecture of the model, which always varies according to
+specific applications.
+
+A multi-layered approach to HMM modeling of sequence data has been
+performed before in other domains. \citet{Zhang:2004:CVPR}, for example,
+explored a two-layer HMM framework for modeling actions in meetings, but
+their definition of ``two-layer'' differs from ours. In their formulation,
+the ``lower-layer'' level is used to label audio-video action sequences
+into basic events, and the ``upper-layer'' is used to label the output of
+the lower-layer to discover higher-level office behavior abstractions. Our
+formulation differs in that we do not feed the labeled sequence of the
+lower level into the input of the higher level. Instead, our lower level is
+actually treated using a non-hidden Markov model, and the higher level is
+modeling transitions between the $K$ different non-hidden Markov models we
+consider.
+
+Our formulation is more closely related to the Hierarchical Hidden Markov
+Model (HHMM) detailed in \citet{Fine:1998:ML}. Here, the ``layers'' are
+modeled by having the hidden Markov model have two kinds of transitions.
+Horizontal transitions move between states within a layer, where vertical
+transitions move between different layers. At the bottom layer lie the
+``production'' states, which output symbols according to some probability
+distribution. Our specific model in this case can be modeled as a HHMM
+where the horizontal transitions between nodes at the highest layer
+(including self-loops) \emph{must} be immediately followed by a vertical
+transition to the lower layer. The output probability distributions over
+symbols in the lower layer ``production'' level are forced to emit only one
+kind of symbol, and vertical transitions are only allowed into the original
+higher-layer state that transitioned down into the lower-layer.
+
+Mixtures of hidden Markov models are also conceptually similar to our
+formulation. \citet{Song:2009:NDSS} explored using a mixture of hidden
+Markov models in the context of anomaly detection in the security domain.
+\citet{Ypma:2002:Springer} use mixtures of HMMs to categorize web pages and
+cluster users by investigating web log data, which is quite similar to the
+clickstream log data we obtain from MOOCs. The major difference between
+our approach and a standard mixture of HMMs is that we also model the
+\emph{transition behavior} between the Markov models that make up our
+model's lowest layer, where a standard mixture of HMMs would ignore the
+potential dependence of the previous sequence's latent state on the next
+sequence's latent state.
 
@@
 
-There a few additional related studies worth mentioning.
+There are a few additional related studies worth mentioning.
 Bayesian Knowledge Tracing~\cite{Corbett:1994:UMUAI} in its basic form uses
@@
 network problem~\cite{Piech:2015:NIPS}.
-\citet{Ypma:2002:Springer} use mixtures of HMMs to categorize web pages and
-cluster users by investigating web log data, which is quite similar to the
-clickstream log data we obtain from MOOCs.

### 2017-04-04 sec-conclusion.tex
Switch to 2L-HMM for the model abbreviation.
Update the reviewer response draft to mention this concrete change.

@@
 techniques that try to predict student outcomes for understanding
-which of the latent behavior patterns discovered by TL-HMM are most predictive
+which of the latent behavior patterns discovered by 2L-HMM are most predictive
 of the performance of student learning.

### 2017-04-04 sec-intro.tex
Switch to 2L-HMM for the model abbreviation.
Update the reviewer response draft to mention this concrete change.

@@
 \emph{flexible to model state changes over different time resolutions}. 
-More specifically, we propose a novel two-layer hidden Markov model (TL-HMM) to
+More specifically, we propose a novel two-layer hidden Markov model (2L-HMM) to
  discover latent student behavior patterns via
 unsupervised learning on large collections of student behavior observation
-sequences. Evaluation results on a MOOC data set on Coursera demonstrate that TL-HMM can effectively 
+sequences. Evaluation results on a MOOC data set on Coursera demonstrate that 2L-HMM can effectively 
 discover a variety of interesting interpretable student behavior patterns at different levels of resolution, many of which are beyond what existing approaches can discover.
-TL-HMM further enables easy use of the discovered patterns for discriminative analysis such as prediction of learning outcome.
+2L-HMM further enables easy use of the discovered patterns for discriminative analysis such as prediction of learning outcome.
 Since our proposed methods are unsupervised, they can potentially be applied to any MOOC data without requiring any manual work to facilitate understanding of student behaviors and their variations, opening up many possibilities for developing intelligent tutoring systems that can adaptive to student behavior.

### 2017-04-04 sec-model.tex
Switch to 2L-HMM for the model abbreviation.
Update the reviewer response draft to mention this concrete change.

@@
 the latent task states gives us a fairly general and powerful two-layer HMM
-(TL-HMM) that can simultaneously learn ``deeply'' the latent task/behavior
+(2L-HMM) that can simultaneously learn ``deeply'' the latent task/behavior
 states and their transitions as well as the corresponding ``micro''
@@
 
-\subsection{Formal Definition of the TL-HMM}
+\subsection{Formal Definition of the 2L-HMM}
 
@@
 where $a_m = o_{t,k,m}$ (for notational conciseness). Our modified EM
-algorithm for TL-HMMs is provided as part of the \meta/
+algorithm for 2L-HMMs is provided as part of the \meta/
 toolkit~\citemeta{}.

### 2017-04-04 sec-results.tex
Fix a few compile errors.
Not sure what changed here; I was able to compile it before...
Move table captions to occur on top of the tables.
JEDM style guide apparently specifies this.
Switch to 2L-HMM for the model abbreviation.
Update the reviewer response draft to mention this concrete change.

@@
 \section{Experiment Results}
-As an analysis tool, the TL-HMM model provides with us the following two
+As an analysis tool, the 2L-HMM model provides with us the following two
 patterns to characterize student behavior:
@@
   \begin{center}
+    \caption{Statistics about the sequences extracted from the two MOOCs.}
+    \label{table:datasets}
     \begin{tabular}{rrrr}
@@
     \end{tabular}
-    \caption{Statistics about the sequences extracted from the two MOOCs.}
-    \label{table:datasets}
   \end{center}
@@
 \subsection{Latent State Representations}
-The TL-HMM is meant to be a tool for exploratory analysis of student
+The 2L-HMM is meant to be a tool for exploratory analysis of student
 behavior. As such, the number of states should be empirically set based on
@@
 finer-grained modeling of student behavior. In our experiments, we explored
-using between 2--6 states. First, we fit a 6-state TL-HMM to the
+using between 2--6 states. First, we fit a 6-state 2L-HMM to the
 \textretrieval{} sequence dataset and show some of the latent state
@@
   \end{subfigure}
-  \caption{Two example states found by the 6-state TL-HMM. (The naming of
+  \caption{Two example states found by the 6-state 2L-HMM. (The naming of
   these states reflects our own interpretation.)}
@@
   \end{subfigure}
-  \caption{Two similar example states found by the 6-state TL-HMM.}
+  \caption{Two similar example states found by the 6-state 2L-HMM.}
   \label{fig:cross-course}
@@
 Figure~\ref{fig:cross-course} we show two latent state representations
-learned by a 6-state TL-HMM, one from \textretrieval{} and one from
+learned by a 6-state 2L-HMM, one from \textretrieval{} and one from
 \sustain{}. These two states were chosen as they are the most similar
@@
 \subsection{Varying the Number of Latent States}
-Our TL-HMM model has a parameter $K$ that sets the number of latent states
+Our 2L-HMM model has a parameter $K$ that sets the number of latent states
 to be learned. We believe that this can allow our model to flexibly
@@
 We observe similar behavior in Figure~\ref{fig:sustain-state-evolution}
-as we increase $K$ when fitting our TL-HMM on the \sustain{} MOOC.\ Again,
+as we increase $K$ when fitting our 2L-HMM on the \sustain{} MOOC.\ Again,
 in the transition between $K=2$ and $K=3$ we discover forum behavior
@@
 latent states. In Figure~\ref{fig:trans-avg} we show the latent state
-transition diagram for a 3-state TL-HMM fit on \textretrieval{}. We can
+transition diagram for a 3-state 2L-HMM fit on \textretrieval{}. We can
 immediately observe two things: \begin{enumerate*}[label=(\arabic*)]
@@
   \end{subfigure}
-  \caption{The latent state transition diagrams for a 4-state TL-HMM fit to
+  \caption{The latent state transition diagrams for a 4-state 2L-HMM fit to
   \protect\textretrieval{} for all students (a) compared to only ``perfect''
@@
 representations to be fixed, in effect only learning initial and transition
-probabilities for the top layer of our TL-HMM.\ We show the updated latent
+probabilities for the top layer of our 2L-HMM.\ We show the updated latent
 state transition diagram in Figure~\ref{fig:trans-perfect}. We can clearly
@@
   \centering
+  \caption{Average rank for ``perfect'' and ``low'' student groups in the
+  ranked lists associated with the four latent states found by a 2L-HMM.
+  $\dagger$ indicates statistically significant different mean ranks at $p
+  < 0.01$ according to an unpaired $t$-test.}
+  \label{table:mean-rank}
   \begin{tabular}{r|rrrrr}
-    \textbf{Group} & \textbf{State 0} & \textbf{State 1}^\dagger &
-    \textbf{State 2}^\dagger & \textbf{State 3} & \textbf{State 2
-    $\rightarrow$ 2}^\dagger\\\hline
+    \textbf{Group} & \textbf{State 0} & \textbf{State
+  1}\textsuperscript\textdagger &
+    \textbf{State 2}\textsuperscript\textdagger & \textbf{State 3} & \textbf{State 2
+    $\rightarrow$ 2}\textsuperscript\textdagger\\\hline
     \textbf{Perfect} & 975.3  & 1001.5 & 999.0  & 1056.5 & 939.6\\
@@
   \end{tabular}
-  \caption{Average rank for ``perfect'' and ``low'' student groups in the
-  ranked lists associated with the four latent states found by a TL-HMM.
-  $\dagger$ indicates statistically significant different mean ranks at $p
-  < 0.01$ according to an unpaired $t$-test.}
-  \label{table:mean-rank}
 \end{table}

### 2017-04-04 sec-intro.tex
Update paper to incorporate review A and review B.
Also update our response to these two reviewers to be more concrete
about the exact changes we have made to the paper and where to find
them.

@@
 
-Our work is motivated by the following observations:
+We view our model as a component of a system that enables collaboration
+between the machine and a human instructor to extract knowledge from these
+large collections of MOOC data. Automatically extracting interpretable
+patterns from the clickstream data associated with MOOCs is a necessary
+step in order for instructors to identify the hidden knowledge in these
+massive interaction datasets. Without the availability of a suitable model
+for identifying these behavioral patterns, instructors are not empowered to
+use this available data to improve their courses without expending
+extraordinary amounts of manual effort.
+
+Our proposed model is motivated by the following observations:
 \begin{enumerate}
@@
       consider~\cite{Faucon:2016:EDM, Kizilcec:2013:LAK, Shih:2010:EDM}. We
-      propose a more flexible model to accommodate different levels of
-      resolution.
+      propose a model that is agnostic to the time resolution considered,
+      allowing it to be applied at different levels of resolution more
+      naturally.
 \end{enumerate}
@@
 patterns that is \emph{probabilistic} for inference and 
-\emph{flexible to model state changes over different time resolutions}. 
-More specifically, we propose a novel two-layer hidden Markov model (2L-HMM) to
- discover latent student behavior patterns via
+does \emph{not} force assumptions about time resolutions, making it
+\emph{flexible to model state changes over different time resolutions} more
+easily.  More specifically, we propose a novel two-layer hidden Markov
+model (2L-HMM) to discover latent student behavior patterns via
 unsupervised learning on large collections of student behavior observation
-sequences. Evaluation results on a MOOC data set on Coursera demonstrate that 2L-HMM can effectively 
-discover a variety of interesting interpretable student behavior patterns at different levels of resolution, many of which are beyond what existing approaches can discover.
-2L-HMM further enables easy use of the discovered patterns for discriminative analysis such as prediction of learning outcome.
-Since our proposed methods are unsupervised, they can potentially be applied to any MOOC data without requiring any manual work to facilitate understanding of student behaviors and their variations, opening up many possibilities for developing intelligent tutoring systems that can adaptive to student behavior.
+sequences.  Evaluation results on a MOOC data set on Coursera demonstrate
+that the 2L-HMM can effectively discover a variety of interesting interpretable
+student behavior patterns at different levels of resolution, many of which
+are beyond what existing approaches can discover. We show that the patterns
+uncovered by the 2L-HMM capture meaningful behavior by quantitatively
+showing that features extracted from a trained 2L-HMM correlate with
+learning outcomes. Since our proposed methods are
+unsupervised, they can potentially be applied to any MOOC data without
+requiring manual annotation effort at the level of sequences. Instead,
+instructors are empowered to use the latent patterns that the 2L-HMM can
+discover from raw data to extract knowledge about the behaviors his/her
+students exhibit in the MOOC.

### 2017-04-04 sec-model.tex
Update paper to incorporate reviewer C's comments.
Also update our response to reviewer C to include more concrete
information about what we have changed in the paper and where to find
these changes.

@@
 actions occur within a 10 hour window of time (and thus our sequences
-roughly correspond to one day's worth of activity).
+roughly correspond to one day's worth of activity). This decision reflects
+our desire to uncover latent state transition behavior at the granularity
+of day-to-day behavior. A different segmentation strategy could be used to
+uncover hour-by-hour behavior, or week-by-week behavior, and this depends
+entirely on the desired time resolution one wishes to be exhibited in the
+transitions between the latent states. Different segmentation strategies
+will result in different underlying training data, and this different
+meanings behind the patterns that the 2L-HMM will extract.
 

### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md
Switch to 2L-HMM for the model abbreviation.
Update the reviewer response draft to mention this concrete change.

@@
 that our knowledge of related work in education is a potential blind spot
-for us.
+for us as CS researchers.
 
@@
 question from a more education-oriented perspective. Our original
-description is perhaps too technically motivated and insufficiently
-addresses this particular perspective.
+description was perhaps too technically motivated and insufficiently
+addresses this particular perspective. We hope our revised introduction
+more clearly articulates why this is an important problem to tackle.
 
@@
 
-We respectfully disagree that we do neither well. We can do a better job of
-framing our work, but we feel that the model we use in this work is novel
-in the education domain and that the behavior patterns we uncover are
-useful, unique, and important.
+We respectfully disagree that we do neither well. We could have done a
+better job of framing our work (which we feel we have improved in the
+revision), but we feel that the model we use in this work is novel in the
+education domain and that the behavior patterns we uncover are useful,
+unique, and important.
 
@@
 
-We cannot adequately address this criticism without a clear articulation of
-exactly where in the paper such mistakes were made, and what the proper
-practice should be in those cases. We would benefit greatly from more
-elaboration on this point.
+We feel that we cannot adequately address this criticism without a clear
+articulation of exactly where in the paper such mistakes were made, and
+what the proper practice should be in those cases. We would benefit greatly
+from more elaboration on this point.
 
@@
 an important assumption that will impact the patterns that are uncovered.
-However, rather than viewing this as a weakness, we view this as a strength
-in the sense that this decision depends entirely on the granularity of the
-patterns you wish to discover.
+However, rather than viewing this as a weakness, we can view this as a
+strength in the sense that this decision depends entirely on the
+granularity of the patterns you wish to discover.
 
@@
 
-We will attempt to do a better job of explaining this assumption, and why
-we chose the segmentation strategy we did for our experiments. We can also
-discuss when it might be appropriate to chose a different segmentation
-depending on the kinds of patterns you wish to uncover.
+We have attempted in the revision to do a better job of explaining this
+assumption, and why we chose the segmentation strategy we did for our
+experiments. We also briefly discuss when it might be appropriate to chose
+a different segmentation depending on the kinds of patterns you wish to
+uncover.
+
+We have also attempted to address the segmentation strategy problem as a
+whole in the new section of the paper dedicated entirely to the discussion
+of the limitations and potential drawbacks with our approach.
 
@@
 We do not believe there is an issue here. The observable variable is indeed
-a sequence, whose cardinality is indeed exponentially large in the number
-of available actions for that sequence. However, the way that we model the
-_generation_ of these sequences, and thus the space of model parameters, is
-as you describe. Each sequence is modeled as being generated from one of
-the K Markov models: its initial state is chosen based on the initial state
-distribution, and then each subsequent action is chosen by sampling a
-transition from the current state to the next state.
+a sequence. The cardinality of the set of all such sequences is indeed
+exponentially large in the number of available actions for a sequence.
+However, the way that we model the _generation_ of these sequences, and
+thus the space of model parameters, is as you describe. Each sequence is
+modeled as being generated from one of the K Markov models: its initial
+state is chosen based on the initial state distribution, and then each
+subsequent action is chosen by sampling a transition from the current state
+to the next state.
 
@@
 we have a discrete distribution over words, we can model the generation of
-a document by repeatedly sampling the next word in the document from this
+a document as repeatedly sampling the next word in the document from this
 discrete distribution. This model is obviously naive and doesn't capture
@@
 can in fact be modeled by repeated sampling of words from some
-distribution. We do not require exponentially many parameters to model
-document generation, and similarly here we do not require exponentially
-many parameters to model sequence generation.
+distribution that covers its vocabulary. We do not require exponentially
+many parameters to model document generation in this setting, and similarly
+in this paper we do not require exponentially many parameters to model
+sequence generation because we make a strong assumption about the kind of
+parametric model that is used to generate the observed sequences.
 
@@
 
-We will modify the paper to be more precise about our claims about what is
-automatically discovered and what is manually discovered. We still feel,
-however, that this method is useful in the sense that these patterns cannot
-be discovered from the raw clickstream data manually by instructors without
-the application of a model that can provide statistical summaries of the
-data such as the ones provided by our model.
+We have attempted to modify the paper to be more precise about our claims
+about what is automatically discovered and what is manually discovered. We
+still feel, however, that this method is useful in the sense that these
+patterns cannot be discovered from the raw clickstream data manually by
+instructors without the application of a model that can provide statistical
+summaries of the data such as the ones provided by our model.
 
@@
 
-We did not intend to come across this way. We will add a new section to the
-paper that discusses the limitations and drawbacks of our approach, and how
-these might be addressed in practice. We will also better highlight what
-gaps remain in this section and in our future work.
+We did not intend to come across this way. We have added a new section to
+the paper that discusses the limitations and drawbacks of our approach, and
+how these might be addressed in practice. We have also attempted to better
+highlight what gaps remain in this section and in our future work.
+
+Specifically, we address what we feel are the two major classes of
+limitations of our method. The first class relates to technical challenges
+and drawbacks due to the particular statistical framework we have proposed
+and the methods we used to perform parameter estimation. This section also
+details some implementation challenges related to numerical stability in
+the face of very small emission probabilities. The second class relates
+more to the interpretability of the patterns we uncover with the current
+model formulation and addresses issues like the pre-processing used to
+segment the action sequences and its impact on the discovered patterns, the
+parameter K for the number of states and its impact on the discovered
+patterns, and the difficulty in choosing the "correct" settings. We also
+acknowledge the manual effort that is still required to go from discovered
+patterns to actionable knowledge.
 
@@
 
-We appreciate this feedback. We will attempt to better describe the model
+We appreciate this feedback. We have attempted to better describe the model
 and more clearly articulate why this particular approach is important
-compared to other approaches.
+and novel compared to other approaches.
 
@@
 Thanks for pointing out the potential ambiguity in our abbreviation. We
-hadn't considered this since we only focused on two layers.
+hadn't considered this since we only focused on two layers. We have
+switched the abbreviation to 2L-HMM instead to avoid this ambiguity.
 
@@
 
-We will be more clear about exactly the contributions we are making that
-are sufficiently supported by our existing experiment results.
+We have attempted be more clear in the introduction about exactly the
+contributions we are making that are sufficiently supported by our existing
+experiment results.
 
@@
 
-I don't know how to address this without a concrete suggestion.
+We would benefit greatly from a concrete suggestion on how to improve the
+reporting in table 2.
 

### 2017-04-04 sec-limitations.tex
Typo: s/the/they

@@
 with very low probability. This flexibility forces a user of our model to
-make some assumptions about the patterns the wish to find (granularity,
+make some assumptions about the patterns they wish to find (granularity,
 diversity), and the model itself does not necessarily provide clear

### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md
Revised the following:  From education perspective, the motivation for our work is two-fold: 1) there is a need to understand student behavior in detail in the online education environment (which is challenging due to the scale of education), and the proposed model addresses this need (as an "intelligent assistant" for humans); 2) the naturally available student interaction data contains valuable knowledge/patterns about student behavior, which the proposed model can help reveal.  We hope our revised introduction more clearly articulates why this is an important problem to tackle.

@@
 description was perhaps too technically motivated and insufficiently
-addresses this particular perspective. We hope our revised introduction
+addresses this particular perspective.
+From education perspective, the motivation for our work is two-fold: 1) there is a need to understand student behavior in detail in the online education environment (which is challenging due to the scale of education), and the proposed model addresses this need (as an "intelligent assistant" for humans); 2) the naturally available student interaction data contains valuable knowledge/patterns about student behavior, which the proposed model can help reveal.
+We hope our revised introduction
 more clearly articulates why this is an important problem to tackle.

### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md
Changed the following: We respectfully disagree that we do neither well. We could have done a better job of framing our work (which we feel we have improved in the revision), but we feel that the model we use in this work is novel in the education domain and that the behavior patterns we uncover are useful, unique, and important. (No existing model can be directly applied to this data to produce the same output as the proposed model produces.)
Revised slightly the following: We view our model as a component in a human-machine collaborative system in which the computational techniques employed

@@
 education domain and that the behavior patterns we uncover are useful,
-unique, and important.
+unique, and important. (No existing model can be directly applied to this data to produce the same output as the proposed model produces.)
 
@@
 they are good at (namely, uncovering statistical patterns/trends) and the
-human effort is leveraged to uncover actionable knowledge based upon these
-statistical discoveries.
+human effort is leveraged to interpret the patterns and turn them into actionable knowledge.
 

### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md
Revised:   we can view this as a strength in the sense that this decision depends entirely on the granularity of the patterns you wish to discover (following our vision of building a software tool based on the proposed model for use as an intelligent assistant for humans).

@@
 articulation of exactly where in the paper such mistakes were made, and
-what the proper practice should be in those cases. We would benefit greatly
-from more elaboration on this point.
+what the proper practice should be in those cases. We would thus benefit greatly from more elaboration on this point.
 
@@
 strength in the sense that this decision depends entirely on the
-granularity of the patterns you wish to discover.
+granularity of the patterns you wish to discover (following our vision of building a software tool based on the proposed model for use as an intelligent assistant for humans).
 

### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md
Changed:  We still feel, however, that this method is useful in the sense that these patterns cannot be discovered from the raw clickstream data manually by instructors without the application of a model that can provide statistical summaries of the data such as the ones provided by our model, and no existing model can be used directly to generate the same behavior patterns as what our model generates.

@@
 instructors without the application of a model that can provide statistical
-summaries of the data such as the ones provided by our model.
+summaries of the data such as the ones provided by our model, and no existing model can be used directly to generate the same behavior patterns as what our model generates.
 

### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md
Minor editing.

@@
 
-We have attempted be more clear in the introduction about exactly the
+We have attempted to be more clear in the introduction about exactly the
 contributions we are making that are sufficiently supported by our existing

### 2017-04-05 jedm-reviews/round-1/reviewer_c_response.md
Several minor typo fixes.

@@
 
-We have also attempted to also do a better job of discussing the potential
+We have also attempted to do a better job of discussing the potential
 limitations and drawbacks of our method, including those related to the

### 2017-04-05 sec-conclusion.tex
Several minor typo fixes.

@@
 is completely general and can be easily applied to analyze the log of any other
-course to enable deep understanding of student behaviors as well as the coorrelations
-of such behaviors as other variables such as grades. We plan to develop a MOOC log analysis
+course to enable deep understanding of student behaviors as well as the correlations
+of such behaviors and other variables such as grades. We plan to develop a MOOC log analysis
 system based on the proposed model to both facilitate education research and 

### 2017-04-05 sec-limitations.tex
Several minor typo fixes.

@@
 potential for finding a particularly bad local maximum. In practice, we
-have found that our model model can be fit to the data fairly quickly on
+have found that our model can be fit to the data fairly quickly on
 commodity hardware, so running it multiple times to address this concern is
@@
 in log-space and using the log-sum-exp trick when we need to take
-summations, which exploits the identify
+summations, which exploits the identity
 \begin{equation}

### 2017-04-05 sec-model.tex
Several minor typo fixes.

@@
 transitions between the latent states. Different segmentation strategies
-will result in different underlying training data, and this different
+will result in different underlying training data, and thus different
 meanings behind the patterns that the 2L-HMM will extract.

### 2017-04-05 sec-related-work.tex
Several minor typo fixes.

@@
 assignment, and jointly learn \emph{representations} for these states,
-which are treated as being \emph{latent} as opposed to pre-definied using
+which are treated as being \emph{latent} as opposed to pre-defined using
 some rule (or set of rules).  Furthermore, our model provides more
 flexibility in how the time segments are defined, allowing for both finer
-(for example, day-by-day) or courser (for example, month-by-month)
+(for example, day-by-day) or coarser (for example, month-by-month)
 granularity. \citet{Shih:2010:EDM} investigated the use of HMM-based

### 2017-04-05 sec-model.tex
Add an intuitive explanation of the EM algorithm.

@@
 facilitate interpretation and analysis of the discovered latent state
-patterns. Next, we present this model more formally and discuss how to
+patterns.
+
+To estimate the parameters of our model, we will use the EM
+algorithm~\citep{Dempster:1977:JRSS} which allows us to perform maximum
+likelihood estimation of the model in the presence of latent (unobserved)
+variables. This algorithm, intuitively, works as follows: first the model
+parameters we wish to estimate are initialized to some random (but valid)
+starting point. Then, we ``guess'' what the latent variable values might
+be given the current model parameters. We can then use this guess to
+re-estimate the model parameters, which will improve their accuracy
+slightly. We can then use these newly estimated parameters to again
+``guess'' what the latent variable values might be, and so on. We repeat
+this process until the parameter estimates no longer shift by much (or,
+equivalently, the log likelihood of the data does not improve by much). The
+computation for the ``guessing'' of latent variable values is somewhat
+involved in the case of HMMs (see Section~\ref{sec:param-estimation} for
+the full details), but the general intuition behind the iterative
+hill-climbing remains valid.
+
+Next, we present this model more formally and discuss how to
 estimate its parameters to uncover these latent patterns in an unsupervised
@@
 \subsection{Parameter Estimation}
+\label{sec:param-estimation}
 To learn the parameters of our model, we may attempt to use maximum

### 2017-04-13 short-paper/poster/poster.tex
Add poster for L@S17.

\documentclass[final,t]{beamer}
\mode<presentation>{\usetheme{I6dv}}
\setbeamerfont{itemize}{size=\normalsize}
\setbeamerfont{itemize/enumerate body}{size=\normalsize}
\setbeamerfont{itemize/enumerate subbody}{size=\normalsize}
\usepackage{amsmath,amsthm,amssymb,latexsym}
\usepackage[english]{babel}
\usepackage[orientation=portrait,scale=1.78]{beamerposter}
\usepackage{mdframed}
\usepackage{tikz}
\usetikzlibrary{calc,arrows.meta,positioning,decorations.pathreplacing}

\usepackage{fancyvrb,color}
\usepackage{subcaption}

%\usepackage{pygmentsdefs}

\def\meta/{\textsc{MeTA}}
\definecolor{answerbg}{rgb}{0.87,0.929,0.996}

\newtheorem{proposedanswer}{Proposed Answer}

\newcommand{\msp}[0]{\\[0.5\baselineskip]}
\newcommand{\vsp}[0]{\vspace{0.21in}}

\newcommand{\textretrieval}{textretrieval-001}

\title{Modeling MOOC Student Behavior with Two-Layer Hidden Markov Models}
\author{Chase Geigle and ChengXiang Zhai}
\institute[University of Illinois]{University of Illinois at
Urbana-Champaign, Department of Computer Science}
\date{}

\begin{document}
\begin{frame}[fragile]
  \begin{columns}[t]
    \begin{column}{0.6\textwidth}
      \begin{block}{Why Use a Two-Layer Hidden Markov Model (TL-HMM)?}
        \begin{itemize}
          \item Student behavior is complicated $\rightarrow$ assuming
            \textbf{latent states} allows \textbf{data} to inform the
            behavior patterns we discover (rather than handwritten rules)
            \msp{}

          \item Student behavior varies over time $\rightarrow$ modeling
            \textbf{behavior transitions} between the latent states is
            important
            \msp{}

          \item Behavior analysis requires flexible time granularity
            $\rightarrow$ model allows for this \textbf{granularity to be
            chosen flexibly} so as to allow modeling behavior patterns over
            different time resolutions
        \end{itemize}
      \end{block}
      \vsp{}
      \begin{block}{Generative Process}
        \begin{itemize}
          \item Student behavior is chunked into \textbf{sessions} with
            flexible granularity (activity gap of $>10$ hours used in our
            study). \msp{}
          \item Within a session, students take actions from an
            \textbf{action set} $\mathbf{A}$.
            \msp{}
          \item Each session is viewed as being ``generated'' from a
            \textbf{latent state} $u$; this generative process is a random
            walk of the Markov model associated with state $u$.
            \msp{}
          \item Between sessions, users \textbf{transition between latent
            states} following another Markov process (hence ``two-layer'').
        \end{itemize}

      \end{block}
      \vsp{}
      \begin{block}{Transitions Between Latent States}
        \begin{figure}
          \centering
          \begin{subfigure}[t]{0.32\textwidth}
            \includegraphics[width=\textwidth,trim={0 2cm 0 0cm}]{../../figures/trans-comp/trans-avg.png}
            \caption{\label{fig:trans-avg}}
          \end{subfigure}%
          \begin{subfigure}[t]{0.32\textwidth}
            \includegraphics[width=\textwidth,trim={0 2cm 0 0cm}]{../../figures/trans-comp/trans-perfect.png}
            \caption{\label{fig:trans-perfect}}
          \end{subfigure}%
          \begin{subfigure}[t]{0.32\textwidth}
            \includegraphics[width=\textwidth,trim={0 2cm 0 0cm}]{../../figures/trans-comp/trans-low.png}
            \caption{\label{fig:trans-low}}
          \end{subfigure}
          \caption{The latent state transition diagrams for a 4-state TL-HMM fit to
            \protect\textretrieval{} for all students (a) compared to only ``perfect''
          students (b) and only ``low'' students (c).}
          \label{fig:trans-comp}
        \end{figure}
      \end{block}
      \vsp{}
      \begin{block}{Latent Variables of Interest}
        Individual \textbf{latent state representations} (right) and
        \textbf{transitions between them} (below). Data: 18,941 students,
        85,240 sequences, avg.\ length of 7.31.
        \begin{itemize}
          \item Size of node $\propto$ probability of visiting during
            random walk
          \item Thickness of edge $\propto$ probability of taking that edge
            from the source node
          \item Parameters learned using modified Baum-Welch
            algorithm\msp{}
        \end{itemize}
        \textbf{Right}:
        \begin{itemize}
          \item Compare state 1 vs state 3: state 1 indicates more varied
            activity during quiz taking (bottom right forum related states
            and transitions) where state 3 indicates more ``pure'' quiz
            taking.
          \item State 2: forum activity\msp{}
        \end{itemize}
        \textbf{Above:}
        \begin{itemize}
          \item ``Perfect'' (b): 100\% on all quizzes
          \item ``Low'' (c): $< 70\%$ on all quizzes (completed all)
          \item Compare: state 1 prevalence (b) vs (c); state 3 prevalence
            (b) vs (c); state 2 prevalence (a) vs (b); state 2
            $\rightarrow$ 1 transition (b) vs (c)
        \end{itemize}
      \end{block}
    \end{column}
    \begin{column}{0.3\textwidth}
      \begin{block}{Latent State Representations}
        \begin{figure}
          \centering
          \begin{subfigure}[t]{0.9\textwidth}
            \includegraphics[width=\textwidth]{../../figures/text-4state/state0.png}
            \caption{\label{fig:state0}State 0}
          \end{subfigure}

          \begin{subfigure}[t]{0.9\textwidth}
            \includegraphics[width=\textwidth]{../../figures/text-4state/state1.png}
            \caption{\label{fig:state1}State 1}
          \end{subfigure}

          \begin{subfigure}[t]{0.9\textwidth}
            \includegraphics[width=\textwidth]{../../figures/text-4state/state2.png}
            \caption{\label{fig:state2}State 2}
          \end{subfigure}
          \begin{subfigure}[t]{0.9\textwidth}
            \includegraphics[width=\textwidth]{../../figures/text-4state/state3.png}
            \caption{\label{fig:state3}State 3}
          \end{subfigure}
          \caption{Example states learned by a 4-state TL-HMM on the
          \textretrieval{} MOOC.}
          \label{fig:states}
        \end{figure}
      \end{block}
    \end{column}
  \end{columns}
\end{frame}
\end{document}

### 2017-04-28 sec-intro.tex
Update abstract/intro to focus on the behavior representation.
This also adds another figure that gives a basic example of an
idealized version of what the behavior representation we propose could
capture.

@@
 \section{Introduction}
+
 The proliferation of massive open online courses (MOOCs) has resulted in a
-profound impact on education. As more and more learners turn to MOOCs to
-educate themselves on various topics, more and more behavioral data is
-being collected as part of the system on which the MOOC is offered. The
-data present in these logs has the power to aid us in understanding the
-behavior of students who take our MOOCs, which is mostly undetectable for
-instructors of these MOOCs today due to its vast scale. As a result, the
-rich data available through these MOOC logs is highly underutilized today.
+profound impact on education. As more and more students participate in
+these novel educational environments, it is of utmost importance that we be
+able to understand the behavioral patterns of students as they interact
+with them. While we can easily observe the changes in behavior of students
+in real classrooms, MOOCs present a challenge due to their hands-off nature
+and sometimes irregular schedule due to being a full-time worker.
+
+At the same time, as more and more learners turn to MOOCs to educate
+themselves on various topics, more and more behavioral data is being
+collected as part of the system on which the MOOC is offered. This
+presents a unique opportunity: the data present in these logs has the power
+to aid us in understanding the behavior of students who take our MOOCs,
+which is mostly undetectable for instructors of these MOOCs today due to
+its vast scale. As a result, the rich data available through these MOOC
+logs is highly underutilized today.
 
@@
 likely to be able to make changes to these courses that positively impact
-learners. While we can easily observe the changes in behavior of students
-in real classrooms, MOOCs present a challenge due to their hands-off
-nature and sometimes irregular schedule due to being a full-time worker.   
-We view this paper as attempting to bridge this gap. Specifically,
-in this paper we propose unsupervised learning methods for automatically
-discovering and characterizing student learning behavior patterns or profiles from large
-collections of click logs associated with MOOCs.
+learners. We view this paper as attempting to bridge this gap.
 
-We view our model as a component of a system that enables collaboration
-between the machine and a human instructor to extract knowledge from these
-large collections of MOOC data. Automatically extracting interpretable
-patterns from the clickstream data associated with MOOCs is a necessary
-step in order for instructors to identify the hidden knowledge in these
-massive interaction datasets. Without the availability of a suitable model
-for identifying these behavioral patterns, instructors are not empowered to
-use this available data to improve their courses without expending
-extraordinary amounts of manual effort.
+How should we represent behavioral patterns, and what does it mean to
+understand changes in student behavior with respect to these patterns?
+These are still very open questions, and are active areas of
+research~\cite{Kizilcec:2013:LAK, Faucon:2016:EDM, Davis:2016:EDM,
+Shih:2010:EDM}. In this paper, we advocate for a particular representation
+of student behavior patterns as well as \emph{behavior transitions} that we
+believe is simultaneously interpretable but also amenable to unsupervised,
+automated discovery via statistical means. Specifically, we choose to
+visualize behavior patterns as labeled directed graphs where node sizes are
+proportional to steady-state probabilities and edge widths are proportional
+to the probability of leaving a node following that edge. We can use this
+same representation for visualizing both the student behavior patterns as
+well as the transitioning behavior between them. In
+Figure~\ref{fig:motivating-example} we show a hypothetical example of the
+kind of output our proposed representation could convey. Here we see two
+different behavioral patterns (\ref{fig:motivating-example-0} and
+\ref{fig:motivating-example-1}) as well as the \emph{transition behavior}
+between these two behavior patterns (\ref{fig:motivating-example-trans}).
+\begin{figure}
+  \centering
+  \begin{subfigure}[t]{0.30\textwidth}
+    \includegraphics[width=\textwidth]{figures/example/state0.png}
+    \caption{Behavior pattern 0\label{fig:motivating-example-0}}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.30\textwidth}
+    \includegraphics[width=\textwidth]{figures/example/state1.png}
+    \caption{Behavior pattern 1\label{fig:motivating-example-1}}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.30\textwidth}
+    \includegraphics[width=\textwidth]{figures/example/trans.png}
+    \caption{Transitions between the two behavior patterns to the
+    left.\label{fig:motivating-example-trans}}
+  \end{subfigure}
+  \caption{An idealized example of what our behavior representation could capture.}
+  \label{fig:motivating-example}
+\end{figure}
+We could infer many things from even such a simple example. The first might
+be that, when students are taking quizzes, they tend to either use the
+forum or the videos for support, but not both. They also tend to take
+quizzes in a sort of ``cycle'' pattern, indicating perhaps that this course
+allows quiz re-takes. Finally, in Figure~\ref{fig:motivating-example-trans}
+we could conclude that users tend to change their quiz-taking behavior over
+time from one that is more video-focused (pattern 0) to one that is more
+forum-focused (pattern 1).
+
+Our goal in this paper is to design a model that can automatically capture
+student behavior in this way via unsupervised learning methods applied to
+large collections of click logs associated with MOOCs. We view our model
+as a component of a system that enables collaboration between the machine
+and a human instructor to extract knowledge from these large collections of
+MOOC data. Automatically extracting interpretable patterns from the
+clickstream data associated with MOOCs is a necessary step in order for
+instructors to identify the hidden knowledge in these massive interaction
+datasets. Without the availability of a suitable model for identifying
+these behavioral patterns, instructors are not empowered to use this
+available data to improve their courses without expending extraordinary
+amounts of manual effort.
 

### 2017-04-28 sec-conclusion.tex
Future work: emphasize comparative analysis opportunities.

@@
 
-Although we only experimented with our model on two MOOCs, the model
-is completely general and can be easily applied to analyze the log of any other
-course to enable deep understanding of student behaviors as well as the correlations
-of such behaviors and other variables such as grades. We plan to develop a MOOC log analysis
-system based on the proposed model to both facilitate education research and 
-help instructors improve course design.
+Although we only experimented with our model on two MOOCs, the model is
+completely general and can be easily applied to analyze the log of any
+other course to enable deep understanding of student behaviors as well as
+the correlations of such behaviors and other variables such as grades. We
+plan to develop a MOOC log analysis system based on the proposed model to
+both facilitate education research and help instructors improve course
+design.
 
-Specifically, our model can be used to produce an ordered list of latent states for each
-student over time. Given these labeled sequences, we could correlate
-course-wide drifts in these latent states with events in a course. For
-example, we might be able to automatically discover difficult or confusing
-parts of a course by noticing spikes in the distribution of students over
-latent states over time.
+Our model can empower many new comparative analyses. For example, we could
+now look at how behavior patterns change between different offerings of
+the same MOOC to understand how changes in course structure or materials
+influences student behavior. Individual students can now also be compared
+against each other or against groups. For example, by decoding the latent
+state sequences for each student, we can measure how ``surprising'' their
+latent state transition sequence is relative to the average we would expect
+according to the model, or to the average ``perfect'' student, etc. We can
+now investigate how certain behavioral patterns correlate with properties
+of a student (e.g., demographics, prior aptitude, etc.). After decoding the
+students' latent state sequences, we could also correlate course-wide
+drifts in these latent states with events in a course. For example, we
+might be able to automatically discover difficult or confusing parts of a
+course by noticing spikes in the distribution of students over latent
+states over time.
 

### 2017-04-30 sec-intro.tex
Revised introduction.

@@
 automated discovery via statistical means. Specifically, we choose to
-visualize behavior patterns as labeled directed graphs where node sizes are
+visualize behavior patterns as labeled directed graphs where a node represents a ``behavior state'' (e.g., watching lecture video, or visiting a forum) and a directed edge indicates a transition from one behavior state to another, and node sizes are
 proportional to steady-state probabilities and edge widths are proportional
@@
 \end{figure}
-We could infer many things from even such a simple example. The first might
+Such a visualized state-transition representation is very informative for describing student behavior. Indeed, we could infer many things from even such a simple example: The first might
 be that, when students are taking quizzes, they tend to either use the
@@
 available data to improve their courses without expending extraordinary
-amounts of manual effort.
+amounts of manual effort (even with which the raw data can still be very hard to interpret).
 
-Our proposed model is motivated by the following observations:
+Our proposed model (as well as the proposed behavior representation) is motivated by the following observations:
 \begin{enumerate}

### 2017-04-30 sec-limitations.tex
Revised limitations.

@@
 ease with which instructors are able to extract knowledge from these
-patterns. We address both of these lines below.
+patterns. We discuss both of these lines below.
 
 \subsection{Technical Limitations and Implementation Challenges}
-In particular, because the model is complex and has many parameters,
-to truly uncover relevant patterns in the data, a large amount of it must
+
+ %irregular behavior
+%patterns In particular, because
+One potential limitation is that the model is complex and has many parameters,
+to truly uncover relevant patterns in the data, thus a large amount of it must
 be available to the training algorithm. The assumption that we have a large
@@
 ``best'' parameters in a global sense, which may lead to suboptimal latent
-state representations and transition patterns. A commonly applied approach
-to address this drawback is to run the model multiple times with different
+state representations and transition patterns. The problem may not be as serious as it 
+appears to be because strong patterns tend to always show up as long as the algorithm reaches a reasonably good local maximum, and the differences of the results tend to be 
+related to weak ``unstable'' patterns which may not be reliable anyway. 
+ A commonly applied approach
+to address the problem of multiple local maxima is to run the model multiple times with different
 starting points to allow the model to explore a larger portion of the
@@
 guarantee finding a global maximum, but it does help alleviate the
-potential for finding a particularly bad local maximum. In practice, we
+potential for finding a particularly bad local maximum.  
+In practice, we
 have found that our model can be fit to the data fairly quickly on
@@
 \subsection{Limitations of Discovered Patterns}
+
+The proposed behavior representation is most suitable for representing
+recurring behavior patterns, which presumably are most interesting 
+to extract from the data, but may not cover all the interesting patterns in the data. 
+It would be interesting to further explore other complementary models such as
+ time series models, which may help capture non-recurring patterns.
 
@@
 actions on the basis of the data. The pattern discovery is an important
-and absolutely necessary component, but it is not the end of the knowledge
-discovery process.
+and absolutely necessary component, without which it would be very difficult, if not impossible, 
+for humans to directly digest the student behavior buried in the large amount of 
+data.  Of course, pattern discovery is only a means to help humans obtain knowledge, not the end of the knowledge discovery process.

### 2017-04-30 sec-conclusion.tex
Revised conclusions.

@@
 % results; argue again for our novelty; discuss future directions
-We proposed a two-layer hidden Markov model for MOOC student behavior
-modeling in an unsupervised learning framework on large collections of
-actions extracted from MOOC log data. This model is different from existing
+As a tool to help instructors and eduation researchers better understand behavior of MOOC students, 
+ we proposed a two-layer hidden Markov model to automatically extract 
+student activity patterns in the form of behavior state-transition graphs from large amounts of 
+MOOC log data. This model is different from existing
 methods in that it treats behavior patterns as a sequence of \emph{latent
@@
 and more active  forum participation as compared with the average students.
+These results show the great potential of the proposed model for serving as a tool
+to help humans discovery knowledge about student behaviors.
 

### 2017-04-30 jedm-reviews/round-1/reviewer_c_response.md
(Whitespace.) Reflow reviewer responses.
Update reviewer responses based on my new edits.
Added "A more important point is that the proposed model is a general model that can be applied to any such student interaction data. It is reasonable to expect that the model would be able to generate even more interesting patterns when it is applied to a data set with such interesting patterns"

@@
 description was perhaps too technically motivated and insufficiently
-addresses this particular perspective.
-From education perspective, the motivation for our work is two-fold: 1) there is a need to understand student behavior in detail in the online education environment (which is challenging due to the scale of education), and the proposed model addresses this need (as an "intelligent assistant" for humans); 2) the naturally available student interaction data contains valuable knowledge/patterns about student behavior, which the proposed model can help reveal.
-We hope our revised introduction
-more clearly articulates why this is an important problem to tackle.
+addresses this particular perspective. From education perspective, the
+motivation for our work is two-fold: 1) there is a need to understand
+student behavior in detail in the online education environment (which is
+challenging due to the scale of education), and the proposed model
+addresses this need (as an "intelligent assistant" for humans); 2) the
+naturally available student interaction data contains valuable
+knowledge/patterns about student behavior, which the proposed model can
+help reveal. We hope our revised introduction more clearly articulates why
+this is an important problem to tackle.
 
@@
 education domain and that the behavior patterns we uncover are useful,
-unique, and important. (No existing model can be directly applied to this data to produce the same output as the proposed model produces.)
+unique, and important. (No existing model can be directly applied to this
+data to produce the same output as the proposed model produces.)
 
@@
 they are good at (namely, uncovering statistical patterns/trends) and the
-human effort is leveraged to interpret the patterns and turn them into actionable knowledge.
+human effort is leveraged to interpret the patterns and turn them into
+actionable knowledge.
 
@@
 articulation of exactly where in the paper such mistakes were made, and
-what the proper practice should be in those cases. We would thus benefit greatly from more elaboration on this point.
+what the proper practice should be in those cases. We would thus benefit
+greatly from more elaboration on this point.
 
@@
 strength in the sense that this decision depends entirely on the
-granularity of the patterns you wish to discover (following our vision of building a software tool based on the proposed model for use as an intelligent assistant for humans).
+granularity of the patterns you wish to discover (following our vision of
+building a software tool based on the proposed model for use as an
+intelligent assistant for humans).
 
@@
 
-I will attempt to make an analogy with text documents. The space of all
+It may be useful to make an analogy with text documents. The space of all
 possible English journal articles is unfathomably large, but we can still
@@
 instructors without the application of a model that can provide statistical
-summaries of the data such as the ones provided by our model, and no existing model can be used directly to generate the same behavior patterns as what our model generates.
+summaries of the data such as the ones provided by our model, and no
+existing model can be used directly to generate the same behavior patterns
+as what our model generates.
 
@@
 correlates negatively with grade outcomes.
+
+A more important point is that the proposed model is a general model that can be applied to any such student interaction data.
+It is reasonable to expect that the model would be able to generate even more interesting patterns when it is applied to a data set with such interesting patterns.
+The specific data set used in our experiments is only to show the capacity of the model to reveal interesting and meaningful patterns in the data.
 

### 2017-04-30 jedm-reviews/round-1/reviewer_c_response.md
Further strengthend our argument about the generality of the model and its potential value for enabling a new kind of education research on student behavior.

--- 2017-04-05 jedm-reviews/round-1/reviewer_c_response.md
@@
 correlates negatively with grade outcomes.
+
+A more important point is that the proposed model is completely general and thus can be applied to analyze any other data set.
+The actual utility of the patterns discovered from a particular data set is inevitably limited by the true patterns that the data set actually contains; it is reasonable to expect the model to be able to reveal even more interesting patterns when applied to data sets that contain such patterns.
 
@@
 1, which is significantly darker in 5(c) than 5(b).
+Moreover, although the magnitude of probabilities of the transitions is small, relative comparison of these probabilities across different "contexts" (e.g., across different student groups or different periods of a course) may reveal sufficiently significant differences to understand how student behavior varies according to the context.
+
+As already mentioned earlier, the specific data set used in our experiments is only to show the potential of the model to reveal interesting and meaningful patterns in the data; the intended main contribution of the paper lies in the novel representation of behavior patterns and the novel algorithm to extract such patterns automatically from data, which would empower all instructors to potentially turn any such interaction data into useful insights and knowledge for improving instruction, as well as enable a novel way of performing data-drive research on student behavior (e.g., by doing large scale analysis of the extracted behavior patterns in association with other "context" variables as mentioned above to reveal detailed understanding of student behavior in general).
 

### 2017-04-30 sec-limitations.tex
Further polished wording about the multiple local maxima where we claimed that the strong patterns tend to always show up.

@@
 ``best'' parameters in a global sense, which may lead to suboptimal latent
-state representations and transition patterns. The problem may not be as serious as it 
-appears to be because strong patterns tend to always show up as long as the algorithm reaches a reasonably good local maximum, and the differences of the results tend to be 
-related to weak ``unstable'' patterns which may not be reliable anyway. 
+state representations and transition patterns. Empirically, however, we observed
+in our experiments that strong patterns tend to always show up as long as the algorithm reaches a reasonably good local maximum, and the differences of the results tend to be 
+related to weak ``unstable'' patterns which may not be reliable anyway. Since
+it is far more important and useful to reveal the strong salient patterns than weak unreliable ones for instructors, the problem might not necessarily affect the utility of the approach so significantly.
  A commonly applied approach

### 2017-06-14 jedm-reviews/round-2/reviewer_b_response.md
Update paper for reviewer b; add response.

> Related works
>
> In prior iterations, the related works did not frame the authors' research
> with respect to similar methods. I think the authors have done an excellent
> job in reworking this section. It is much clearer how the authors' work is
> novel in both the education and data mining/machine learning domains.

Thank you. We greatly appreciate your helpful feedback that helped improve
this aspect of the paper's presentation.

> Abstract and style
>
> The abstract is now sufficiently long and adequately written. The table
> captions have also been fixed. Certain parts of the figures are still hard
> to read, but I understand that the possible fixes here are limited. In that
> regard, the visual issues do not significantly hinder interpretability of
> the concepts.

We again thank you for your help in improving the quality of our abstract.

> Challenges / Limitations
>
> The authors have created a full challenges and limitations section. The
> authors have added a great deal of clarity to the core issues they
> encountered with both the methods and outcomes of their work.

We agree that this added section brings much more clarity to the technical
challenges and potential limitations of our approach, and thank you for the
feedback leading to this improvement to the paper.

> General readability comments:
>
> Previously, I made comments about the readability of the work and the
> authors requested a few examples. I have written down a few quotes from the
> paper along with explanations or fixes for the problems (denoted by an
> arrow ->). It is important to note that these are not all of the issues in
> the paper. I think that authors could still make many improvements to
> increase both the readability and clarity of the work.

Thank you for the detailed list of suggestions. In the revision, we have
incorporated these suggested changes to the wording and have also done
another editing pass to attempt to correct additional errors of similar
form. We hope that the readability and clarity of the work has improved as
a direct result.

### 2017-06-14 sec-intro.tex
Update paper for reviewer b; add response.

@@
 these novel educational environments, it is of utmost importance that we be
-able to understand the behavioral patterns of students as they interact
-with them. While we can easily observe the changes in behavior of students
-in real classrooms, MOOCs present a challenge due to their hands-off nature
-and sometimes irregular schedule due to being a full-time worker.
+able to understand the behavioral patterns students exhibit in these
+environments. While we can easily observe the changes in behavior of
+students in real classrooms, MOOC environments present some significant
+challenges in this regard: the structure of the course itself is more
+hands-off in nature than that of the traditional classroom (in most cases),
+and thus attracts more students that are full-time workers with irregular
+learning schedules.
 
-At the same time, as more and more learners turn to MOOCs to educate
-themselves on various topics, more and more behavioral data is being
-collected as part of the system on which the MOOC is offered. This
-presents a unique opportunity: the data present in these logs has the power
-to aid us in understanding the behavior of students who take our MOOCs,
-which is mostly undetectable for instructors of these MOOCs today due to
-its vast scale. As a result, the rich data available through these MOOC
-logs is highly underutilized today.
+At the same time, this influx of learners turning to MOOC platforms to
+educate themselves directly leads to the collection of larger datasets of
+behavioral data through the platform's logging. This presents a unique
+opportunity: the data present in these logs has the power to aid us in
+understanding the behavior of students who take our MOOCs. However, due to
+the vast scale of these behavioral logs, student behavior patterns are
+mostly undetectable for MOOC instructors and as a result the rich data
+available through MOOC logs is highly underutilized today.
 
-What stands in the way? Clearly, intelligent systems to create concise and
-digestible summaries of the massive amount of interaction data collected
-are needed in order to empower the instructors of these courses. If we can
-understand how users are interacting with our MOOCs, we are much more
-likely to be able to make changes to these courses that positively impact
-learners. We view this paper as attempting to bridge this gap.
+What stands in the way? Instructors require intelligent systems to create
+concise and digestible summaries of the massive amount of interaction data
+collected. If we can understand how users are interacting with our MOOCs,
+we are much more likely to be able to make changes to these courses that
+positively impact learners. We view this paper as attempting to bridge this
+gap.
 
@@
 automated discovery via statistical means. Specifically, we choose to
-visualize behavior patterns as labeled directed graphs where a node represents a ``behavior state'' (e.g., watching lecture video, or visiting a forum) and a directed edge indicates a transition from one behavior state to another, and node sizes are
-proportional to steady-state probabilities and edge widths are proportional
-to the probability of leaving a node following that edge. We can use this
-same representation for visualizing both the student behavior patterns as
-well as the transitioning behavior between them. In
-Figure~\ref{fig:motivating-example} we show a hypothetical example of the
-kind of output our proposed representation could convey. Here we see two
-different behavioral patterns (\ref{fig:motivating-example-0} and
-\ref{fig:motivating-example-1}) as well as the \emph{transition behavior}
-between these two behavior patterns (\ref{fig:motivating-example-trans}).
+visualize behavior patterns as labeled directed graphs where a node
+represents a ``behavior state'' (such as watching a lecture video or
+visiting a forum), a directed edge indicates a transition from one
+behavior state to another, node sizes are proportional to steady-state
+probabilities, and edge widths are proportional to the probability of
+leaving a node following that edge. We can use this same representation for
+visualizing both the student behavior patterns as well as the transitioning
+behavior between them. In Figure~\ref{fig:motivating-example} we show a
+hypothetical example of the kind of output our proposed representation
+could convey.  Here we see two different behavioral patterns
+(\ref{fig:motivating-example-0} and \ref{fig:motivating-example-1}) as well
+as the \emph{transition behavior} between these two behavior patterns
+(\ref{fig:motivating-example-trans}).
 \begin{figure}
@@
 \end{figure}
-Such a visualized state-transition representation is very informative for describing student behavior. Indeed, we could infer many things from even such a simple example: The first might
-be that, when students are taking quizzes, they tend to either use the
-forum or the videos for support, but not both. They also tend to take
-quizzes in a sort of ``cycle'' pattern, indicating perhaps that this course
-allows quiz re-takes. Finally, in Figure~\ref{fig:motivating-example-trans}
-we could conclude that users tend to change their quiz-taking behavior over
-time from one that is more video-focused (pattern 0) to one that is more
-forum-focused (pattern 1).
+Such a visualized state-transition representation is very informative for
+describing student behavior. Indeed, we could infer many things from even
+such a simple example: The first might be that, when students are taking
+quizzes, they tend to either use the forum or the videos for support, but
+not both. They also tend to take quizzes in a sort of ``cycle'' pattern,
+indicating perhaps that this course allows quiz re-takes. Finally, in
+Figure~\ref{fig:motivating-example-trans} we could conclude that users tend
+to change their quiz-taking behavior over time from one that is more
+video-focused (pattern 0) to one that is more forum-focused (pattern 1).
 
@@
 as a component of a system that enables collaboration between the machine
-and a human instructor to extract knowledge from these large collections of
+and a human instructor to extract knowledge from large collections of
 MOOC data. Automatically extracting interpretable patterns from the
 clickstream data associated with MOOCs is a necessary step in order for
-instructors to identify the hidden knowledge in these massive interaction
+instructors to identify the hidden knowledge in massive interaction
 datasets. Without the availability of a suitable model for identifying
-these behavioral patterns, instructors are not empowered to use this
-available data to improve their courses without expending extraordinary
-amounts of manual effort (even with which the raw data can still be very hard to interpret).
+behavioral patterns, instructors are not empowered to use this available
+data to improve their courses without expending extraordinary amounts of
+manual effort (even with which the raw data can still be very hard to
+interpret).
 
-Our proposed model (as well as the proposed behavior representation) is motivated by the following observations:
+Our proposed model (as well as the proposed behavior representation) is
+motivated by the following observations:
 \begin{enumerate}

### 2017-06-14 sec-limitations.tex
Update paper for reviewer b; add response.

@@
 
- %irregular behavior
-%patterns In particular, because
-One potential limitation is that the model is complex and has many parameters,
-to truly uncover relevant patterns in the data, thus a large amount of it must
-be available to the training algorithm. The assumption that we have a large
-amount of sequence data available for training on should generally hold for
-most MOOC courses, but this assumption may be problematic if attempting to
-apply our model to smaller online (or on-campus) courses.
+One potential limitation is that the model is complex and has many parameters
+in order to truly uncover the relevant patterns in the data. In order to
+properly estimate these parameters at training time, a large amount of data
+must be available to the training algorithm. The assumption that we have a
+large amount of sequence data available for training on should generally
+hold for most MOOC courses, but this assumption may be problematic if
+attempting to apply our model to smaller online (or on-campus) courses.
 
@@
 ``best'' parameters in a global sense, which may lead to suboptimal latent
-state representations and transition patterns. Empirically, however, we observed
-in our experiments that strong patterns tend to always show up as long as the algorithm reaches a reasonably good local maximum, and the differences of the results tend to be 
-related to weak ``unstable'' patterns which may not be reliable anyway. Since
-it is far more important and useful to reveal the strong salient patterns than weak unreliable ones for instructors, the problem might not necessarily affect the utility of the approach so significantly.
- A commonly applied approach
-to address the problem of multiple local maxima is to run the model multiple times with different
-starting points to allow the model to explore a larger portion of the
-potential parameter space. One can then compare the log-likelihood of the
-data between the models that were started from different initialization
+state representations and transition patterns. Empirically, however, we
+observed in our experiments that strong patterns tend to always show up as
+long as the algorithm reaches a reasonably good local maximum, and the
+differences of the results tend to be related to weak ``unstable'' patterns
+which may not be reliable anyway. Since it is far more important and useful
+to reveal the strong salient patterns than weak unreliable ones for
+instructors, the problem might not necessarily affect the utility of the
+approach so significantly.  A commonly applied approach to address the
+problem of multiple local maxima is to run the model multiple times with
+different starting points to allow the model to explore a larger portion of
+the potential parameter space. One can then compare the log-likelihood of
+the data between the models that were started from different initialization
 points and select the one that has the highest value. This still does not
 guarantee finding a global maximum, but it does help alleviate the
-potential for finding a particularly bad local maximum.  
-In practice, we
+potential for finding a particularly bad local maximum.  In practice, we
 have found that our model can be fit to the data fairly quickly on
@@
 changing the segmentation strategy one uses to divide the user action
-stream into discrete ``sessions''. The other is in the number of latent
+stream into discrete ``sessions.'' The other is in the number of latent
 states $K$ that are used to describe the segmented action sequences. One
@@
 patterns which then enable a human actor to extract knowledge and take
-actions on the basis of the data. The pattern discovery is an important
-and absolutely necessary component, without which it would be very difficult, if not impossible, 
-for humans to directly digest the student behavior buried in the large amount of 
-data.  Of course, pattern discovery is only a means to help humans obtain knowledge, not the end of the knowledge discovery process.
+actions on the basis of the data. The pattern discovery is an important and
+absolutely necessary component, without which it would be very difficult,
+if not impossible, for humans to directly digest the student behavior
+buried in the large amount of data.  Of course, pattern discovery is only a
+means to help humans obtain knowledge, not the end of the knowledge
+discovery process.

### 2017-06-14 sec-model.tex
Update paper for reviewer b; add response.

@@
 
-Our general idea is to use a probabilistic generative model to model 
-the student activities as recorded in a MOOC log, which means we will 
-assume that all the observed student activities are samples drawn (i.e., ``generated'') from 
-a parameterized probabilistic model. We can then estimate the 
-parameter values of the probabilistic model by fitting the model to a specific
-MOOC log data set. The estimated parameter values could then be treated as 
-the latent ``knowledge'' discovered from the data. Because such a generative model 
-attempts to fit {\em all} the data, it enables us to discover interesting patterns
-that can explain the {\em overall} behavior of a student or the {\em common} behavior patterns shared by many students.
+Our general idea is to use a probabilistic generative model to model the
+student activities as recorded in a MOOC log, which means we will assume
+that all the observed student activities are samples drawn (i.e.,
+``generated'') from a parameterized probabilistic model. We can then
+estimate the parameter values of the probabilistic model by fitting the
+model to a specific MOOC log data set. The estimated parameter values could
+then be treated as the latent ``knowledge'' discovered from the data.
+Because such a generative model attempts to fit \emph{all} the data, it
+enables us to discover interesting patterns that can explain the
+\emph{overall} behavior of a student or the \emph{common} behavior patterns
+shared by many students.
 
-An HMM  is a specific probabilistic generative model with a ``built-in'' state transition system
-that would control the data to be generated by the model, thus it is especially 
-suitable for modeling sequence data~\cite{Rabiner:1990:RSR, Huang:1990:HMM}. At any moment, the HMM would be in 
-one of $k$ states $U=\{u_1, ...,u_k\}$, and at the next moment, the HMM would move to 
-another state stochastically according to a transition matrix that specifies the probability of
-going to state $u_i$ when the HMM is currently in state $u_j$, i.e., $p(u_i|u_j)$. 
-When the HMM is in state $u$, the HMM can generate an observable data point $x$ 
-according to an output probabilistic model $p(x|u)$. Thus if we ``run'' an HMM for 
-$N$ time points denoted by $t=1, ..., N$, the HMM could ``generate'' a sequence of 
-observations $x_1 ... x_N$, where each $x_i$ is an output symbol by 
-going through a sequence of {\em hidden states} $w_1 ... w_N$ where $w_i \in U$ is a state. 
-The association of such a latent sequence of state transitions with the observed symbols makes
-it possible to use HMM to ``decode'' the latent behavior of students behind the surface behavior we directly observe in the log data, allowing for understanding student behavior more deeply than a model with no latent state variables.
+An HMM is a specific probabilistic generative model with a ``built-in''
+state transition system that would control the data to be generated by the
+model, thus it is especially suitable for modeling sequence
+data~\cite{Rabiner:1990:RSR, Huang:1990:HMM}. At any moment, the HMM would
+be in one of $k$ states $U=\{u_1,\ldots,u_k\}$, and at the next moment, the
+HMM would move to another state stochastically according to a transition
+matrix that specifies the probability $p(u_i \mid u_j)$ of going to state
+$u_i$ when the HMM is currently in state $u_j$. When the HMM is in state
+$u$, the HMM can generate an observable data point $x$ according to an
+output probabilistic model $p(x \mid u)$. Thus if we ``run'' an HMM for $N$
+time points denoted by $t=1,\ldots,N$, the HMM could ``generate'' a sequence
+of observations $x_1 \ldots x_N$, where each $x_i$ is an output symbol by
+going through a sequence of \emph{hidden states} $w_1 \ldots w_N$ where
+$w_i \in U$ is a state. The association of such a latent sequence of state
+transitions with the observed symbols makes it possible to use the HMM to
+``decode'' the latent behavior of students behind the surface behavior we
+directly observe in the log data, allowing for understanding student
+behavior more deeply than a model with no latent state variables.
 
 In many ways, the generation process behind an HMM is meant to simulate the
-actual behavior of a student. We may say that students transition through different ``task states'' (or ``behavior states'')  in the process of study. 
-One such task state may be to learn about a topic by mostly watching
-lecture videos, another task state may be to work on quizzes, and yet
-another may be to participate in forum discussions. While in each of these
-different states, the student would tend to exhibit different surface
+actual behavior of a student. We may say that students transition through
+different ``task states'' (or ``behavior states'')  in the process of
+study.  One such task state may be to learn about a topic by mostly
+watching lecture videos, another task state may be to work on quizzes, and
+yet another may be to participate in forum discussions. While in each of
+these different states, the student would tend to exhibit different surface
 ``micro'' behaviors. For example, in the lecture study state, the student
@@
 forum. Note that due to the complexity of the student behavior, it is very
-difficult to accurately {\em prescribe} the specific surface ``micro''
-behavior patterns for each state in advance, especially without  prior knowledge about the students.
-For example, forum activities are likely interleaved with other activities   in every task state and the interleaving pattern can be somewhat irregular with potentially many variations.
-The major motivations for using an HMM are that 
+difficult to accurately \emph{prescribe} the specific surface ``micro''
+behavior patterns for each state in advance, especially without  prior
+knowledge about the students. For example, forum activities are likely
+interleaved with other activities   in every task state and the
+interleaving pattern can be somewhat irregular with potentially many
+variations.  The major motivations for using an HMM are that
 \begin{enumerate*}[label=(\arabic*)]
-    \item it uses a probabilistic model (i.e., the output probability distribution $p(x|u)$ conditioned on each state) to directly capture the inevitable uncertainty in the association of surface ``micro'' activities with their corresponding latent task/behavior state, which is often our main target to discover and characterize, and
+  \item it uses a probabilistic model (the output probability distribution
+    $p(x \mid u)$ conditioned on each state) to directly capture the
+    inevitable uncertainty in the association of surface ``micro''
+    activities with their corresponding latent task/behavior state, which
+    is often our main target to discover and characterize, and
   \item it does not make any assumption about which latent task/behavior
@@
     student would move from one state to another, but instead allows our
-      data to ``tell'' us what kind of associations are most likely,
-      what kind of transitions are most probable, and which states tend to be more long-lasting for any particular set of students. 
+    data to ``tell'' us what kind of associations are most likely, what
+    kind of transitions are most probable, and which states tend to be more
+    long-lasting for any particular set of students.
 \end{enumerate*}
@@
 However, if we use an ordinary HMM to analyze our data, we would treat each
-observed ``micro'' activity (e.g., video watching, or forum post reading)
-as an output symbol, and thus the output distribution $p(x|u)$ for each
+observed ``micro'' activity (such as video watching or forum post reading)
+as an output symbol, and thus the output distribution $p(x \mid u)$ for each
 discovered latent state would be a simple distribution over all kinds of
 observable micro activities recorded in our log data (e.g., 50\% lecture
-watching, 8\% quiz taking, 7\% quiz submission, 2\% course wiki reading,
-...). While such a distribution is meaningful and can already help us
+watching, 8\% quiz taking, 7\% quiz submission, 2\% course wiki reading
+\ldots). While such a distribution is meaningful and can already help us
 interpret the corresponding latent state, it only gives us a rather
@@
 
-Ideally, we want $p(x|u)$ to characterize the directly observable ``micro''
-behavior in more detail to further capture the relations and dependencies
-of these micro activities. To this end, we would treat an {\em entire
-sequence} of micro activities (e.g., one session of activities) as an
-observed ``symbol'' from a latent state, and further model the generation
-of such a sequence with another Markov model where we treat each micro
-activity as an {\em observable} state, and model the transitions between
-these activity states in very much the same way as the state transitions in
-HMM.
+Ideally, we want $p(x \mid u)$ to characterize the directly observable
+``micro'' behavior in more detail to further capture the relations and
+dependencies of these ``micro'' activities. To this end, we would treat an
+\emph{entire sequence} of ``micro'' activities (e.g., one session of
+activities) as an observed ``symbol'' from a latent state, and further
+model the generation of such a sequence with another Markov model where we
+treat each micro activity as an \emph{observable} state, and model the
+transitions between these activity states in very much the same way as the
+state transitions in an HMM.
 

### 2017-06-14 sec-related-work.tex
Update paper for reviewer b; add response.

@@
 method for finding the common behavioral patterns involves a manual
-clustering step to identify behavioral ``motifs'', which is then followed
+clustering step to identify behavioral ``motifs,'' which is then followed
 by an automatic (rule-based) assignment of all sequences to these motifs.
@@
 students in the form of semi-Markov models like our proposed model does,
-but they differ from our work in two major ways. First, they assume that a
-student can belong to only one behavior profile across the entire course
-rather than allowing this profile to change over time.  Because we do not
-have this restriction, our model is also able to learn the transition
-probabilities between the different behavior profiles we discover.
+but they assume that a student can belong to only one behavior profile
+across the entire course rather than allowing this profile to change over
+time. Because we do not have this restriction, our model is also able to
+learn the transition probabilities between the different behavior profiles
+we discover.
 

### 2017-06-14 sec-related-work.tex
Elaborate on Hamilton (1990) to appease reviewer c.

@@
 appropriate architecture of the model, which always varies according to
-specific applications.
+specific applications. For example, \citet{Hamilton:1990:JoE} explores HMMs
+in the context of ``regime-switching''. Essentially, these are HMMs with
+observation distributions such as the multivariate Gaussian or an
+auto-regressive model in order to properly capture real-valued vector data
+in the domain of econometrics. Our model can be seen as such a
+``regime-switching'' model where the output distributions are (observable)
+Markov chains.
 

### 2017-06-14 sec-model.tex
Update for reviewer c comments; add reviewer c response.

@@
 forum''. For each student in the course $\ell \in \mathbf{L}$, we then
-extract a list of action sequences $\mathbf{S}_\ell$ that he or she
-produced as observed in the log, where each sequence $\mathbf{s} \in
-\mathbf{S}_\ell$ is itself a list of actions $(a_1, a_2, \ldots, a_M)$ with
+extract a list of action sequences $\mathbf{O}_\ell$ that he or she
+produced as observed in the log, where each sequence $\mathbf{o} \in
+\mathbf{O}_\ell$ is itself a list of actions $(a_1, a_2, \ldots, a_T)$ with
 each $a_i \in \mathbf{A}$.  Each sequence can be divided flexibly; in this
@@
 
-Each sequence $\mathbf{s} \in \mathbf{S}_\ell$ is associated with a latent
+Each sequence $\mathbf{o} \in \mathbf{O}_\ell$ is associated with a latent
 state $u \in \{1,\ldots,K\}$ (where $K$ is a fixed constant picked in
-advance). The actions within the sequence $\mathbf{s} = (a_1, a_2, \ldots,
-a_M)$ are then modeled as a first-order Markov process conditioned upon $u$
+advance). The actions within the sequence $\mathbf{o} = (a_1, a_2, \ldots,
+a_T)$ are then modeled as a first-order Markov process conditioned upon $u$
 where each action is drawn from a distribution conditioned upon the
@@
 
-Thus, the probability of a sequence $\mathbf{s}$ of length $M$ given its
+Thus, the probability of a sequence $\mathbf{o}$ of length $M$ given its
 latent state $u$ is
 \begin{equation}
-  P(\mathbf{s} \mid \lambda^{(u)}) = P(a_1 \mid \pi^{(u)}) \prod_{i=2}^M P(a_i \mid
+  P(\mathbf{o} \mid \lambda^{(u)}) = P(a_1 \mid \pi^{(u)}) \prod_{i=2}^T P(a_i \mid
   a_{i-1}, A^{(u)})
@@
 We can compute the likelihood of a list of action sequences
-$\mathbf{S}_\ell$ of length $N$ for a student $\ell$ by marginalizing over
+$\mathbf{O}_\ell$ of length $N$ for a student $\ell$ by marginalizing over
 all possible latent state sequences $(v_1, \ldots, v_N) \in U$ as
 \begin{equation}
-  P(\mathbf{S}_\ell \mid \Lambda)
+  P(\mathbf{O}_\ell \mid \Lambda)
   = \sum_{(v_1, \ldots, v_N) \in U} \left(P(v_1 \mid \Lambda)
-  P(\mathbf{s}_1 \mid \lambda^{(v_1)}) \times \prod_{i=2}^N P(v_i \mid
-  v_{i-1}, \Lambda) P(\mathbf{s}_i \mid \lambda^{(v_i)})\right)
+  P(\mathbf{o}_1 \mid \lambda^{(v_1)}) \times \prod_{i=2}^N P(v_i \mid
+  v_{i-1}, \Lambda) P(\mathbf{o}_i \mid \lambda^{(v_i)})\right)
 \end{equation}
 where $\Lambda$ is the set of all model parameters.  In our model, we let
-$\Lambda = (\pi, A, \lambda^{(1)}, \ldots, \lambda^{(K)})$ where $\pi$ and
+$\Lambda = (\pi, A, B)$ where $\pi$ and
 $A$ are the parameters of a first-order Markov model over the latent states
-and each $\lambda^{(i)}$ consists of the parameters for the first-order
-Markov model over action sequences for latent state $i$. Thus $\pi$
-(without superscripts) is an initial probability vector of length $K$ and
-$A$ (without superscripts) is a $K \times K$ transition probability matrix,
-analogous to the case with the individual first-order Markov model
-parameters $\lambda^{(i)}$ for each latent state.
+and $B = (\lambda^{(1)}, \ldots, \lambda^{(K)})$ where each $\lambda^{(i)}$
+consists of the parameters for the first-order Markov model over action
+sequences for latent state $i$. Thus $\pi$ (without superscripts) is an
+initial probability vector of length $K$ and $A$ (without superscripts) is
+a $K \times K$ transition probability matrix, analogous to the case with
+the individual first-order Markov model parameters $\lambda^{(i)}$ for each
+latent state. We have in total $K + K^2 + K(|\mathbf{A}| + |\mathbf{A}|^2)$
+parameters to be estimated from our sequence data.
 
 This can be seen as a modification of the traditional hidden Markov
-model~\cite{Rabiner:1990:RSR} where instead of discrete observations (one
-for each latent state transition) we have observations that take the form
-of entire sequences $\mathbf{s}_i = (a_1, a_2, \ldots a_M)$ whose
-probabilities are computed using another (non-hidden) Markov model
-conditioned upon the latent state $u_i$.
+model with categorical outputs~\cite{Rabiner:1990:RSR} where instead of
+discrete observations (one for each latent state transition) we have
+observations that take the form of entire sequences $\mathbf{o}_i = (a_1,
+a_2, \ldots a_T)$ whose probabilities are computed using another
+(non-hidden) Markov model conditioned upon the latent state $u_i$.
 
@@
 hidden Markov models in an unsupervised setting. In the following sections,
-we will provide a brief description of the original Baum-Welch algorithm,
-and then describe our modification.
+we will provide a brief description of the original Baum-Welch algorithm
+for unsupervised parameter estimation for categorical valued hidden Markov
+models, and then describe our modification to allow for parameter
+estimation for our 2L-HMM modification for sequence valued observations.
 
 \subsubsection{Baum-Welch for Traditional HMMs}
-In the traditional HMM formulation, we have $\Lambda = (\pi, A, B)$ where
-$\pi$ is the initial probability distribution over the latent states,
-$A$ is a $N\times N$ matrix indicating the latent state transition
-probabilities, and $B$ is a $N \times V$ matrix indicating the probability
-of generating a discrete observation symbol from a given latent state. The
-Baum-Welch algorithm (also called the forward-backward algorithm) defines
-two sets of variables $\alpha_t(i)$ called the forward variables and
-$\beta_t(i)$ called the backward variables.
+In the traditional HMM formulation with categorical outputs, we have
+$\Lambda = (\pi, A, B)$ where $\pi$ is the initial probability distribution
+over the latent states (of length $K$), $A$ is a $K\times K$ matrix
+indicating the latent state transition probabilities, and $B = (b_1,
+\ldots, b_k)$ is a length $K$ vector where each entry $b_i$ is a
+probability distribution over the possible discrete output symbols from
+$\mathbf{A}$.
 
-$\alpha_t(i) = P(o_1, \ldots, o_t, q_t = i \mid \Lambda)$ is the probability of
-generating the sequence of observations $(o_1, o_2, \ldots, o_t)$ up to time
-$t$ and arriving in state $i$ at that time. They are typically defined
-using the following recursion:
+The goal of the Baum-Welch algorithm (also called the forward-backward
+algorithm) is to learn the values for the parameters $\Lambda$ from a
+collection of observed sequences of values from $\mathbf{A}$. Concretely,
+our training data $D = (\mathbf{o}^{(1)}, \ldots, \mathbf{o}^{(M)})$ is a
+collection of $M$ sequences $\mathbf{o}^{(k)}$, each of which consists of $T_k$
+symbols (observations) from $\mathbf{A}$. The Baum-Welch algorithm defines
+two sets of variables $\alpha^{(\mathbf{o})}_t(i)$ called the forward
+variables and $\beta^{(\mathbf{o})}_t(i)$ called the backward
+variables~\cite{Rabiner:1990:RSR} for each sequence $\mathbf{o} \in
+\mathbf{D}$.
+
+$\alpha^{(\mathbf{o})}_t(i) = P(o_1, \ldots, o_t, v_t = i \mid \Lambda)$ is
+the probability of generating the sequence of observations $(o_1, o_2,
+\ldots, o_t)$ up to time $t$ and arriving in state $i$ at that time. They
+are typically defined using the following recursion:
 \begin{itemize}
-  \item $\alpha_1(i) = \pi_i b_i(o_1)$, the probability of starting in
+  \item $\alpha^{(\mathbf{o})}_1(i) = \pi_i b_i(o_1)$, the probability of starting in
     state $i$ ($\pi_i$) times the probability of generating the first
     observation $o_1$ from state $i$.
-  \item $\alpha_{t+1}(i) = b_i(o_{t+1})\sum_{j=1}^N \alpha_t(j) A_{ji}$,
-    the probability of generating the observation $o_{t+1}$ from state $i$
-    times the probability that we arrive in state $i$ from any other
-    previous state after generating all of the other observations.
+  \item $\alpha^{(\mathbf{o})}_{t+1}(i) = b_i(o_{t+1})\sum_{j=1}^K
+    \alpha^{(\mathbf{o})}_t(j) A_{ji}$, the probability of generating the observation
+    $o_{t+1}$ from state $i$ times the probability that we arrive in state
+    $i$ from any other previous state after generating all of the other
+    observations.
 \end{itemize}
 
-Analogously, $\beta_t(i) = P(o_{t+1}, \ldots, o_{T} \mid q_t = i, \Lambda)$
-is the probability of generating the \emph{rest of the sequence} given that
-we are in state $i$ at time $t$. They are also defined using a recursion:
+Analogously, $\beta^{(\mathbf{o})}_t(i) = P(o_{t+1}, \ldots, o_{T} \mid v_t
+= i, \Lambda)$ is the probability of generating the \emph{rest of the
+sequence} given that we are in state $i$ at time $t$. They are also defined
+using a recursion:
 \begin{itemize}
-  \item $\beta_T(i) = 1$
-  \item $\beta_t(i) = \sum_{j=1}^N \beta_{t+1}(j) A_{ij} b_j(o_{t+1})$, the
-    probability of transitioning to any state $j$ and generating the
-    observation $o_{t+1}$ times the probability of generating the rest of
-    the sequence given that we transitioned to state $j$.
+  \item $\beta^{(\mathbf{o})}_T(i) = 1$
+  \item $\beta^{(\mathbf{o})}_t(i) = \sum_{j=1}^K
+    \beta^{(\mathbf{o})}_{t+1}(j) A_{ij} b_j(o_{t+1})$, the probability of
+    transitioning to any state $j$ and generating the observation $o_{t+1}$
+    times the probability of generating the rest of the sequence given that
+    we transitioned to state $j$.
 \end{itemize}
 
-Given the $\alpha$s and the $\beta$s, we can compute $\gamma_t(i)$, the posterior
-probability of being in a given state $i$ at time $t$, and $\xi_t(i,j)$,
-the posterior probability of going through a transition from state $i$ to
-state $j$ at time $t$ as
+Given the $\alpha$s and the $\beta$s, we can compute
+$\gamma^{(\mathbf{o})}_t(i)$, the posterior probability of being in a given
+state $i$ at time $t$, and $\xi^{(\mathbf{o})}_t(i,j)$, the posterior
+probability of going through a transition from state $i$ to state $j$ at
+time $t$ as
 \begin{equation}
-  \gamma_t(i) = \frac{\alpha_t(i)\beta_t(i)}{\sum_{j=1}^N
-  \alpha_t(j)\beta_t(j)}
+  \gamma^{(\mathbf{o})}_t(i) =
+  \frac{\alpha^{(\mathbf{o})}_t(i)\beta^{(\mathbf{o})}_t(i)}{\sum_{j=1}^K
+  \alpha^{(\mathbf{o})}_t(j)\beta^{(\mathbf{o})}_t(j)}
 \end{equation}
@@
 \begin{equation}
-  \xi_t(i,j) = \frac{\alpha_t(i) A_{ij} b_j(o_{t+1}) \beta_{t+1}(j)}
-  {\sum_{j=1}^N \alpha_t(j)\beta_t(j)}
+  \xi^{(\mathbf{o})}_t(i,j) = \frac{\alpha^{(\mathbf{o})}_t(i) A_{ij}
+  b_j(o_{t+1}) \beta^{(\mathbf{o})}_{t+1}(j)} {\sum_{j=1}^K
+  \alpha^{(\mathbf{o})}_t(j)\beta^{(\mathbf{o})}_t(j)}
 \end{equation}
-respectively. This is the E-step of the EM algorithm.
+respectively~\cite{Rabiner:1990:RSR}. Computing these variables for each
+sequence $\mathbf{o} \in D$ is the E-step of the EM algorithm.
 
-Given $\gamma_t(i)$ and $\xi_t(i,j)$ for each sequence $\mathbf{o}_k$ in
-some training data of length $M$, we can update our model parameters
-$\Lambda$ as
+Given $\gamma^{(\mathbf{o})}_t(i)$ and $\xi^{(\mathbf{o})}_t(i,j)$ for each
+sequence $\mathbf{o} \in \mathbf{D}$, we can update our model parameters $\Lambda$
+as
 \begin{align}
-  \pi_i &= \frac{\sum_{k=1}^M \gamma^{(\mathbf{o}_k)}_1(i)}{M},\\
-  A_{ij} &= \frac{\sum_{k=1}^M \sum_{t=1}^T \xi^{(\mathbf{o}_k)}_t(i,j)}
-  {\sum_{k=1}^M \sum_{j=1}^N \sum_{t=1}^T \xi^{(\mathbf{o}_k)}_t(i,j)},
+  \pi_i &= \frac{\sum_{\mathbf{o} \in \mathbf{D}}
+  \gamma^{(\mathbf{o})}_1(i)}{|\mathbf{D}|},\\
+  A_{ij} &= \frac{\sum_{\mathbf{o} \in \mathbf{D}} \sum_{t=1}^T
+  \xi^{(\mathbf{o})}_t(i,j)} {\sum_{\mathbf{o} \in \mathbf{D}} \sum_{j=1}^N
+  \sum_{t=1}^T \xi^{(\mathbf{o})}_t(i,j)},
   \text{ and }\\
-  b_i(v) &= \frac{\sum_{k=1}^M \sum_{t=1, o_{k,t} = v}^T
-  \gamma^{(\mathbf{o}_k)}_t(i)}
-  {\sum_{k=1}^M \sum_{t=1}^T \gamma^{(\mathbf{o}_{k,t})}_t(i)}.
+  b_i(a) &= \frac{\sum_{\mathbf{o} \in \mathbf{D}} \sum_{t=1, o_t = a}^T
+  \gamma^{(\mathbf{o})}_t(i)} {\sum_{\mathbf{o} \in \mathbf{D}}
+  \sum_{t=1}^T \gamma^{(\mathbf{o})}_t(i)}
 \end{align}
-This is the M-step of the EM algorithm.
+respectively~\cite{Rabiner:1990:RSR}. This is the M-step of the EM
+algorithm.
 
 \subsubsection{Baum-Welch for HMMs with Sequence Observations}
-The major deviation of our model from the traditional HMM is that our
-observations are themselves sequences. Thus, in all of the equations
-presented in the previous section, we treat an observation $o_t =
-\mathbf{s}_t$ at time $t$ as consisting itself of a list of actions $(a_1,
-\ldots, a_M)$. Thus, instead of defining $b_i(v)$ as the probability that a
-discrete observation \emph{token} $v$ is generated from state $i$, we have
+The major deviation of our model from the traditional categorical-valued
+HMM is that our observations are themselves sequences of actions from
+$\mathbf{A}$ rather than individual tokens. We again denote our parameters
+as $\Lambda = (\pi, A, B)$ where $\pi$ is the initial probability
+distribution over the latent states (of length $K$), $A$ is a $K \times K$
+matrix indicating the latent state transition probabilities, but $B =
+(\lambda^{(1)}, \ldots, \lambda^{(K)})$ is now a vector of length $K$ where
+each element $\lambda^{(i)}$ consists of the parameters for a first-order
+Markov model over the action space $\mathbf{A}$. Recall that each
+$\lambda^{(i)} = (\pi^{(i)}, A^{(i)})$ where $\pi^{(i)}$ is an initial
+action distribution over the $|\mathbf{A}|$ available actions, and
+$A^{(i)}$ is the $|\mathbf{A}| \times |\mathbf{A}|$ transition matrix
+between those actions.
+
+The goal of the Baum-Welch algorithm is still to learn the paramters
+$\Lambda$ for our model. What differs from the categorical-valued HMM case
+is that our data now consists of a collection of \emph{lists} of sequences,
+rather than just a collection of sequences. This means that our observation
+values $\mathbf{o}_t$ are now \emph{themselves} sequences of values from
+$\mathbf{A}$, rather than just being a single token from $\mathbf{A}$.
+Formally, our training data $\mathbf{D} = (\mathbf{O}^{(1)}, \ldots,
+\mathbf{O}^{(|L|)})$, where each $\mathbf{O}^{(\ell)} = (\mathbf{o}_1,
+\ldots, \mathbf{o}_{T_\ell})$ is itself a list of sequences. Each sequence
+$\mathbf{o}_k \in \mathbf{O}^{(\ell)}$ consists of a list of actions $(a_1,
+\ldots, a_{T_k})$, each of which is a member of $\mathbf{A}$.
+
+In order to run the forward-backward algorithm for an element $\mathbf{O}
+\in \mathbf{D}$ to compute the $\alpha$ and $\beta$ recursions like before,
+we must define $b_i(\mathbf{o}_t)$ in this setting where $\mathbf{o}_t =
+(a_1, \ldots, a_{T_t})$ is now a sequence instead of a single token. We
+define it as follows:
 \begin{equation}
-  b_i(\mathbf{s}_t) = P(\mathbf{s}_t \mid \lambda^{(i)})
-  = P(a_1 \mid \pi^{(i)}) \prod_{k=2}^M P(a_k \mid
-  a_{k-1}, A^{(i)})
+  b_i(\mathbf{o}_t) = P(\mathbf{o}_t \mid \lambda^{(i)})
+  = P(a_1 \mid \pi^{(i)}) \prod_{k=2}^{T_t} P(a_k \mid a_{k-1}, A^{(i)})
+  = \pi^{(i)}_{a_1} \prod_{k=2}^{T_t} A^{(i)}_{a_{k-1},a_{k}}.
 \end{equation}
-where $b_i(\mathbf{s})$ now defines the probability of an action sequence
-using a first-order (non-hidden) Markov model.
+Fortunately, the recursions for $\alpha^{(\mathbf{O})}_t(i)$ and
+$\beta^{(\mathbf{O})}_t(i)$ remain the same, as do the definitions of
+$\gamma^{(\mathbf{O})}_t(i)$ and $\xi^{(\mathbf{O})}_t(i,j)$, in the
+E-step. We can simply substitute the new definition for the output
+distribution $b_i(\mathbf{o}_t)$ in those equations.
 
-Fortunately, the recursions for the $\alpha$ and $\beta$ variables remain
-the same, as do the definitions of $\gamma_t(i)$ and $\xi_t(i,j)$ in the
-E-step. The only change is in the updating equations in the M-step,
-where we replace the update for $b_i(v)$ by a pair of updates: one for
-$\pi^{(i)}_a$ one for $A^{(i)}_{ab}$. We denote the $m$-th action in the
-$t$-th sequence for the $k$-th student as $o_{k,t,m}$. The two updates can
-be written as
+The substantial change is in the updating equations in the M-step,
+where we replace the update for $b_i(a)$ (which used to be a categorical
+distribution) by a pair of updates for the Markov chain for state $i$: one
+for $\pi^{(i)}_a$ one for $A^{(i)}_{ab}$. The two updates can be written as
 \begin{align}
   \pi^{(i)}_{a}
-  &= \frac{\sum_{k=1}^M \sum_{t=1,a_1 = a}^T \gamma^{(\mathbf{o}_k)}_t(i)}
-  {\sum_{k=1}^M \sum_{t=1}^T \gamma^{(\mathbf{o}_k)}_t(i)}, \text{ and }\\
+  &= \frac{\sum_{\mathbf{O} \in \mathbf{D}} \sum_{\mathbf{o}_t \in
+  \mathbf{O},\mathbf{o}_{t,1} = a} \gamma^{(\mathbf{O})}_t(i)}
+  {\sum_{\mathbf{O} \in \mathbf{D}} \sum_{\mathbf{o}_t \in \mathbf{O}}
+  \gamma^{(\mathbf{O})}_t(i)}, \text{ and }\\
   A^{(i)}_{ab}
-  &= \frac{\sum_{k=1}^M \sum_{t=1}^T
-  \sum_{m=2,a_{m-1}=a \land a_m = b}^{|\mathbf{o}_{k,t}|}
-  \gamma^{(\mathbf{o}_k)}_t(i)}
-  {\sum_{k=1}^M \sum_{t=1}^T \sum_{m=2}^{|\mathbf{o}_{k,t}|}
-  \gamma^{(\mathbf{o}_k)}_t(i)}
+  &= \frac{\sum_{\mathbf{O} \in \mathbf{D}} \sum_{\mathbf{o}_t \in \mathbf{O}}
+  \sum_{m=2,\mathbf{o}_{t,m-1}=a \land \mathbf{o}_{t,m} = b}^{|\mathbf{o}_{t}|}
+  \gamma^{(\mathbf{O})}_t(i)}
+  {\sum_{\mathbf{O} \in \mathbf{D}} \sum_{\mathbf{o}_t \in \mathbf{O}}
+  \sum_{m=2,\mathbf{o}_{t,m-1}=a}^{|\mathbf{o}_{t}|} \gamma^{(\mathbf{O})}_t(i)}.
 \end{align}
-where $a_m = o_{t,k,m}$ (for notational conciseness). Our modified EM
-algorithm for 2L-HMMs is provided as part of the \meta/
+These updates can be understood as follows. $\pi^{(i)}_a$ is the
+probability that a sequence being generated by state $i$ begins with action
+$a$. $\gamma^{(\mathbf{O})}_t(i)$ gives the probability of generating the
+sequence $\mathbf{o}_t$ from state $i$, so we simply aggregate this for all
+sequences $\mathbf{o}_t$ where the first action is $a$. We then normalize
+this distribution to sum to 1 across all possible actions $a \in
+\mathbf{A}$ to obtain our new estimate for $\pi^{(i)}_a$.  $A^{(i)}_{ab}$
+is the probability that a sequence being generated by state $i$ currently
+at action $a \in \mathbf{A}$ transitions to action $b \in \mathbf{A}$.
+Thus, we compute the expected number of times we observe a transition from
+$a$ to $b$ in a sequence generated by state $i$, and we normalize this
+distribution to sum to 1 across all possible actions $b \in \mathbf{A}$.
+Our modified EM algorithm for 2L-HMMs is provided as part of the \meta/
 toolkit~\citemeta{}.

### 2017-06-14 sec-related-work.tex
Update for reviewer c comments; add reviewer c response.

--- 2017-06-14 sec-related-work.tex
@@
 specific applications.
+
+For example, in part-of-speech tagging~\cite{Jurafsky:2009:SLP} the output
+distributions are categorical (distributions over words from a fixed
+vocabulary) and the latent states represent the part-of-speech category for
+a word. In speech recognition~\cite{Huang:1990:HMM} the output
+distributions might be mixtures of Gaussians to predict real-valued vectors
+extracted from short windows of a speech signal. In the domain of
+econometrics, \citet{Hamilton:1990:JoE} explores HMMs in the context of
+``regime-switching.'' In this framing, the goal is to understand how
+econometric data changes by modeling discrete changes in ``regime'' as
+having an impact on the resulting real-valued vector data observed. The
+``regimes'' are represented with some sort of model that can produce
+real-valued vector data, such as a multivariate Gaussian or an
+auto-regressive model. The analogy with HMMs is that a ``regime'' is a
+latent state, and the characterization of the regime itself is the output
+distribution for that latent state. Our model can be seen as such a
+``regime-switching'' model where the output of the ``regimes'' that
+students are switching between are discrete-valued \emph{sequences} (as
+opposed to real numbers, vectors of real numbers, or categorical symbols)
+and the model used to represent a specific ``regime'' is an (observable)
+Markov chain over the observed student actions. We view the switching
+between ``regimes'' as the first ``layer'' of our model, and the
+transitioning behavior \emph{within} a ``regime'' between the actions a
+student takes as the second ``layer'' of our model.
 

### 2017-06-18 sec-conclusion.tex
Incorporate new citation from the editor.
Also shuffle around some figures to make the page layout prettier with
the now longer related work and model "derivation" sections.

@@
 which of the latent behavior patterns discovered by 2L-HMM are most predictive
-of the performance of student learning.
+of the performance of student learning. One could also relax this somewhat
+and posit that \emph{groups} of students transition between the lower layer
+patterns our method identifies in distinct ways and instead do a soft
+clustering of students into $K_2$ groups based on the similarity of their
+transitioning behavior between the higher-level behaviors we can identify.
 

### 2017-06-18 sec-related-work.tex
Incorporate new citation from the editor.
Also shuffle around some figures to make the page layout prettier with
the now longer related work and model "derivation" sections.

@@
 into basic events, and the ``upper-layer'' is used to label the output of
-the lower-layer to discover higher-level office behavior abstractions. Our
-formulation differs in that we do not feed the labeled sequence of the
-lower level into the input of the higher level. Instead, our lower level is
-actually treated using a non-hidden Markov model, and the higher level is
-modeling transitions between the $K$ different non-hidden Markov models we
-consider.
+the lower-layer to discover higher-level office behavior abstractions.
+\citet{Oliver:2004:CVIU} propose a similar layered HMM approach
+for modeling office activity at multiple different levels of time
+granularity. In their approach, each layer $L$ is represented as a ``bank''
+of $K$ different HMMs that model sequences of some length $T_L$. At the
+bottom layer ($L = 1$), the bank of $K$ HMMs corresponding to that layer is
+run on some initial observation data, considering windows of observations
+of length $T_1$. Then an output is generated by using the inferential
+results of these $K$ HMMs to make a prediction: which of the $K$ HMMs was
+most likely to produce that sequence of observations? This output is then
+fed to the next layer of HMMs, which considers sequences of length $T_2$
+and outputs prediction results as to which of the $K$ HMMs at layer two
+were mostly likely to produce the sequence of outputs produced by the
+previous layer, and so on.
 
-Our formulation is more closely related to the Hierarchical Hidden Markov
-Model (HHMM) detailed in \citet{Fine:1998:ML}. Here, the ``layers'' are
-modeled by having the hidden Markov model have two kinds of transitions.
-Horizontal transitions move between states within a layer, where vertical
-transitions move between different layers. At the bottom layer lie the
-``production'' states, which output symbols according to some probability
-distribution. Our specific model in this case can be modeled as a HHMM
-where the horizontal transitions between nodes at the highest layer
+Our formulation differs from both in that we do not feed the labeled
+sequence of the lower level into the input of the higher level. Instead,
+our lower level is actually treated using a non-hidden Markov model, and
+the higher level is modeling transitions between the $K$ different
+non-hidden Markov models we consider. The problem to be solved is similar
+in that we wish to predict a ``label'' for a sequence of actions a student
+takes as well as understand the transition behavior between those labels.
+However, one of the consequences of modeling the lower layer using a
+non-hidden Markov model instead of an HMM directly is that the meaning of
+the $K$ different latent states can be more clearly captured by using our
+proposed behavior representation. If we were to use an HMM as our first
+layer, the behavior patterns (like from
+Figure~\ref{fig:motivating-example}) would instead have nodes that
+represent another set of latent states instead of being concrete actions
+themselves. Each of these latent states would then be associated with some
+other output distribution over the possible concrete actions to be
+considered. In order to understand a single behavior pattern uncovered, one
+would first have to understand the different output distributions for the
+latent states in that pattern in order to understand the meaning of that
+latent state. This further complicates the understanding of higher level
+patterns because understanding the higher layer patterns requires
+understanding the lower layer patterns. By taking a more restrictive view
+on the first layer, we are able to produce a representation that can be
+more readily interpreted due to the states in our first layer
+representation having an immediately clear meaning (the concrete action
+they represent).
+
+The 2L-HMM model we propose is more closely related to the Hierarchical
+Hidden Markov Model (HHMM) detailed in \citet{Fine:1998:ML}. Here, the
+``layers'' are modeled by having the hidden Markov model have two kinds of
+transitions.  Horizontal transitions move between states within a layer,
+where vertical transitions move between different layers. At the bottom
+layer lie the ``production'' states, which output symbols according to some
+probability distribution. Our specific model in this case can be modeled as
+a HHMM where the horizontal transitions between nodes at the highest layer
 (including self-loops) \emph{must} be immediately followed by a vertical

### 2017-06-20 sec-conclusion.tex
Minor revision to related work, the parameter estimation section, conclusions, and acknowledgments.

@@
-\section{Conclusion}
+\section{Conclusions and Future Work}
 % summarize the main contributions of the paper; summarize the experimental
@@
 
-\subsection{Future Work}
+%\subsection{Future Work}
 
-Although we only experimented with our model on two MOOCs, the model is
-completely general and can be easily applied to analyze the log of any
-other course to enable deep understanding of student behaviors as well as
-the correlations of such behaviors and other variables such as grades. We
-plan to develop a MOOC log analysis system based on the proposed model to
-both facilitate education research and help instructors improve course
+There are a number of interesting directions to further extend our work in the future. 
+First, the proposed model is 
+completely general and can thus be easily applied to analyze the log of many
+other courses to enable deep understanding of student behaviors as well as
+the correlations of such behaviors and other variables such as grades. To realize these benefits, 
+it would be useful 
+to develop a MOOC log analysis system based on the proposed model so as to facilitate education research and help instructors improve course
 design.
 
-Our model can empower many new comparative analyses. For example, we could
+Second, our model can empower many new comparative analyses. For example, we could
 now look at how behavior patterns change between different offerings of
@@
 
-Our model does not explicitly model drop-out like
-\citet{Kizilcec:2013:LAK}, but doing so is an obvious extension. Our model
-should be able to provide predictions of when a student is ``at risk'' for
+Third, there is more recent work on better learning algorithms for mixtures of
+Markov models~\citep{Gupta:2016:NIPS}. It would be worth exploring whether
+the advances proposed in this and similar work can be applied to our model
+to address some of the concerns surrounding our use of the EM algorithm for
+our parameter estimation.
+
+Finally, the proposed model can be extended in several ways. For example, 
+although our model does not explicitly model drop-out like
+\citet{Kizilcec:2013:LAK}, doing so is an obvious extension. Our model
+would be able to provide predictions of when a student is ``at risk'' for
 dropping out under such an extension.
-
-Currently, the model learns a transition matrix over the latent states that
+Also, currently, the model learns a transition matrix over the latent states that
 is \emph{shared} across all students. It would be interesting to instead
@@
 transitioning behavior between the higher-level behaviors we can identify.
-
-There is more recent work on better learning algorithms for mixtures of
-Markov models~\citep{Gupta:2016:NIPS}. It would be worth exploring whether
-the advances proposed in this and similar work can be applied to our model
-to address some of the concerns surrounding our use of the EM algorithm for
-our parameter estimation.

### 2017-06-20 sec-model.tex
Minor revision to related work, the parameter estimation section, conclusions, and acknowledgments.

@@
 particular, we propose a minor modification of the Baum-Welch
-algorithm~\cite{Rabiner:1990:RSR} which is used to learn the parameters for
-hidden Markov models in an unsupervised setting. In the following sections,
+algorithm~\cite{Rabiner:1990:RSR} which is an efficient EM algorithm
+for learning the parameters for
+hidden Markov models in an unsupervised setting where the Markovian assumption is exploited
+ to significantly reduce the computational complexity of the EM algorithm by avoiding explicit enumeration of all the possible state transitions.  In the following sections,
 we will provide a brief description of the original Baum-Welch algorithm

### 2017-06-20 sec-related-work.tex
Minor revision to related work, the parameter estimation section, conclusions, and acknowledgments.

@@
 
-Our formulation differs from both in that we do not feed the labeled
+Our formulation differs from the work~\cite{Zhang:2004:CVPR,Oliver:2004:CVIU} in that we do not feed the labeled
 sequence of the lower level into the input of the higher level. Instead,

### 2017-06-20 sec-related-work.tex
Minor: change one citation format in related work.

--- 2017-06-18 sec-related-work.tex
@@
 
-Our formulation differs from both in that we do not feed the labeled
-sequence of the lower level into the input of the higher level. Instead,
-our lower level is actually treated using a non-hidden Markov model, and
-the higher level is modeling transitions between the $K$ different
-non-hidden Markov models we consider. The problem to be solved is similar
-in that we wish to predict a ``label'' for a sequence of actions a student
-takes as well as understand the transition behavior between those labels.
-However, one of the consequences of modeling the lower layer using a
-non-hidden Markov model instead of an HMM directly is that the meaning of
-the $K$ different latent states can be more clearly captured by using our
-proposed behavior representation. If we were to use an HMM as our first
-layer, the behavior patterns (like from
-Figure~\ref{fig:motivating-example}) would instead have nodes that
-represent another set of latent states instead of being concrete actions
-themselves. Each of these latent states would then be associated with some
-other output distribution over the possible concrete actions to be
-considered. In order to understand a single behavior pattern uncovered, one
-would first have to understand the different output distributions for the
-latent states in that pattern in order to understand the meaning of that
-latent state. This further complicates the understanding of higher level
-patterns because understanding the higher layer patterns requires
+Our formulation differs from both \citet{Zhang:2004:CVPR} and
+\citet{Oliver:2004:CVIU} in that we do not feed the labeled sequence of the
+lower level into the input of the higher level. Instead, our lower level is
+actually treated using a non-hidden Markov model, and the higher level is
+modeling transitions between the $K$ different non-hidden Markov models we
+consider. The problem to be solved is similar in that we wish to predict a
+``label'' for a sequence of actions a student takes as well as understand
+the transition behavior between those labels.  However, one of the
+consequences of modeling the lower layer using a non-hidden Markov model
+instead of an HMM directly is that the meaning of the $K$ different latent
+states can be more clearly captured by using our proposed behavior
+representation. If we were to use an HMM as our first layer, the behavior
+patterns (like from Figure~\ref{fig:motivating-example}) would instead have
+nodes that represent another set of latent states instead of being concrete
+actions themselves. Each of these latent states would then be associated
+with some other output distribution over the possible concrete actions to
+be considered. In order to understand a single behavior pattern uncovered,
+one would first have to understand the different output distributions for
+the latent states in that pattern in order to understand the meaning of
+that latent state. This further complicates the understanding of higher
+level patterns because understanding the higher layer patterns requires
 understanding the lower layer patterns. By taking a more restrictive view

### 2017-06-20 presentation/presentation.tex
Add initial draft of presentation for EDM 2017.

\documentclass[10pt]{beamer}
\usetheme{metropolis}
\usepackage{tikz}
\usetikzlibrary{fit,positioning}
\usepackage{attrib}
\usepackage{rotating}
\usepackage{biblatex}
\usepackage{subcaption}
\addbibresource{../bib.bib}

\setbeamerfont{footnote}{size=\tiny}

\usepackage{amsmath}
\DeclareMathOperator*{\argmax}{arg\,max}

\setsansfont{Noto Sans}
\setmonofont{Droid Sans Mono}

\title{Modeling MOOC Student Behavior With Two-Layer Hidden Markov Models}
\author{Chase Geigle and ChengXiang Zhai}
\institute{Department of Computer Science, University of Illinois at
Urbana-Champaign}
\date{2017-06-25}

\begin{document}

\maketitle

\section{Motivation}

\begin{frame}
  \begin{center}
      \includegraphics[width=\textwidth]{figures/Growth-of-MOOCs.png}

      \small{\url{https://www.class-central.com/report/moocs-stats-and-trends-2016/}}
  \end{center}
\end{frame}

\begin{frame}{The Back of My Envelope}
  If every course looked like Cheng's\ldots
  \begin{center}
      \[
        \sim 7,000\text{ courses } \times \frac{\sim 85,000\text{ browsing
        sessions }}{\text{ course }} \times \frac{\sim 7\text{
        actions }}{\text{ browsing session }}
      \]
      \Large
      \begin{align*}
        \approx &\text{ } \mathbf{4.2} \text{ \textbf{billion} actions }
        \\
        &\text{ across } \mathbf{595} \text{ \textbf{million} sessions }
      \end{align*}
  \end{center}
\end{frame}

\begin{frame}{Where we come in\ldots}
    We need \textbf{\alert{concise}} and \textbf{\alert{digestible}}
    summaries of these logs in order to understand them!
    \\[\baselineskip]
    This paper contributes the following:
    \begin{enumerate}
      \item A suggestion for a \textbf{behavior pattern representation}, and
      \item A model that can efficiently and \textbf{automatically} capture
          this kind of representation in an \textbf{unsupervised} way
    \end{enumerate}
\end{frame}

\begin{frame}{Why do we need a different model/representation?}
    Rule-based
    methods\footfullcite{Kizilcec:2013:LAK}\textsuperscript{,}\footfullcite{Davis:2016:EDM}
    require \textbf{knowing what you're looking for} before you look for
    it.
    \\[\baselineskip]
    We suggest instead that patterns are \textbf{latent} and may be
    difficult to specify in advance via a series of rules.
    \\[\baselineskip]
    Let the data speak: \textbf{probabilistically} discover these patterns
    from large data.
\end{frame}

\begin{frame}{Why do we need a different model/representation?}
  Previous probabilistic representations\footfullcite{Faucon:2016:EDM}
  assume that students exhibit \textbf{only one behavioral pattern over
  time}.
  \\[\baselineskip]
  We suggest that student behavior is \textbf{dynamic}, and that we should
  have a model to understand how students move between the patterns we
  discover.
  \\[\baselineskip]
  Let the data speak: \textbf{probabilistically} discover \textbf{latent
  state transition} behavior
\end{frame}

\begin{frame}{Why do we need a different model/representation?}
  Previous models \textbf{assume} a certain level of \textbf{time
  granularity}\footfullcite{Shih:2010:EDM}\textsuperscript{,}\footfullcite{Kizilcec:2013:LAK}\textsuperscript{,}\footfullcite{Faucon:2016:EDM}
  for the patterns to be discovered.
  \\[\baselineskip]
  We propose a model that is \textbf{agnostic to the time resolution
  considered} to allow it to be flexible to model different levels of
  granularity as needed.
\end{frame}

\section{Example}

\begin{frame}{Our Behavior Representation}
  Visualize behavior patterns with \textbf{labeled directed graphs} where
  \begin{itemize}
      \item \textbf{node size} is proportional to the \textbf{probability
          of being at that node} at any given time,
          \\[\baselineskip]
      \item a \textbf{directed edge} represents a \textbf{transition from one behavior to
          another}, and
          \\[\baselineskip]
      \item the \textbf{thickness of an edge} is proportional to the
          \textbf{probability of taking that edge} to leave the source node.
  \end{itemize}
\end{frame}

\begin{frame}{(Theoretical Example) Behavior Pattern 0}
  \centering
  \includegraphics[width=0.7\textwidth]{../figures/example/state0.png}
\end{frame}

\begin{frame}{(Theoretical Example) Behavior Pattern 1}
  \centering
  \includegraphics[width=0.7\textwidth]{../figures/example/state1.png}
\end{frame}

\begin{frame}{(Theoretical Example) Transitions Between Patterns}
  \centering
  \includegraphics[width=0.7\textwidth]{../figures/example/trans.png}
\end{frame}

\section{Modeling Details}

\begin{frame}{The Model}
  Notation:
  \begin{itemize}
    \item $\mathbf{A}$: a set of actions students take in the MOOC log,
    \item $\mathbf{L}$: the set of all students in the MOOC,
    \item $\mathbf{O}^{(\ell)}$: the list of action sequences (list of
        lists) for each student $\ell \in \mathbf{L}$
    \item $\mathbf{o} \in \mathbf{O}^{(\ell)}$: a list of actions a student
        took in one ``session'' $\mathbf{o} = (a_1, a_2, \ldots, a_T)$
        where $a_i \in \mathbf{A}$.
  \end{itemize}
\end{frame}

\begin{frame}{The Model}
  Each pattern is associated with a latent state $u \in \{1,\ldots,K\}$,
  which is itself a Markov model represented with $\lambda^{(u)} =
  (\pi^{(u)}, A^{(u)})$ where
  \begin{itemize}
    \item $\pi^{(u)}$ is a $|\mathbf{A}|$ length probability vector to
        indicate the probability of starting with any action $a \in
        \mathbf{A}$, and
    \item $A^{(u)}$ is a $|\mathbf{A}|\times|\mathbf{A}|$ matrix indicating
        the transition probabilities between each pair of actions in
        $\mathbf{A}$.
  \end{itemize}

  \textbf{Assumption}: Each action list $\mathbf{o} \in
  \mathbf{O}^{(\ell)}$ is generated from one specific latent state.
  If $\mathbf{o} = (a_1, \ldots, a_T)$, then
  \[
      P(\mathbf{o} \mid \lambda^{(u)})
      = P(a_1 \mid \pi^{(u)}) \prod_{i=2}^T P(a_i \mid a_{i-1}, A^{(u)})
  \]
\end{frame}

\begin{frame}{The Model}
  Finally, we can compute the likelihood of a list of action sequences
  $\mathbf{O}^{(\ell)}$ of length $N$ for any student $\ell \in \mathbf{L}$
  as
  \begin{align*}
      P(\mathbf{O}^{(\ell)} \mid \Lambda) = \sum_{(v_1,\ldots,v_N) \in U}
          &\left(P(v_1 \mid \Lambda) P(\mathbf{o}_1 \mid
          \lambda^{(v_1)})\phantom{\prod_{i}^N}\right.\\
          & \left.\times
          \prod_{i=2}^N P(v_i \mid v_{i-1}, \Lambda)P(\mathbf{o}_i \mid
      \lambda^{(v_i)})\right)
  \end{align*}
  where $\Lambda = (\pi, A, B)$ is the tuple of all model parameters.
  \begin{itemize}
    \item $\pi$: distribution over initial latent states,
    \item $A$: $K\times K$ transition matrix for each pair of latent
        states,
    \item $B = (\lambda^{(1)}, \ldots, \lambda^{(K)})$, the latent state
        Markov model representations
  \end{itemize}
\end{frame}

\begin{frame}{Inference}
    \small
    Given $\mathbf{D} = (\mathbf{O}^{(1)}, \ldots,
    \mathbf{O}^{(|\mathbf{L}|)})$, our goal is to infer the $\Lambda = (\pi, A,
    B)$ that maximizes the likelihood our model generates $\mathbf{D}$.
    \\[\baselineskip]
    Modified Baum-Welch algorithm\footfullcite{Rabiner:1990:RSR} for hidden
    Markov models. M-step updated to learn the latent state
    representations\footnote{See the full paper for more details.}:
    \begin{align*}
      \pi^{(i)}_{a} &= \frac{\sum_{\mathbf{O} \in \mathbf{D}}
      \sum_{\mathbf{o}_t \in \mathbf{O},\mathbf{o}_{t,1} = a}
      \gamma^{(\mathbf{O})}_t(i)} {\sum_{\mathbf{O} \in \mathbf{D}}
      \sum_{\mathbf{o}_t \in \mathbf{O}} \gamma^{(\mathbf{O})}_t(i)},
      \text{ and }\\
      A^{(i)}_{ab} &= \frac{\sum_{\mathbf{O} \in \mathbf{D}}
      \sum_{\mathbf{o}_t \in \mathbf{O}} \sum_{m=2,\mathbf{o}_{t,m-1}=a
      \land \mathbf{o}_{t,m} = b}^{|\mathbf{o}_{t}|}
      \gamma^{(\mathbf{O})}_t(i)} {\sum_{\mathbf{O} \in \mathbf{D}}
      \sum_{\mathbf{o}_t \in \mathbf{O}}
      \sum_{m=2,\mathbf{o}_{t,m-1}=a}^{|\mathbf{o}_{t}|}
      \gamma^{(\mathbf{O})}_t(i)}
    \end{align*}
    where $\gamma^{(\mathbf{O})}(i)$ is the posterior probability that a
    list of action sequences $\mathbf{O}$ is generated from latent state
    $i$ (efficiently recovered via Baum-Welch).
\end{frame}

\section{Experiments}

\newcommand{\textretrieval}{textretrieval-001}
\newcommand{\sustain}{sustain-001}
\newcommand{\UIUC}{UIUC}

\begin{frame}{Data}
\begin{table}
  \begin{center}
    \caption{Statistics about the sequences extracted from the two MOOCs.}
    \label{table:datasets}
    \begin{tabular}{rrrr}
      \textbf{MOOC} & \textbf{Students} & \textbf{Sequences} & \textbf{Avg.
      $|\mathbf{s}|$}\\\hline
      \textretrieval{} & 18,941 & 85,240 & 7.31\\
      \sustain{} & 85,240 & 231,881 & 15.4
    \end{tabular}
  \end{center}
\end{table}
\end{frame}

\begin{frame}{Behavior Pattern 0 ($K = 4$)}
  \centering
  \includegraphics[width=0.70\textwidth]{../figures/text-4state/state0.png}
\end{frame}

\begin{frame}{Behavior Pattern 1 ($K = 4$)}
  \centering
  \includegraphics[width=0.70\textwidth]{../figures/text-4state/state1.png}
\end{frame}

\begin{frame}{Behavior Pattern 2 ($K = 4$)}
  \centering
  \includegraphics[width=0.70\textwidth]{../figures/text-4state/state2.png}
\end{frame}

\begin{frame}{Behavior Pattern 3 ($K = 4$)}
  \centering
  \includegraphics[width=0.70\textwidth]{../figures/text-4state/state3.png}
\end{frame}

\begin{frame}{A Closer Look}
 State 1 and state 3 capture most of the quiz taking behavior, but are
 actually quite different from each other. Let's look closer\ldots
\end{frame}

\begin{frame}{``For more on this, it's time for \emph{A Closer Look}.''
    (State 1)}
  \centering
  \includegraphics[width=0.70\textwidth]{../figures/text-4state/state1.png}
\end{frame}

\begin{frame}{``For more on this, it's time for \emph{A Closer Look}.''
    (State 3)}
  \centering
  \includegraphics[width=0.70\textwidth]{../figures/text-4state/state3.png}
\end{frame}

\begin{frame}{A Look at Transitions (All)}
  \centering
  \includegraphics[width=0.70\textwidth]{../figures/trans-comp/trans-avg.png}
\end{frame}

\begin{frame}{A Look at Transitions (Perfect retrofit)}
  \centering
  \includegraphics[width=0.70\textwidth]{../figures/trans-comp/trans-perfect.png}
\end{frame}

\begin{frame}{A Look at Transitions (Low retrofit)}
  \centering
  \includegraphics[width=0.70\textwidth]{../figures/trans-comp/trans-low.png}
\end{frame}

\begin{frame}{A Look at Transitions (All)}
  \centering
  \includegraphics[width=0.70\textwidth]{../figures/trans-comp/trans-avg.png}
\end{frame}

\begin{frame}{Ranking experiment}
\begin{table}
  \centering
  \caption{Average rank for ``perfect'' and ``low'' student groups in the
  ranked lists associated with the four latent states found by a 2L-HMM.
  $\dagger$ indicates statistically significant different mean ranks at $p
  < 0.01$ according to an unpaired $t$-test.}
  \label{table:mean-rank}
  \begin{tabular}{r|rrrrr}
    \textbf{Group} & \textbf{State 0} & \textbf{State
  1}\textsuperscript\textdagger &
    \textbf{State 2}\textsuperscript\textdagger & \textbf{State 3} & \textbf{State 2
    $\rightarrow$ 2}\textsuperscript\textdagger\\\hline
    \textbf{Perfect} & 975.3  & 1001.5 & 999.0  & 1056.5 & 939.6\\
    \textbf{Low}     & 1024.9 & 816.4  & 1230.5 & 1161.2 & 1187.4\\
  \end{tabular}
\end{table}
\end{frame}

\section{Concluding Remarks}

\begin{frame}{Conclusion}
\small
We proposed using \textbf{\alert{Two-layer Hidden Markov Models}} to
discover \textbf{latent student behavior patterns} as well as the
\textbf{transitioning behavior between them}. The patterns we extract are
meaningful and can be used to extract features that correlate with student
performance.
\\[\baselineskip]
The code for learning the parameters of a 2L-HMM is available as part of
the \textsc{MeTA} toolkit\footfullcite{Massung:2016:ACL} available at
\url{https://meta-toolkit.org}.
\\[\baselineskip]
The code for generating the visualizations for a 2L-HMM on MOOC data is
available on GitHub: \url{https://github.com/skystrife/clickstream-hmm}
\end{frame}

\begin{frame}
    \centering
    \Huge
    Thank you!
\end{frame}

\end{document}

### 2017-06-22 presentation/presentation.tex
Improve visualization for EDM 2017 presentation.

@@
 \begin{frame}{The Back of My Envelope}
+  Data looks like:
+  \begin{center}
+    view lecture \rightarrow\, quiz \rightarrow\, forum \rightarrow\,
+    \ldots\, \rightarrow\, view lecture \rightarrow\, quiz \rightarrow\,
+    quiz \rightarrow\, forum \rightarrow\, \ldots
+  \end{center}
+
   If every course looked like Cheng's\ldots
@@
     \begin{enumerate}
-      \item A suggestion for a \textbf{behavior pattern representation}, and
+      \item A suggestion for a \textbf{behavior pattern representation} to
+       understand patterns and \textbf{transitions between them}, and
+       \\[0.5\baselineskip]
       \item A model that can efficiently and \textbf{automatically} capture
@@
     \end{enumerate}
+\end{frame}
+
+\section{Example}
+
+\begin{frame}{Our Behavior Representation}
+  Visualize behavior patterns with \textbf{labeled directed graphs} where
+  \begin{itemize}
+      \item \textbf{node size} is proportional to the \textbf{probability
+          of being at that node} at any given time,
+          \\[\baselineskip]
+      \item a \textbf{directed edge} represents a \textbf{transition from one behavior to
+          another}, and
+          \\[\baselineskip]
+      \item the \textbf{thickness of an edge} is proportional to the
+          \textbf{probability of taking that edge} to leave the source node.
+  \end{itemize}
+\end{frame}
+
+\begin{frame}{(Theoretical Example) Behavior Pattern 0}
+  \centering
+  \includegraphics[width=0.7\textwidth]{../figures/example/state0.png}
+\end{frame}
+
+\begin{frame}{(Theoretical Example) Behavior Pattern 1}
+  \centering
+  \includegraphics[width=0.7\textwidth]{../figures/example/state1.png}
+\end{frame}
+
+\begin{frame}{(Theoretical Example) Transitions Between Patterns}
+  \centering
+  \includegraphics[width=0.7\textwidth]{../figures/example/trans.png}
 \end{frame}
@@
 
-\section{Example}
-
-\begin{frame}{Our Behavior Representation}
-  Visualize behavior patterns with \textbf{labeled directed graphs} where
-  \begin{itemize}
-      \item \textbf{node size} is proportional to the \textbf{probability
-          of being at that node} at any given time,
-          \\[\baselineskip]
-      \item a \textbf{directed edge} represents a \textbf{transition from one behavior to
-          another}, and
-          \\[\baselineskip]
-      \item the \textbf{thickness of an edge} is proportional to the
-          \textbf{probability of taking that edge} to leave the source node.
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{(Theoretical Example) Behavior Pattern 0}
-  \centering
-  \includegraphics[width=0.7\textwidth]{../figures/example/state0.png}
-\end{frame}
-
-\begin{frame}{(Theoretical Example) Behavior Pattern 1}
-  \centering
-  \includegraphics[width=0.7\textwidth]{../figures/example/state1.png}
-\end{frame}
-
-\begin{frame}{(Theoretical Example) Transitions Between Patterns}
-  \centering
-  \includegraphics[width=0.7\textwidth]{../figures/example/trans.png}
-\end{frame}
-
 \section{Modeling Details}
 
-\begin{frame}{The Model}
-  Notation:
-  \begin{itemize}
-    \item $\mathbf{A}$: a set of actions students take in the MOOC log,
-    \item $\mathbf{L}$: the set of all students in the MOOC,
-    \item $\mathbf{O}^{(\ell)}$: the list of action sequences (list of
-        lists) for each student $\ell \in \mathbf{L}$
-    \item $\mathbf{o} \in \mathbf{O}^{(\ell)}$: a list of actions a student
-        took in one ``session'' $\mathbf{o} = (a_1, a_2, \ldots, a_T)$
-        where $a_i \in \mathbf{A}$.
-  \end{itemize}
-\end{frame}
+\begin{frame}{The Generative Model}
+  \begin{center}
+    \begin{tikzpicture}
+      \tikzstyle{main}=[circle, minimum size = 10mm, thick, draw =black!80, node distance = 10mm]
+      \tikzstyle{observed}=[circle, minimum size = 10mm, thick, draw=black!80, node distance = 10mm, fill = black!10]
+      \tikzstyle{invismain}=[circle, minimum size = 10mm, thick, draw =white!100, node distance = 10mm]
+      \tikzstyle{connect}=[-latex, thick]
+      \tikzstyle{box}=[rectangle, draw=black!100]
 
-\begin{frame}{The Model}
-  Each pattern is associated with a latent state $u \in \{1,\ldots,K\}$,
-  which is itself a Markov model represented with $\lambda^{(u)} =
-  (\pi^{(u)}, A^{(u)})$ where
-  \begin{itemize}
-    \item $\pi^{(u)}$ is a $|\mathbf{A}|$ length probability vector to
-        indicate the probability of starting with any action $a \in
-        \mathbf{A}$, and
-    \item $A^{(u)}$ is a $|\mathbf{A}|\times|\mathbf{A}|$ matrix indicating
-        the transition probabilities between each pair of actions in
-        $\mathbf{A}$.
-  \end{itemize}
+      % tags
+      \node[main, fill = white!100] (s1) [] { $s_1$ };
+      \node[main, fill = white!100] (s2) [right=of s1] { $s_2$ };
+      \node[main, fill = white!100] (s3) [right=of s2] { $s_3$ };
+      \node[main, fill = white!100] (s4) [right=of s3] { $s_4$ };
+      \node[invismain, opacity=0, text opacity=1] (sn) [right=of s4] {$\ldots$};
 
-  \textbf{Assumption}: Each action list $\mathbf{o} \in
-  \mathbf{O}^{(\ell)}$ is generated from one specific latent state.
-  If $\mathbf{o} = (a_1, \ldots, a_T)$, then
+      % words
+      \node[observed] (o1) [below=of s1] { $\mathbf{o_1}$ };
+      \node[observed] (o2) [below=of s2] { $\mathbf{o_2}$ };
+      \node[observed] (o3) [below=of s3] { $\mathbf{o_3}$ };
+      \node[observed] (o4) [below=of s4] { $\mathbf{o_4}$ };
+      \node[invismain, opacity=0, text opacity=1] (on) [below=of sn] {$\ldots$};
+
+      \path (s1) edge [connect] (s2)
+        (s2) edge [connect] (s3)
+        (s3) edge [connect] (s4)
+        (s4) edge [connect] (sn);
+
+      \path (s1) edge [connect] (o1);
+      \path (s2) edge [connect] (o2);
+      \path (s3) edge [connect] (o3);
+      \path (s4) edge [connect] (o4);
+
+    \end{tikzpicture}
+  \end{center}
   \[
-      P(\mathbf{o} \mid \lambda^{(u)})
-      = P(a_1 \mid \pi^{(u)}) \prod_{i=2}^T P(a_i \mid a_{i-1}, A^{(u)})
+    \underbrace{\overbrace{\langle \overbrace{lecture}^{\text{action } a_1}
+      \rightarrow\, \overbrace{quiz}^{\text{action } a_2} \rightarrow\,
+      \ldots\, \rightarrow\, \overbrace{wiki}^{\text{action } a_T}
+    \rangle}^{\text{observation } \mathbf{o}_1 \text{ ``generated'' from }
+    s_1}, \overbrace{\langle \overbrace{wiki}^{\text{action } a_1}
+      \rightarrow\, \overbrace{quiz}^{\text{action } a_2} \rightarrow\,
+    \ldots \rangle}^{\text{observation } \mathbf{o}_2 \text{ ``generated'' from } s_2},
+    \ldots}_{\text{student sequence } \mathbf{O}^{(\ell)}}
   \]
@@
 
-\begin{frame}{The Model}
-  Finally, we can compute the likelihood of a list of action sequences
-  $\mathbf{O}^{(\ell)}$ of length $N$ for any student $\ell \in \mathbf{L}$
-  as
-  \begin{align*}
-      P(\mathbf{O}^{(\ell)} \mid \Lambda) = \sum_{(v_1,\ldots,v_N) \in U}
-          &\left(P(v_1 \mid \Lambda) P(\mathbf{o}_1 \mid
-          \lambda^{(v_1)})\phantom{\prod_{i}^N}\right.\\
-          & \left.\times
-          \prod_{i=2}^N P(v_i \mid v_{i-1}, \Lambda)P(\mathbf{o}_i \mid
-      \lambda^{(v_i)})\right)
-  \end{align*}
-  where $\Lambda = (\pi, A, B)$ is the tuple of all model parameters.
+\begin{frame}{The Generative Model}
+  \begin{center}
+    \begin{tikzpicture}
+      \tikzstyle{main}=[circle, minimum size = 10mm, thick, draw =black!80, node distance = 10mm]
+      \tikzstyle{observed}=[circle, minimum size = 10mm, thick, draw=black!80, node distance = 10mm, fill = black!10]
+      \tikzstyle{invismain}=[circle, minimum size = 10mm, thick, draw =white!100, node distance = 10mm]
+      \tikzstyle{connect}=[-latex, thick]
+      \tikzstyle{box}=[rectangle, draw=black!100]
+
+      % tags
+      \node[main, fill = white!100] (s1) [] { $s_1$ };
+      \node[main, fill = white!100] (s2) [right=of s1] { $s_2$ };
+      \node[main, fill = white!100] (s3) [right=of s2] { $s_3$ };
+      \node[main, fill = white!100] (s4) [right=of s3] { $s_4$ };
+      \node[invismain, opacity=0, text opacity=1] (sn) [right=of s4] {$\ldots$};
+
+      % words
+      \node[observed] (o1) [below=of s1] { $\mathbf{o_1}$ };
+      \node[observed] (o2) [below=of s2] { $\mathbf{o_2}$ };
+      \node[observed] (o3) [below=of s3] { $\mathbf{o_3}$ };
+      \node[observed] (o4) [below=of s4] { $\mathbf{o_4}$ };
+      \node[invismain, opacity=0, text opacity=1] (on) [below=of sn] {$\ldots$};
+
+      \path (s1) edge [connect] (s2)
+        (s2) edge [connect] (s3)
+        (s3) edge [connect] (s4)
+        (s4) edge [connect] (sn);
+
+      \path (s1) edge [connect] (o1);
+      \path (s2) edge [connect] (o2);
+      \path (s3) edge [connect] (o3);
+      \path (s4) edge [connect] (o4);
+
+    \end{tikzpicture}
+  \end{center}
+
+  Model parameters $\Lambda = (\pi, A, B)$:
   \begin{itemize}
-    \item $\pi$: distribution over initial latent states,
-    \item $A$: $K\times K$ transition matrix for each pair of latent
-        states,
-    \item $B = (\lambda^{(1)}, \ldots, \lambda^{(K)})$, the latent state
-        Markov model representations
+    \item $\pi$: probability of starting student sequence in latent state
+      $s_i$
+    \item $A$: transition probability between latent states
+    \item $B = (\lambda^{(1)}, \ldots, \lambda^{(K)})$ where
+      \begin{itemize}
+        \item $\lambda^{(i)} = (\pi^{(i)}, A^{(i)})$, initial probability
+          and transition matrix between student \emph{actions} (e.g. view
+          forum, take quiz, \ldots) for latent state $s_i$
+      \end{itemize}
   \end{itemize}

### 2017-07-25 sec-conclusion.tex
Grammar fixes from Grammarly analysis.

@@
 % results; argue again for our novelty; discuss future directions
-As a tool to help instructors and eduation researchers better understand behavior of MOOC students, 
- we proposed a two-layer hidden Markov model to automatically extract 
-student activity patterns in the form of behavior state-transition graphs from large amounts of 
-MOOC log data. This model is different from existing
-methods in that it treats behavior patterns as a sequence of \emph{latent
-states}, rather than assigning these states in a rule-based manner. It
-captures the variable behaviors of students over time, and allows
-analysis at different levels of granularity.
+As a tool to help instructors and education researchers better understand
+the behavior of MOOC students, we proposed a two-layer hidden Markov model
+to automatically extract student activity patterns in the form of behavior
+state-transition graphs from large amounts of MOOC log data. This model is
+different from existing methods in that it treats behavior patterns as a
+sequence of \emph{latent states}, rather than assigning these states in a
+rule-based manner. It captures the variable behaviors of students over time
+and allows analysis at different levels of granularity.
 
-We showed that such a model does in fact capture meaningful behavior
+We showed that such a model does, in fact, capture meaningful behavior
 patterns and produces descriptions of these behavior patterns that are easy
@@
 These results show the great potential of the proposed model for serving as a tool
-to help humans discovery knowledge about student behaviors.
+to help humans discover knowledge about student behaviors.
 
@@
 the same MOOC to understand how changes in course structure or materials
-influences student behavior. Individual students can now also be compared
+influence student behavior. Individual students can now also be compared
 against each other or against groups. For example, by decoding the latent

### 2017-07-25 sec-intro.tex
Grammar fixes from Grammarly analysis.

@@
 the vast scale of these behavioral logs, student behavior patterns are
-mostly undetectable for MOOC instructors and as a result the rich data
+mostly undetectable for MOOC instructors and as a result, the rich data
 available through MOOC logs is highly underutilized today.
@@
 understand changes in student behavior with respect to these patterns?
-These are still very open questions, and are active areas of
+These are still very open questions and are active areas of
 research~\cite{Kizilcec:2013:LAK, Faucon:2016:EDM, Davis:2016:EDM,
@@
       student behavior dynamics in a course. We propose a latent space
-      model with {\em latent state transitions} to flexibly model the
+      model with \emph{latent state transitions} to flexibly model the
       dynamics.
@@
 
-Thus, what we propose is a \emph{latent variable approach} to mining student behavior
+Thus, what we propose is a \emph{latent variable approach} for mining student behavior
 patterns that is \emph{probabilistic} for inference and

### 2017-07-25 sec-limitations.tex
Grammar fixes from Grammarly analysis.

@@
 our model parameters. Second, there are limitations in the kinds of
-patterns our model is able to discover in its current formulation, and the
+patterns our model is able to discover in its current formulation and the
 ease with which instructors are able to extract knowledge from these
@@
 optimizing in a highly non-convex parameter space. Thus, it can only
-guarantee reaching a local maximum in practice~\citep{Dempster:1977:JRSS}.
-This may mean that the parameters found for the model may not be the
-``best'' parameters in a global sense, which may lead to suboptimal latent
-state representations and transition patterns. Empirically, however, we
-observed in our experiments that strong patterns tend to always show up as
-long as the algorithm reaches a reasonably good local maximum, and the
-differences of the results tend to be related to weak ``unstable'' patterns
-which may not be reliable anyway. Since it is far more important and useful
-to reveal the strong salient patterns than weak unreliable ones for
-instructors, the problem might not necessarily affect the utility of the
-approach so significantly.  A commonly applied approach to address the
-problem of multiple local maxima is to run the model multiple times with
-different starting points to allow the model to explore a larger portion of
-the potential parameter space. One can then compare the log-likelihood of
-the data between the models that were started from different initialization
-points and select the one that has the highest value. This still does not
-guarantee finding a global maximum, but it does help alleviate the
-potential for finding a particularly bad local maximum.  In practice, we
-have found that our model can be fit to the data fairly quickly on
-commodity hardware, so running it multiple times to address this concern is
-not computationally unfeasible.
+guarantee that we reach a local maximum in
+practice~\citep{Dempster:1977:JRSS}.  This may mean that the parameters
+found for the model may not be the ``best'' parameters in a global sense,
+which may lead to suboptimal latent state representations and transition
+patterns. Empirically, however, we observed in our experiments that strong
+patterns tend to always show up as long as the algorithm reaches a
+reasonably good local maximum, and the differences of the results tend to
+be related to weak ``unstable'' patterns which may not be reliable anyway.
+Since it is far more important and useful to reveal the strong salient
+patterns than weak unreliable ones for instructors, the problem might not
+necessarily affect the utility of the approach so significantly.  A
+commonly applied approach to address the problem of multiple local maxima
+is to run the model multiple times with different starting points to allow
+the model to explore a larger portion of the potential parameter space. One
+can then compare the log-likelihood of the data between the models that
+were started from different initialization points and select the one that
+has the highest value. This still does not guarantee that we find a global
+maximum, but it does help alleviate the potential for finding a
+particularly bad local maximum.  In practice, we have found that our model
+can be fit to the data fairly quickly on commodity hardware, so running it
+multiple times to address this concern is not computationally unfeasible.
 
@@
 changing the segmentation strategy one uses to divide the user action
-stream into discrete ``sessions.'' The other is in the number of latent
+stream into discrete ``sessions.'' The other is the number of latent
 states $K$ that are used to describe the segmented action sequences. One

### 2017-07-25 sec-model.tex
Minor notation fixes in model section.
Grammar fixes from Grammarly analysis.

--- 2017-06-14 sec-model.tex
@@
 likelihood estimation of the model in the presence of latent (unobserved)
-variables. This algorithm, intuitively, works as follows: first the model
+variables. This algorithm, intuitively, works as follows: first, the model
 parameters we wish to estimate are initialized to some random (but valid)
@@
 each $a_i \in \mathbf{A}$.  Each sequence can be divided flexibly; in this
-paper we chose to denote the end of a sequence as occurring when no further
-actions occur within a 10 hour window of time (and thus our sequences
+paper, we chose to denote the end of a sequence as occurring when no further
+actions occur within a 10-hour window of time (and thus our sequences
 roughly correspond to one day's worth of activity). This decision reflects
@@
 of day-to-day behavior. A different segmentation strategy could be used to
-uncover hour-by-hour behavior, or week-by-week behavior, and this depends
+uncover hour-by-hour behavior or week-by-week behavior, and this depends
 entirely on the desired time resolution one wishes to be exhibited in the
@@
 
-Thus, the probability of a sequence $\mathbf{o}$ of length $M$ given its
+Thus, the probability of a sequence $\mathbf{o}$ of length $T$ given a
 latent state $u$ is
@@
 particular, we propose a minor modification of the Baum-Welch
-algorithm~\cite{Rabiner:1990:RSR} which is used to learn the parameters for
-hidden Markov models in an unsupervised setting. In the following sections,
+algorithm~\cite{Rabiner:1990:RSR} which is an efficient EM algorithm
+for learning the parameters of
+hidden Markov models in an unsupervised setting where the Markovian assumption is exploited
+ to significantly reduce the computational complexity of the EM algorithm by avoiding explicit enumeration of all the possible state transitions.  In the following sections,
 we will provide a brief description of the original Baum-Welch algorithm
@@
 
-The goal of the Baum-Welch algorithm is still to learn the paramters
+The goal of the Baum-Welch algorithm is still to learn the parameters
 $\Lambda$ for our model. What differs from the categorical-valued HMM case
@@
 Formally, our training data $\mathbf{D} = (\mathbf{O}^{(1)}, \ldots,
-\mathbf{O}^{(|L|)})$, where each $\mathbf{O}^{(\ell)} = (\mathbf{o}_1,
+\mathbf{O}^{(|\mathbf{L}|)})$, where each $\mathbf{O}^{(\ell)} = (\mathbf{o}_1,
 \ldots, \mathbf{o}_{T_\ell})$ is itself a list of sequences. Each sequence

### 2017-07-25 sec-related-work.tex
Grammar fixes from Grammarly analysis.

@@
 
-Our model is based heavily upon the prior art of Hidden Markov
+Our model is based heavily on the prior art of Hidden Markov
 Models~\cite{Rabiner:1990:RSR} for sequence labeling tasks. These types of
@@
 vocabulary) and the latent states represent the part-of-speech category for
-a word. In speech recognition~\cite{Huang:1990:HMM} the output
+a word. In speech recognition~\cite{Huang:1990:HMM}, the output
 distributions might be mixtures of Gaussians to predict real-valued vectors
@@
 and outputs prediction results as to which of the $K$ HMMs at layer two
-were mostly likely to produce the sequence of outputs produced by the
+were most likely to produce the sequence of outputs produced by the
 previous layer, and so on.
@@
 understanding the lower layer patterns. By taking a more restrictive view
-on the first layer, we are able to produce a representation that can be
+of the first layer, we are able to produce a representation that can be
 more readily interpreted due to the states in our first layer
@@
 layer lie the ``production'' states, which output symbols according to some
-probability distribution. Our specific model in this case can be modeled as
-a HHMM where the horizontal transitions between nodes at the highest layer
+probability distribution. Our specific model, in this case, can be modeled as
+an HHMM where the horizontal transitions between nodes at the highest layer
 (including self-loops) \emph{must} be immediately followed by a vertical
@@
 previous models are not well tuned toward the student behavior task and
-thus cannot adequately address all the aspects of complexity of student
+thus cannot adequately address all the aspects of the complexity of student
 learning behavior.  A main technical contribution of this paper is to
@@
 thus enable discovery of more sophisticated behavior patterns and provide
-more detailed characterization of student behavior than the previous
+a more detailed characterization of student behavior than the previous
 models.
@@
 probabilistic framework to account for uncertainty in this state
-assignment, and jointly learn \emph{representations} for these states,
+assignment and jointly learn \emph{representations} for these states,
 which are treated as being \emph{latent} as opposed to pre-defined using

### 2017-07-25 sec-results.tex
Update paper to incorporate review A and review B.
Also update our response to these two reviewers to be more concrete
about the exact changes we have made to the paper and where to find
them.
Add an intuitive explanation of the EM algorithm.
Incorporate new citation from the editor.
Also shuffle around some figures to make the page layout prettier with
the now longer related work and model "derivation" sections.
Grammar fixes from Grammarly analysis.

@@
 \section{Experiment Results}
-As an analysis tool, the 2L-HMM model provides with us the following two
+As an analysis tool, the 2L-HMM model provides us with the following two
 patterns to characterize student behavior:
@@
 
-\begin{figure*}
+\begin{figure}
   \centering
@@
   \label{fig:states}
-\end{figure*}
-
-\begin{figure*}
-  \centering
-  \begin{subfigure}[t]{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figures/text-6state/state4.png}
-    \caption{A state from \textretrieval{}.}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figures/sustain-6state/state4.png}
-    \caption{A state from \sustain{}.}
-  \end{subfigure}
-  \caption{Two similar example states found by the 6-state 2L-HMM.}
-  \label{fig:cross-course}
-\end{figure*}
+\end{figure}
 
@@
 
-We also argue that it is important that the latent state representation be
+We also argue that it is important that the latent state representation is
 a Markov model rather than just a discrete distribution over actions in
@@
 
+\begin{figure}
+  \centering
+  \begin{subfigure}[t]{0.5\textwidth}
+    \includegraphics[width=\textwidth]{figures/text-6state/state4.png}
+    \caption{A state from \textretrieval{}.}
+  \end{subfigure}%
+  \begin{subfigure}[t]{0.5\textwidth}
+    \includegraphics[width=\textwidth]{figures/sustain-6state/state4.png}
+    \caption{A state from \sustain{}.}
+  \end{subfigure}
+  \caption{Two similar example states found by the 6-state 2L-HMM.}
+  \label{fig:cross-course}
+\end{figure}
+
 We can also use our model for cross-course behavior analysis. In
@@
 \sustain{}. These two states were chosen as they are the most similar
-across the two courses. However, if we look at the transitions we can see
+between the two courses. However, if we look at the transitions we can see
 some important differences. First, in the state from \textretrieval{}, we
@@
 as we increase $K$ when fitting our 2L-HMM on the \sustain{} MOOC.\ Again,
-in the transition between $K=2$ and $K=3$ we discover forum behavior
+in the transition between $K=2$ and $K=3$, we discover forum behavior
 patterns as a latent state. However, in the transition between $K=3$ to
-$K=4$ we instead see a refining of that discovered forum state, where
+$K=4$, we instead see a refining of that discovered forum state, where
 state 3 captures asymmetric transition probabilities between ``forum thread
@@
 latent states. In Figure~\ref{fig:trans-avg} we show the latent state
-transition diagram for a 3-state 2L-HMM fit on \textretrieval{}. We can
+transition diagram for a 4-state 2L-HMM fit on \textretrieval{}. We can
 immediately observe two things: \begin{enumerate*}[label=(\arabic*)]
@@
 relatively lower probability than the other states as we might expect. It
-also makes sense that state 0 has rather high probability as this state
+also makes sense that state 0 has a rather high probability as this state
 likely captures all sequences where a student logged in to the platform and
@@
   \label{fig:trans-comp}
-\end{figure*}
+\end{figure}
 
@@
 
-Returning to Figure~\ref{fig:trans-comp}, we an also see a difference in
+Returning to Figure~\ref{fig:trans-comp}, we can also see a difference in
 the transitions between the latent states. In particular, look at the
@@
 nearly all of this probability mass is allocated for the self-loop ($p
-\approx 0.93$). In the ``low'' group, this self loop is less strong ($p
+\approx 0.93$). In the ``low'' group, this self-loop is less strong ($p
 \approx 0.80$; most easily seen by noting that the edges leaving state 2

### 2017-08-24 sec-conclusion.tex
Proofread and revised the entire paper.

@@
 and posit that \emph{groups} of students transition between the lower layer
-patterns our method identifies in distinct ways and instead do a soft
+patterns identified by our method in distinct ways; this way, we can do a soft
 clustering of students into $K_2$ groups based on the similarity of their

### 2017-08-24 sec-intro.tex
Proofread and revised the entire paper.

@@
 unsupervised, they can potentially be applied to any MOOC data without
-requiring manual annotation effort at the level of sequences. Instead,
-instructors are empowered to use the latent patterns that the 2L-HMM can
-discover from raw data to extract knowledge about the behaviors his/her
+requiring manual annotation effort at the level of sequences, thus empowering
+% Instead,
+instructors 
+%are empowered 
+to use the latent patterns discovered by the 2L-HMM to further extract knowledge about the behaviors his/her
 students exhibit in the MOOC.

### 2017-08-24 sec-model.tex
Proofread and revised the entire paper.

@@
 student can take at any given time. For example, an action $a \in
-\mathbf{A}$ might be ``viewing lecture'', ``taking quiz'', or ``viewing
-forum''. For each student in the course $\ell \in \mathbf{L}$, we then
+\mathbf{A}$ might be ``viewing lecture,'' ``taking quiz,'' or ``viewing
+forum.'' For each student in the course $\ell \in \mathbf{L}$, we then
 extract a list of action sequences $\mathbf{O}_\ell$ that he or she
@@
 will result in different underlying training data, and thus different
-meanings behind the patterns that the 2L-HMM will extract.
+meanings behind the patterns that the 2L-HMM will extract. The flexibility of
+using different segmentation strategies is intentional as it allows a user to
+flexibly adjust the segmentation so as to obtain patterns at different granularity levels as needed.
 
@@
 \label{sec:param-estimation}
-To learn the parameters of our model, we may attempt to use maximum
+To learn the parameters of our model, we may use maximum
 likelihood estimation. Unfortunately, a closed-form solution does not

### 2017-08-24 sec-related-work.tex
Proofread and revised the entire paper.

@@
 Our model is based heavily on the prior art of Hidden Markov
-Models~\cite{Rabiner:1990:RSR} for sequence labeling tasks. These types of
-models are widely applicable and have been used for tasks such as speech
+Models (HMMs)~\cite{Rabiner:1990:RSR} for sequence labeling tasks.  As a member
+of the more general family of probabilistic graphical models~\cite{Koller:2009:PGM}, HMMs are widely applicable and have been used for tasks such as speech
 recognition~\cite{Huang:1990:HMM}, part-of-speech
-tagging~\cite{Jurafsky:2009:SLP}, and econometrics~\cite{Hamilton:1990:JoE}
-and are a member of the more general family of probabilistic graphical
-models~\cite{Koller:2009:PGM}.  A major challenge in applying HMMs and
+tagging~\cite{Jurafsky:2009:SLP}, and econometrics~\cite{Hamilton:1990:JoE}.
+%and are a member of the more general family of probabilistic graphical
+%models~\cite{Koller:2009:PGM}.  
+A major challenge in applying HMMs and
 other graphical models successfully to solve a problem is to design an
@@
 
-For example, in part-of-speech tagging~\cite{Jurafsky:2009:SLP} the output
+For example, in part-of-speech tagging~\cite{Jurafsky:2009:SLP}, the output
 distributions are categorical (distributions over words from a fixed

### 2017-08-24 sec-conclusion.tex
Whitespace changes and minor edits.

--- 2017-07-25 sec-conclusion.tex
@@
 \section{Conclusions and Future Work}
-% summarize the main contributions of the paper; summarize the experimental
-% results; argue again for our novelty; discuss future directions
 As a tool to help instructors and education researchers better understand
@@
 over actions in order to capture information present in bigrams of actions
-(or above). By varying the number of latent
-states inferred, we showed that the model is flexible and can capture
-patterns of differing levels of specificity in this way.  Finally, we
-investigated whether we can detect differences in student behavior patterns
-as they correlate with course performance. Specifically, we demonstrated
-that high-performing students produce substantially different HMM
-transition diagrams that tend to show longer concentration span in quiz-taking
-and more active  forum participation as compared with the average students.
-These results show the great potential of the proposed model for serving as a tool
-to help humans discover knowledge about student behaviors.
+(or above). By varying the number of latent states inferred, we showed that
+the model is flexible and can capture patterns of differing levels of
+specificity in this way.  Finally, we investigated whether we can detect
+differences in student behavior patterns as they correlate with course
+performance. Specifically, we demonstrated that high-performing students
+produce substantially different HMM transition diagrams that tend to show
+longer concentration span in quiz-taking and more active  forum
+participation as compared with the average students.  These results show
+the great potential of the proposed model for serving as a tool to help
+humans discover knowledge about student behaviors.
 
@@
 
-There are a number of interesting directions to further extend our work in the future. 
-First, the proposed model is 
-completely general and can thus be easily applied to analyze the log of many
-other courses to enable deep understanding of student behaviors as well as
-the correlations of such behaviors and other variables such as grades. To realize these benefits, 
-it would be useful 
-to develop a MOOC log analysis system based on the proposed model so as to facilitate education research and help instructors improve course
-design.
+There are a number of interesting directions to further extend our work in
+the future.  First, the proposed model is completely general and can thus
+be easily applied to analyze the log of many other courses to enable deep
+understanding of student behaviors as well as the correlations of such
+behaviors and other variables such as grades. To realize these benefits, it
+would be useful to develop a MOOC log analysis system based on the proposed
+model so as to facilitate education research and help instructors improve
+course design.
 
-Second, our model can empower many new comparative analyses. For example, we could
-now look at how behavior patterns change between different offerings of
-the same MOOC to understand how changes in course structure or materials
-influence student behavior. Individual students can now also be compared
-against each other or against groups. For example, by decoding the latent
-state sequences for each student, we can measure how ``surprising'' their
-latent state transition sequence is relative to the average we would expect
-according to the model, or to the average ``perfect'' student, etc. We can
-now investigate how certain behavioral patterns correlate with properties
-of a student (e.g., demographics, prior aptitude, etc.). After decoding the
-students' latent state sequences, we could also correlate course-wide
-drifts in these latent states with events in a course. For example, we
-might be able to automatically discover difficult or confusing parts of a
-course by noticing spikes in the distribution of students over latent
-states over time.
+Second, our model can empower many new comparative analyses. For example,
+we could now look at how behavior patterns change between different
+offerings of the same MOOC to understand how changes in course structure or
+materials influence student behavior. Individual students can now also be
+compared against each other or against groups. For example, by decoding the
+latent state sequences for each student, we can measure how ``surprising''
+their latent state transition sequence is relative to the average we would
+expect according to the model, or to the average ``perfect'' student, etc.
+We can now investigate how certain behavioral patterns correlate with
+properties of a student (e.g., demographics, prior aptitude, etc.). After
+decoding the students' latent state sequences, we could also correlate
+course-wide drifts in these latent states with events in a course. For
+example, we might be able to automatically discover difficult or confusing
+parts of a course by noticing spikes in the distribution of students over
+latent states over time.
 
@@
 would be able to provide predictions of when a student is ``at risk'' for
-dropping out under such an extension.
-Also, currently, the model learns a transition matrix over the latent states that
-is \emph{shared} across all students. It would be interesting to instead
-learn a different latent state transition matrix for each individual
-student, but keep the second-level Markov models shared. This would provide
-the model with more flexibility which may be desirable itself, but would
-also naturally result in a description of a student (via his or her HMM
-transitions) that could be incorporated into existing supervised learning
-techniques that try to predict student outcomes for understanding
-which of the latent behavior patterns discovered by 2L-HMM are most predictive
-of the performance of student learning. One could also relax this somewhat
-and posit that \emph{groups} of students transition between the lower layer
-patterns our method identifies in distinct ways and instead do a soft
+dropping out under such an extension. Also, currently, the model learns a
+transition matrix over the latent states that is \emph{shared} across all
+students. It would be interesting to instead learn a different latent state
+transition matrix for each individual student, but keep the second-level
+Markov models shared. This would provide the model with more flexibility
+which may be desirable itself, but would also naturally result in a
+description of a student (via his or her HMM transitions) that could be
+incorporated into existing supervised learning techniques that try to
+predict student outcomes for understanding which of the latent behavior
+patterns discovered by 2L-HMM are most predictive of the performance of
+student learning. One could also relax this somewhat and posit that
+\emph{groups} of students transition between the lower layer patterns
+identified by our method in distinct ways; this way, we can do a soft
 clustering of students into $K_2$ groups based on the similarity of their

### 2017-08-24 sec-intro.tex
Whitespace changes and minor edits.

--- 2017-07-25 sec-intro.tex
@@
       to a particular pattern or group.
-  \item Student behavior can vary over time. Previous models that treat students
-      as exhibiting only one behavioral pattern over
+  \item Student behavior can vary over time. Previous models that treat
+      students as exhibiting only one behavioral pattern over
       time~\cite{Faucon:2016:EDM} miss out on the opportunity to understand
-      student behavior dynamics in a course. We propose a latent space
-      model with \emph{latent state transitions} to flexibly model the
-      dynamics.
+      student behavior dynamics in a course. We propose a latent space model
+      with \emph{latent state transitions} to flexibly model the dynamics.
   \item Analysis of student behavior can and should be performed at varying
@@
 
-Thus, what we propose is a \emph{latent variable approach} for mining student behavior
-patterns that is \emph{probabilistic} for inference and
+Thus, what we propose is a \emph{latent variable approach} for mining
+student behavior patterns that is \emph{probabilistic} for inference and
 does \emph{not} force assumptions about time resolutions, making it
@@
 sequences.  Evaluation results on a MOOC data set on Coursera demonstrate
-that the 2L-HMM can effectively discover a variety of interesting interpretable
-student behavior patterns at different levels of resolution, many of which
-are beyond what existing approaches can discover. We show that the patterns
-uncovered by the 2L-HMM capture meaningful behavior by quantitatively
-showing that features extracted from a trained 2L-HMM correlate with
-learning outcomes. Since our proposed methods are
+that the 2L-HMM can effectively discover a variety of interesting
+interpretable student behavior patterns at different levels of resolution,
+many of which are beyond what existing approaches can discover. We show
+that the patterns uncovered by the 2L-HMM capture meaningful behavior by
+quantitatively showing that features extracted from a trained 2L-HMM
+correlate with learning outcomes. Since our proposed methods are
 unsupervised, they can potentially be applied to any MOOC data without
-requiring manual annotation effort at the level of sequences. Instead,
-instructors are empowered to use the latent patterns that the 2L-HMM can
-discover from raw data to extract knowledge about the behaviors his/her
-students exhibit in the MOOC.
+requiring manual annotation effort at the level of sequences, thus
+empowering instructors to use the latent patterns discovered by the 2L-HMM
+to further extract knowledge about the behaviors his/her students exhibit
+in the MOOC.

### 2017-08-24 sec-model.tex
Whitespace changes and minor edits.

@@
 will result in different underlying training data, and thus different
-meanings behind the patterns that the 2L-HMM will extract. The flexibility of
-using different segmentation strategies is intentional as it allows a user to
-flexibly adjust the segmentation so as to obtain patterns at different granularity levels as needed.
+meanings behind the patterns that the 2L-HMM will extract. The flexibility
+of using different segmentation strategies is intentional as it allows a
+user to adjust the segmentation as needed so as to obtain patterns at
+different granularity levels.
 
@@
 \label{sec:param-estimation}
-To learn the parameters of our model, we may use maximum
-likelihood estimation. Unfortunately, a closed-form solution does not
-exist, so we must appeal to the EM algorithm~\cite{Dempster:1977:JRSS}. In
-particular, we propose a minor modification of the Baum-Welch
-algorithm~\cite{Rabiner:1990:RSR} which is an efficient EM algorithm
-for learning the parameters of
-hidden Markov models in an unsupervised setting where the Markovian assumption is exploited
- to significantly reduce the computational complexity of the EM algorithm by avoiding explicit enumeration of all the possible state transitions.  In the following sections,
-we will provide a brief description of the original Baum-Welch algorithm
-for unsupervised parameter estimation for categorical valued hidden Markov
-models, and then describe our modification to allow for parameter
-estimation for our 2L-HMM modification for sequence valued observations.
+To learn the parameters of our model, we may use maximum likelihood
+estimation. Unfortunately, a closed-form solution does not exist, so we
+must appeal to the EM algorithm~\cite{Dempster:1977:JRSS}. In particular,
+we propose a minor modification of the Baum-Welch
+algorithm~\cite{Rabiner:1990:RSR} which is an efficient EM algorithm for
+learning the parameters of hidden Markov models in an unsupervised setting
+where the Markovian assumption is exploited to significantly reduce the
+computational complexity of the EM algorithm by avoiding explicit
+enumeration of all the possible state transitions.  In the following
+sections, we will provide a brief description of the original Baum-Welch
+algorithm for unsupervised parameter estimation for categorical valued
+hidden Markov models, and then describe our modification to allow for
+parameter estimation for our 2L-HMM modification for sequence valued
+observations.
 
@@
 \begin{itemize}
-  \item $\alpha^{(\mathbf{o})}_1(i) = \pi_i b_i(o_1)$, the probability of starting in
-    state $i$ ($\pi_i$) times the probability of generating the first
-    observation $o_1$ from state $i$.
+  \item $\alpha^{(\mathbf{o})}_1(i) = \pi_i b_i(o_1)$, the probability of
+    starting in state $i$ ($\pi_i$) times the probability of generating the
+    first observation $o_1$ from state $i$.
   \item $\alpha^{(\mathbf{o})}_{t+1}(i) = b_i(o_{t+1})\sum_{j=1}^K
-    \alpha^{(\mathbf{o})}_t(j) A_{ji}$, the probability of generating the observation
-    $o_{t+1}$ from state $i$ times the probability that we arrive in state
-    $i$ from any other previous state after generating all of the other
-    observations.
+    \alpha^{(\mathbf{o})}_t(j) A_{ji}$, the probability of generating the
+    observation $o_{t+1}$ from state $i$ times the probability that we
+    arrive in state $i$ from any other previous state after generating all
+    of the other observations.
 \end{itemize}
@@
 Given $\gamma^{(\mathbf{o})}_t(i)$ and $\xi^{(\mathbf{o})}_t(i,j)$ for each
-sequence $\mathbf{o} \in \mathbf{D}$, we can update our model parameters $\Lambda$
-as
+sequence $\mathbf{o} \in \mathbf{D}$, we can update our model parameters
+$\Lambda$ as
 \begin{align}

### 2017-08-24 sec-limitations.tex
Whitespace changes and minor edits.
Specify what exact additive constant was used for smoothing.

@@
 non-zero probability of generating the observations. We employed a simple
-additive smoothing in our implementation with a small additive constant for
-all of our transition matrices to avoid this problem.
+additive smoothing in our implementation with a small additive constant
+($10^{-6}$) for all of our transition matrices to avoid this problem.
 
@@
 The proposed behavior representation is most suitable for representing
-recurring behavior patterns, which presumably are most interesting
-to extract from the data, but may not cover all the interesting patterns in the data.
-It would be interesting to further explore other complementary models such as
- time series models, which may help capture non-recurring patterns.
+recurring behavior patterns, which presumably are most interesting to
+extract from the data, but may not cover all the interesting patterns in
+the data. It would be interesting to further explore other complementary
+models such as time series models, which may help capture non-recurring
+patterns.
 

### 2017-08-28 sec-conclusion.tex
Address copy-editing feedback.

@@
 patterns with more sophisticated models than simple discrete distributions
-over actions in order to capture information present in bigrams of actions
-(or above). By varying the number of latent states inferred, we showed that
-the model is flexible and can capture patterns of differing levels of
-specificity in this way.  Finally, we investigated whether we can detect
-differences in student behavior patterns as they correlate with course
-performance. Specifically, we demonstrated that high-performing students
-produce substantially different HMM transition diagrams that tend to show
-longer concentration span in quiz-taking and more active  forum
+over actions to capture information present in bigrams of actions (or
+larger sequences). By varying the number of latent states inferred, we
+showed that the model is flexible and can capture patterns of differing
+levels of specificity in this way.  Finally, we investigated whether we can
+detect differences in student behavior patterns as they correlate with
+course performance. Specifically, we demonstrated that high-performing
+students produce substantially different HMM transition diagrams that tend
+to show longer concentration span in quiz-taking and more active  forum
 participation as compared with the average students.  These results show
@@
 would be useful to develop a MOOC log analysis system based on the proposed
-model so as to facilitate education research and help instructors improve
-course design.
+model to facilitate education research and help instructors improve course
+design.
 

### 2017-08-28 sec-intro.tex
Address copy-editing feedback.

@@
 research~\cite{Kizilcec:2013:LAK, Faucon:2016:EDM, Davis:2016:EDM,
-Shih:2010:EDM}. In this paper, we advocate for a particular representation
-of student behavior patterns as well as \emph{behavior transitions} that we
-believe is simultaneously interpretable but also amenable to unsupervised,
-automated discovery via statistical means. Specifically, we choose to
-visualize behavior patterns as labeled directed graphs where a node
-represents a ``behavior state'' (such as watching a lecture video or
-visiting a forum), a directed edge indicates a transition from one
-behavior state to another, node sizes are proportional to steady-state
-probabilities, and edge widths are proportional to the probability of
-leaving a node following that edge. We can use this same representation for
-visualizing both the student behavior patterns as well as the transitioning
-behavior between them. In Figure~\ref{fig:motivating-example} we show a
-hypothetical example of the kind of output our proposed representation
-could convey.  Here we see two different behavioral patterns
-(\ref{fig:motivating-example-0} and \ref{fig:motivating-example-1}) as well
-as the \emph{transition behavior} between these two behavior patterns
-(\ref{fig:motivating-example-trans}).
+Shih:2010:EDM}. In this paper, we advocate for a representation of student
+behavior patterns as well as \emph{behavior transitions} that we believe is
+simultaneously interpretable but also amenable to unsupervised, automated
+discovery via statistical means. Specifically, we choose to visualize
+behavior patterns as labeled directed graphs where a node represents a
+``behavior state'' (such as watching a lecture video or visiting a forum),
+a directed edge indicates a transition from one behavior state to another,
+node sizes are proportional to steady-state probabilities, and edge widths
+are proportional to the probability of leaving a node following that edge.
+We can use this same representation for visualizing both the student
+behavior patterns as well as the transitioning behavior between them. In
+Figure~\ref{fig:motivating-example} we show a hypothetical example of the
+kind of output our proposed representation could convey.  Here we see two
+different behavioral patterns (\ref{fig:motivating-example-0} and
+\ref{fig:motivating-example-1}) as well as the \emph{transition behavior}
+between these two behavior patterns (\ref{fig:motivating-example-trans}).
 \begin{figure}
@@
 MOOC data. Automatically extracting interpretable patterns from the
-clickstream data associated with MOOCs is a necessary step in order for
-instructors to identify the hidden knowledge in massive interaction
-datasets. Without the availability of a suitable model for identifying
-behavioral patterns, instructors are not empowered to use this available
-data to improve their courses without expending extraordinary amounts of
-manual effort (even with which the raw data can still be very hard to
-interpret).
+clickstream data associated with MOOCs is a necessary step for instructors
+to identify the hidden knowledge in massive interaction datasets. Without
+the availability of a suitable model for identifying behavioral patterns,
+instructors are not empowered to use this available data to improve their
+courses without expending extraordinary amounts of manual effort (even with
+which the raw data can still be very hard to interpret).
 
@@
       to automatically capture patterns that we might not have been able to
-      articulate clearly a priori via a series of rules, and also allows us
-      to model the inherent uncertainty in assigning a student's behavior
-      to a particular pattern or group.
+      articulate clearly a priori via a series of rules, and allows us to
+      model the inherent uncertainty in assigning a student's behavior to a
+      pattern or group.
   \item Student behavior can vary over time. Previous models that treat
@@
       with \emph{different levels of resolution}; existing models tend to come
-      with a particular assumption about the resolution of time they
+      with an assumption about the resolution of time they
       consider~\cite{Faucon:2016:EDM, Kizilcec:2013:LAK, Shih:2010:EDM}. We

### 2017-08-28 sec-limitations.tex
Address copy-editing feedback.

--- 2017-07-25 sec-limitations.tex
@@
 ease with which instructors are able to extract knowledge from these
-patterns. We discuss both of these lines below.
+patterns. We discuss both lines below.
 
@@
 One potential limitation is that the model is complex and has many parameters
-in order to truly uncover the relevant patterns in the data. In order to
-properly estimate these parameters at training time, a large amount of data
-must be available to the training algorithm. The assumption that we have a
-large amount of sequence data available for training on should generally
-hold for most MOOC courses, but this assumption may be problematic if
-attempting to apply our model to smaller online (or on-campus) courses.
+in order to truly uncover the relevant patterns in the data. To properly
+estimate these parameters at training time, a large amount of data must be
+available to the training algorithm. The assumption that we have a large
+amount of sequence data available for training on should hold for most MOOC
+courses, but this assumption may be problematic if attempting to apply our
+model to smaller online (or on-campus) courses.
 
@@
 patterns. Empirically, however, we observed in our experiments that strong
-patterns tend to always show up as long as the algorithm reaches a
-reasonably good local maximum, and the differences of the results tend to
-be related to weak ``unstable'' patterns which may not be reliable anyway.
-Since it is far more important and useful to reveal the strong salient
-patterns than weak unreliable ones for instructors, the problem might not
-necessarily affect the utility of the approach so significantly.  A
-commonly applied approach to address the problem of multiple local maxima
-is to run the model multiple times with different starting points to allow
-the model to explore a larger portion of the potential parameter space. One
-can then compare the log-likelihood of the data between the models that
-were started from different initialization points and select the one that
-has the highest value. This still does not guarantee that we find a global
-maximum, but it does help alleviate the potential for finding a
-particularly bad local maximum.  In practice, we have found that our model
-can be fit to the data fairly quickly on commodity hardware, so running it
-multiple times to address this concern is not computationally unfeasible.
+patterns tend to always show up if algorithm reaches a reasonably good
+local maximum, and the differences of the results tend to be related to
+weak ``unstable'' patterns which may not be reliable anyway.  Since it is
+far more important and useful to reveal the strong salient patterns than
+weak unreliable ones for instructors, the problem might not necessarily
+affect the utility of the approach so significantly.  A commonly applied
+approach to address the problem of multiple local maxima is to run the
+model multiple times with different starting points to allow the model to
+explore a larger portion of the parameter space. One can then compare the
+log-likelihood of the data between the models that were started from
+different initialization points and select the one that has the highest
+value. This still does not guarantee that we find a global maximum, but it
+does help alleviate the potential for finding a particularly bad local
+maximum. In practice, we have found that our model can be fit to the data
+quickly on commodity hardware, so running it multiple times to address this
+concern is not computationally unfeasible.
 
@@
 non-zero probability of generating the observations. We employed a simple
-additive smoothing in our implementation with a small additive constant for
-all of our transition matrices to avoid this problem.
+additive smoothing in our implementation with a small additive constant
+($10^{-6}$) for all our transition matrices to avoid this problem.
 
@@
 The proposed behavior representation is most suitable for representing
-recurring behavior patterns, which presumably are most interesting
-to extract from the data, but may not cover all the interesting patterns in the data.
-It would be interesting to further explore other complementary models such as
- time series models, which may help capture non-recurring patterns.
+recurring behavior patterns, which presumably are most interesting to
+extract from the data, but may not cover all the interesting patterns in
+the data. It would be interesting to explore other complementary models
+such as time series models, which may help capture non-recurring patterns.
 
@@
 scenario. Varying the segmentation strategy changes the granularity of the
-patterns that are required to be explained by the latent states, which will
-change their meaning. Varying the number of latent states increases the
-flexibility of the model, but also may result in latent states that are not
-substantially different from the other latent states and/or latent states
-with very low probability. This flexibility forces a user of our model to
-make some assumptions about the patterns they wish to find (granularity,
-diversity), and the model itself does not necessarily provide clear
-guidance as to what the best approach is.
+patterns the latent states explain, which will change their meaning.
+Varying the number of latent states increases the flexibility of the model,
+but also may result in latent states that are not substantially different
+from the other latent states and/or latent states with very low
+probability. This flexibility forces a user of our model to make some
+assumptions about the patterns they wish to find (granularity, diversity),
+and the model itself does not necessarily provide clear guidance as to what
+the best approach is.
 
@@
 latent states only occur at action sequence boundaries. One can model how
-much time a user stays in a given state in terms of the number of sessions
+much time a user stays in each state in terms of the number of sessions
 they remain there before transitioning, but it may be better to directly
-model the amount of time we expect a user to stay in the given state
-directly. In other words, a more powerful model might be one in which the
-segmentation and the latent behavior pattern discovery are jointly modeled
-in a single probabilistic framework rather than being completely separate
-pieces as we investigate in this work.
+model the amount of time we expect a user to stay in the given state.  In
+other words, a more powerful model might be one in which the segmentation
+and the latent behavior pattern discovery are jointly modeled in a single
+probabilistic framework rather than being separate pieces as we investigate
+in this work.
 

### 2017-08-28 sec-model.tex
Address copy-editing feedback.

@@
 $u$, the HMM can generate an observable data point $x$ according to an
-output probabilistic model $p(x \mid u)$. Thus if we ``run'' an HMM for $N$
+output probabilistic model $p(x \mid u)$. Thus, if we ``run'' an HMM for $N$
 time points denoted by $t=1,\ldots,N$, the HMM could ``generate'' a sequence
@@
 would tend to have many video-watching related behaviors and occasionally
-forum activities, while in the quiz-taking state (in order to pass each
-module), the student would tend to show many quiz-related ``micro''
-activities as well as asking questions or checking discussions on the
-forum. Note that due to the complexity of the student behavior, it is very
-difficult to accurately \emph{prescribe} the specific surface ``micro''
-behavior patterns for each state in advance, especially without  prior
-knowledge about the students. For example, forum activities are likely
-interleaved with other activities   in every task state and the
-interleaving pattern can be somewhat irregular with potentially many
-variations.  The major motivations for using an HMM are that
+forum activities, while in the quiz-taking state (to pass each module), the
+student would tend to show many quiz-related ``micro'' activities as well
+as asking questions or checking discussions on the forum. Note that due to
+the complexity of the student behavior, it is very difficult to accurately
+\emph{prescribe} the specific surface ``micro'' behavior patterns for each
+state in advance, especially without  prior knowledge about the students.
+For example, forum activities are likely interleaved with other activities
+in every task state and the interleaving pattern can be somewhat irregular
+with potentially many variations. The major motivations for using an HMM
+are that
 \begin{enumerate*}[label=(\arabic*)]
@@
     kind of transitions are most probable, and which states tend to be more
-    long-lasting for any particular set of students.
+    long-lasting for any set of students.
 \end{enumerate*}
 
-However, if we use an ordinary HMM to analyze our data, we would treat each
-observed ``micro'' activity (such as video watching or forum post reading)
-as an output symbol, and thus the output distribution $p(x \mid u)$ for each
-discovered latent state would be a simple distribution over all kinds of
-observable micro activities recorded in our log data (e.g., 50\% lecture
-watching, 8\% quiz taking, 7\% quiz submission, 2\% course wiki reading
-\ldots). While such a distribution is meaningful and can already help us
-interpret the corresponding latent state, it only gives us a rather
-superficial characterization of student behavior.
+However, if we were to use an ordinary HMM to analyze our data, we would
+treat each observed ``micro'' activity (such as video watching or forum
+post reading) as an output symbol, and thus the output distribution $p(x
+\mid u)$ for each discovered latent state would be a simple distribution
+over all kinds of observable micro activities recorded in our log data
+(e.g., 50\% lecture watching, 8\% quiz taking, 7\% quiz submission, 2\%
+course wiki reading, etc.). While such a distribution is meaningful and
+can already help us interpret the corresponding latent state, it only gives
+us a rather superficial characterization of student behavior.
 
@@
 model over the ``micro'' actions with the ``deep'' hidden Markov model over
-the latent task states gives us a fairly general and powerful two-layer HMM
+the latent task states gives us a general and powerful two-layer HMM
 (2L-HMM) that can simultaneously learn ``deeply'' the latent task/behavior
@@
 of using different segmentation strategies is intentional as it allows a
-user to adjust the segmentation as needed so as to obtain patterns at
-different granularity levels.
+user to adjust the segmentation as needed to obtain patterns at different
+granularity levels.
 

### 2017-08-28 sec-related-work.tex
Whitespace changes and minor edits.
Address copy-editing feedback.

@@
 
-Our model is based heavily on the prior art of Hidden Markov
-Models (HMMs)~\cite{Rabiner:1990:RSR} for sequence labeling tasks.  As a member
-of the more general family of probabilistic graphical models~\cite{Koller:2009:PGM}, HMMs are widely applicable and have been used for tasks such as speech
-recognition~\cite{Huang:1990:HMM}, part-of-speech
-tagging~\cite{Jurafsky:2009:SLP}, and econometrics~\cite{Hamilton:1990:JoE}.
-%and are a member of the more general family of probabilistic graphical
-%models~\cite{Koller:2009:PGM}.  
-A major challenge in applying HMMs and
-other graphical models successfully to solve a problem is to design an
+Our model is based heavily on the prior art of Hidden Markov Models
+(HMMs)~\cite{Rabiner:1990:RSR} for sequence labeling tasks. As a member of
+the more general family of probabilistic graphical
+models~\cite{Koller:2009:PGM}, HMMs are widely applicable and have been
+used for tasks such as speech recognition~\cite{Huang:1990:HMM},
+part-of-speech tagging~\cite{Jurafsky:2009:SLP}, and
+econometrics~\cite{Hamilton:1990:JoE}. A major challenge in applying HMMs
+and other graphical models successfully to solve a problem is to design an
 appropriate architecture of the model, which always varies according to
@@
 having an impact on the resulting real-valued vector data observed. The
-``regimes'' are represented with some sort of model that can produce
-real-valued vector data, such as a multivariate Gaussian or an
-auto-regressive model. The analogy with HMMs is that a ``regime'' is a
-latent state, and the characterization of the regime itself is the output
-distribution for that latent state. Our model can be seen as such a
-``regime-switching'' model where the output of the ``regimes'' that
-students are switching between are discrete-valued \emph{sequences} (as
-opposed to real numbers, vectors of real numbers, or categorical symbols)
-and the model used to represent a specific ``regime'' is an (observable)
-Markov chain over the observed student actions. We view the switching
-between ``regimes'' as the first ``layer'' of our model, and the
-transitioning behavior \emph{within} a ``regime'' between the actions a
-student takes as the second ``layer'' of our model.
+``regimes'' are thus represented with a model that can produce real-valued
+vector data, such as a multivariate Gaussian or an auto-regressive model.
+The analogy with HMMs is that a ``regime'' is a latent state, and the
+characterization of the regime itself is the output distribution for that
+latent state. Our model can be seen as such a ``regime-switching'' model
+where the output of the ``regimes'' that students are switching between are
+discrete-valued \emph{sequences} (as opposed to real numbers, vectors of
+real numbers, or categorical symbols) and the model used to represent a
+specific ``regime'' is an (observable) Markov chain over the observed
+student actions. We view the switching between ``regimes'' as the first
+``layer'' of our model, and the transitioning behavior \emph{within} a
+``regime'' between the actions a student takes as the second ``layer'' of
+our model.
 
@@
 lower level into the input of the higher level. Instead, our lower level is
-actually treated using a non-hidden Markov model, and the higher level is
-modeling transitions between the $K$ different non-hidden Markov models we
-consider. The problem to be solved is similar in that we wish to predict a
-``label'' for a sequence of actions a student takes as well as understand
-the transition behavior between those labels.  However, one of the
-consequences of modeling the lower layer using a non-hidden Markov model
-instead of an HMM directly is that the meaning of the $K$ different latent
-states can be more clearly captured by using our proposed behavior
-representation. If we were to use an HMM as our first layer, the behavior
-patterns (like from Figure~\ref{fig:motivating-example}) would instead have
-nodes that represent another set of latent states instead of being concrete
-actions themselves. Each of these latent states would then be associated
-with some other output distribution over the possible concrete actions to
-be considered. In order to understand a single behavior pattern uncovered,
-one would first have to understand the different output distributions for
-the latent states in that pattern in order to understand the meaning of
-that latent state. This further complicates the understanding of higher
-level patterns because understanding the higher layer patterns requires
-understanding the lower layer patterns. By taking a more restrictive view
-of the first layer, we are able to produce a representation that can be
-more readily interpreted due to the states in our first layer
-representation having an immediately clear meaning (the concrete action
-they represent).
+treated using a non-hidden Markov model, and the higher level is modeling
+transitions between the $K$ different non-hidden Markov models we consider.
+The problem to be solved is similar in that we wish to predict a ``label''
+for a sequence of actions a student takes as well as understand the
+transition behavior between those labels.  However, one of the consequences
+of modeling the lower layer using a non-hidden Markov model instead of an
+HMM directly is that the meaning of the $K$ different latent states can be
+more clearly captured by using our proposed behavior representation. If we
+were to use an HMM as our first layer, the behavior patterns (like from
+Figure~\ref{fig:motivating-example}) would instead have nodes that
+represent another set of latent states instead of being concrete actions
+themselves. Each of these latent states would then be associated with some
+other output distribution over the possible concrete actions to be
+considered. In order to understand a single behavior pattern uncovered, one
+would first have to understand the different output distributions for the
+latent states in that pattern to understand the meaning of that latent
+state. This further complicates the understanding of higher level patterns
+because understanding the higher layer patterns requires understanding the
+lower layer patterns. By taking a more restrictive view of the first layer,
+we can produce a representation that can be more readily interpreted due to
+the states in our first layer representation having an immediately clear
+meaning (the concrete action they represent).
 
@@
 clustering techniques for automatic discovery of student learning
-strategies when solving a particular problem.  This is similar to our
-approach in that the description of behavior profiles is a Markov model,
-but cannot further characterize each latent state with another informative
-HMM. Thus their work can be regarded to modeling ``micro'' behavior,
-whereas our model can model both ``micro'' and ``macro'' behavior.
+strategies when solving a problem. This is similar to our approach in that
+the description of behavior profiles is a Markov model, but cannot further
+characterize each latent state with another informative HMM. Thus, their
+work can be regarded to modeling ``micro'' behavior, whereas our model can
+model both ``micro'' and ``macro'' behavior.
 
 \citet{Davis:2016:EDM} investigate frequent student behavior pattern chains
-with a set of actions that is defined similarly to ours. However, their
+with a set of actions that are defined similarly to ours. However, their
 method for finding the common behavioral patterns involves a manual

### 2017-08-28 sec-results.tex
Add link to the source code for the analysis.
Address copy-editing feedback.

@@
 \end{enumerate*}
-Thus to evaluate the proposed model, we conduct experiments to
+Thus, to evaluate the proposed model, we conduct experiments to
 qualitatively analyze both types of patterns discovered from empirical MOOC
@@
 the text as we discuss the plots. The plots were created using
-python-igraph: \url{http://igraph.org/python/}.}.
+python-igraph: \url{http://igraph.org/python/}.}. In the interest of
+reproducibility, the source code for analyzing the MOOC logs we use in this
+paper and for producing the figures themselves is publicly available
+as open-source
+software\footnote{\url{https://github.com/skystrife/clickstream-hmm}}.
 
@@
 these two latent states: in the first you are more likely to visit the
-forum \emph{looking for a particular post}, where in the second you are
-more likely to visit the forum \emph{to browse existing posts}. Thus,
-capturing the action transition matrix within a latent state is important
-for capturing detailed insights involving bigrams of actions.
+forum \emph{looking for a post}, where in the second you are more likely to
+visit the forum \emph{to browse existing posts}. Thus, capturing the action
+transition matrix within a latent state is important for capturing detailed
+insights involving bigrams of actions.
 
@@
 insight into the lecture viewing behavior of students in these two MOOCs
-which can possibly be a reflection of the course's structure (which, as
-demonstrated in \citet{Davis:2016:EDM}, can differ substantially across
-different MOOCs).  In \textretrieval{}, students are likely to view lecture
-videos in succession directly without first visiting the course wiki,
-whereas in \sustain{} students are much more likely to first return the
-course wiki before watching the next lecture video. This observation would
-be lost if we did not consider the transitions between the actions within
-the latent states.
+which might reflect the course's structure (which, as demonstrated in
+\citet{Davis:2016:EDM}, can differ substantially across different MOOCs).
+In \textretrieval{}, students are likely to view lecture videos in
+succession directly without first visiting the course wiki, whereas in
+\sustain{} students are much more likely to first return the course wiki
+before watching the next lecture video. This observation would be lost if
+we did not consider the transitions between the actions within the latent
+states.
 
@@
 discover patterns of different granularity, and we can show this by varying
-$K$ for a particular course and observing how the latent state
-representations evolve.
+$K$ for a course and observing how the latent state representations evolve.
 
@@
 a video watching pattern (state 1) and a course material browsing pattern
-(state 0). However, we can see that when $K=3$ we are able to uncover a new
-pattern involving forum behavior in state 3 (notice the node sizes on the
-bottom right). As we increase $K$ to four, we can see that state 1 splits
-into state 1 and state 3. These states appear quite similar at a glance,
-but there are still some key differences. First, we can see that state 1
-now has a non-negligible forum component, whereas state 3 has hardly any
-weight on the forum.
+(state 0). However, we can see that when $K=3$ we can uncover a new pattern
+involving forum behavior in state 3 (notice the node sizes on the bottom
+right). As we increase $K$ to four, we can see that state 1 splits into
+state 1 and state 3. These states appear quite similar at a glance, but
+there are still some key differences. First, we can see that state 1 now
+has a non-negligible forum component, whereas state 3 has hardly any weight
+on the forum.
 
@@
 that they attempted all required quizzes in the course, but such that their
-average quiz score was $\leq 70\%$. Here, we see that \emph{state 1} has a
+average quiz score was $\leq 70\%$. Here, we see that state 1 has a
 large increase in size, where we might have expected state 3 to grow
@@
 Figure~\ref{fig:trans-low} that leave state 2. In the ``perfect'' group,
-nearly all of this probability mass is allocated for the self-loop ($p
+nearly all this probability mass is allocated for the self-loop ($p
 \approx 0.93$). In the ``low'' group, this self-loop is less strong ($p

### 2017-08-29 sec-model.tex
Minor wording update for clarity in model section.

@@
 going through a sequence of \emph{hidden states} $w_1 \ldots w_N$ where
-$w_i \in U$ is a state. The association of such a latent sequence of state
-transitions with the observed symbols makes it possible to use the HMM to
-``decode'' the latent behavior of students behind the surface behavior we
-directly observe in the log data, allowing for understanding student
-behavior more deeply than a model with no latent state variables.
+each $w_i$ is a random variable taking a value from the HMM's state set
+$U$. The association of such a latent sequence of state transitions with
+the observed symbols makes it possible to use the HMM to ``decode'' the
+latent behavior of students behind the surface behavior we directly observe
+in the log data, allowing for understanding student behavior more deeply
+than a model with no latent state variables.
 
